---
title: "Basic stats sgRNA library"
author: "Author: Stefanus Bernard"
date: "Date: `r format(Sys.time(), '%d %B %Y')`"
output: 
    html_document:
      theme: cerulean
      toc: true
      toc_float: true
      number_sections: true
      self_contained: true
---

```{r}
library(tidyverse)
library(readxl)
library(VennDiagram)
library(grid)
library(paletteer)
library(RColorBrewer)
source('../CRISPR-KO-library-refinement-pipeline/sgRNA_library_refinement_pipeline_functions.R')
source("./00_CRISPR-KO-off-target-paralog-functions.R")
```

# Import the data of sgRNA library report

```{r}
brunello <- 'broadgpp-brunello-library-contents'
toronto <- 'tkov3_guide_sequence'
avana <- 'avana_library'
yusa <- 'yusa_hcrispr_ko_grnas'

brunello_report <- read_excel(paste('../CRISPR-KO-library-refinement-pipeline/output_cleaning/T2T-CHM13/', brunello, '_full_report.xlsx', sep = ''))
toronto_report <- read_excel(paste('../CRISPR-KO-library-refinement-pipeline/output_cleaning/T2T-CHM13/', toronto, '_full_report.xlsx', sep = ''))
avana_report <- read_excel(paste('../CRISPR-KO-library-refinement-pipeline/output_cleaning/T2T-CHM13/', avana, '_full_report.xlsx', sep = ''))
yusa_report <- read_excel(paste('../CRISPR-KO-library-refinement-pipeline/output_cleaning/T2T-CHM13/', yusa, '_full_report.xlsx', sep = ''))

```

## Import paralog data from Barbara's paper

```{r}

# import predicted paralog pairs from Barbara's paper

predicted_paralog_pairs <- list_paralog_gene_pairs()

```

# Calculate the percentage of multi-target, single-mismatch, and PAM-distal double mismatch sgRNA in each libraries

```{r}

brunello_multi_target <- round(calculate_percentage(brunello_report, "multi-target sgRNA", "sgRNA number"), 2)
toronto_multi_target <- round(calculate_percentage(toronto_report, "multi-target sgRNA", "sgRNA number"), 2)
avana_multi_target <- round(calculate_percentage(avana_report, "multi-target sgRNA", "sgRNA number"), 2)
yusa_multi_target <- round(calculate_percentage(yusa_report, "multi-target sgRNA", "sgRNA number"), 2)

brunello_single_mismatch <- round(calculate_percentage(brunello_report, "single mismatch sgRNA", "sgRNA number"), 2)
toronto_single_mismatch <- round(calculate_percentage(toronto_report, "single mismatch sgRNA", "sgRNA number"), 2)
avana_single_mismatch <- round(calculate_percentage(avana_report, "single mismatch sgRNA", "sgRNA number"), 2)
yusa_single_mismatch <- round(calculate_percentage(yusa_report, "single mismatch sgRNA", "sgRNA number"), 2)

brunello_pam_distal <- round(calculate_percentage(brunello_report, "PAM distal double mismatch", "sgRNA number"), 2)
toronto_pam_distal <- round(calculate_percentage(toronto_report, "PAM distal double mismatch", "sgRNA number"), 2)
avana_pam_distal <- round(calculate_percentage(avana_report, "PAM distal double mismatch", "sgRNA number"), 2)
yusa_pam_distal <- round(calculate_percentage(yusa_report, "PAM distal double mismatch", "sgRNA number"), 2)


# Create dataframe for multi-target, single-mismatch, and pam-distal from all library

multi_target <- c(brunello_multi_target, toronto_multi_target, yusa_multi_target, avana_multi_target)
single_mismatch <- c(brunello_single_mismatch, toronto_single_mismatch, yusa_single_mismatch, avana_single_mismatch)
pam_distal <- c(brunello_pam_distal, toronto_pam_distal, yusa_pam_distal, avana_pam_distal)

library <- c("brunello", "toronto_v3", "yusa", "avana")
factor <- c(1,2,3,4)

sum_dataframe <- data.frame(library, multi_target, single_mismatch, pam_distal)


# Histogram of multi-target/off-target sgRNAs in Cellecta, Brunello, and Toronto
# Reorder the library factor to ensure the bars are ordered as Brunello, Toronto, Yusa, Avana
sum_dataframe_long <- sum_dataframe %>%
    pivot_longer(cols = c("multi_target", "single_mismatch", "pam_distal"), 
                 names_to = "category", 
                 values_to = "percentage") %>%
    mutate(category = factor(category, levels = c("multi_target", "single_mismatch", "pam_distal")),
           library = factor(library, levels = c("brunello", "toronto_v3", "yusa", "avana")))


# Define custom colors for the libraries
my_palette <- brewer.pal(n = 4, name = "Set2")


# Create the grouped multibar chart for each sgRNA libraries
multi_bar_chart <- ggplot(sum_dataframe_long, aes(x = category, y = percentage, fill = library)) +
    geom_bar(stat = "identity", position = position_dodge2(padding = 0.2), width = 0.7) +
    theme_minimal() +
    labs(title = "", x = "", y = "sgRNA (%)") +
    scale_x_discrete(labels = c('multi_target' = 'Multi-target sgRNA',
                                'single_mismatch' = 'Single mismatch sgRNA',
                                'pam_distal' = 'PAM distal \n double mismatch sgRNA')) +
    geom_text(aes(label = paste(percentage, "%", sep = '')), position = position_dodge2(width = 0.7), vjust = -0.5, size = 9) +
    theme(axis.text.x = element_text(size = 40, vjust = 0.7, colour = 'black'),
          axis.text.y = element_text(size = 40),
          axis.title.y = element_text(size = 40, vjust = 0.5, margin = margin(r = 20)),
          panel.grid.major = element_blank(),
          panel.grid.minor = element_blank(),
          axis.line.x = element_line(size = 0.5),
          axis.line.y = element_line(size = 0.5),
          legend.title = element_text(size = 40),
          legend.text = element_text(size = 40),
          plot.margin = margin(1, 1, 1, 1), 
          line = element_line((size = 4), colour = 'black'),
          legend.spacing.x = unit(1, 'cm')) +
    scale_fill_manual(values = my_palette,
                      name = 'Library',
                      breaks = c("brunello", "toronto_v3", "yusa", "avana"),
                      labels = c("brunello" = "Brunello",
                                 "toronto_v3" = "Toronto Knockout V3",
                                 "yusa" = "Yusa Project Score",
                                 "avana" = "Avana")) + 
  scale_y_continuous(breaks=seq(0, 10, 2), limits=c(0, 10))

multi_bar_chart

ggsave(
    path = './',
    width = 30,
    height = 15,
    dpi = 300,
    plot = multi_bar_chart,
    filename = paste('./plot/percentage_off_target_T2T.png', sep = ''))
```

# Percentage of sgRNAs not represented after library refinement
# Percentage of genes not represented after library refinement

```{r}

# brunello_library <- import_library("../CRISPR-KO-library-refinement-pipeline/public_crispr_library/broadgpp-brunello-library-contents.tsv")
# toronto_library <- import_library("../CRISPR-KO-library-refinement-pipeline/public_crispr_library/tkov3_guide_sequence.tsv")
# avana_library <- import_library("../CRISPR-KO-library-refinement-pipeline/public_crispr_library/avana_library.tsv")
# yusa_library <- import_library("../CRISPR-KO-library-refinement-pipeline/public_crispr_library/yusa_hcrispr_ko_grnas.tsv")
# 
# brunello_refined <- import_library("../CRISPR-KO-library-refinement-pipeline/output_cleaning/hg38/broadgpp-brunello-library-contents_65K_refined.tsv")
# toronto_refined <- import_library("../CRISPR-KO-library-refinement-pipeline/output_cleaning/hg38/tkov3_guide_sequence_62K_refined.tsv")
# avana_refined <- import_library("../CRISPR-KO-library-refinement-pipeline/output_cleaning/hg38/avana_library_56K_refined.tsv")
# yusa_refined <- import_library("../CRISPR-KO-library-refinement-pipeline/output_cleaning/hg38/yusa_hcrispr_ko_grnas_76K_refined.tsv")


avana_result_gene <- percentages_and_list_genes(avana_report, predicted_paralog_pairs[[2]])
brunello_result_gene <- percentages_and_list_genes(brunello_report, predicted_paralog_pairs[[2]])
toronto_result_gene <- percentages_and_list_genes(toronto_report, predicted_paralog_pairs[[2]])
yusa_result_gene <- percentages_and_list_genes(yusa_report, predicted_paralog_pairs[[2]])


df_percentage <- data.frame(library = c("brunello", "toronto_v3", "yusa", "avana"),
                            paralog = c(brunello_result_gene[[2]], toronto_result_gene[[2]], yusa_result_gene[[2]], avana_result_gene[[2]]),
                            not_paralog = c(brunello_result_gene[[3]], toronto_result_gene[[3]], yusa_result_gene[[3]], avana_result_gene[[3]])
                            )

df_percentage <- df_percentage %>%
  mutate(library = factor(library, levels = c("brunello", "toronto_v3", "yusa", "avana"))) %>%
  pivot_longer(!library, names_to = "not_represented_genes", values_to = "percentage")


```

# Plot of genes that are not well represented in the refined library

```{r}

removed_paralog <- ggplot(df_percentage, aes(x = library, y = percentage, fill = not_represented_genes)) +
      geom_bar(stat = "identity", position = "stack") +
      scale_fill_discrete(
          name = "Gene type",
          labels = c("Not paralog", "Paralog")) +
        theme_minimal() +
        labs(title = "", x = "", y = "Genes targeted by < 3 sgRNAs (%)") +
        scale_x_discrete(labels = c("brunello" = "Brunello",
                                     "toronto_v3" = "Toronto \nKnockout V3",
                                     "yusa" = "Yusa \nProject Score",
                                     "avana" = "Avana")) +
        geom_text(aes(label = paste(percentage, "%", sep = '')), position = position_stack(), vjust = 2, size = 6) +
        theme(axis.text.x = element_text(size = 25, vjust = 0.7, colour = 'black'),
              axis.text.y = element_text(size = 25),
              axis.title.y = element_text(size = 25, vjust = 0.5, margin = margin(r = 20)),
              panel.grid.major = element_blank(),
              panel.grid.minor = element_blank(),
              axis.line.x = element_line(size = 0.5),
              axis.line.y = element_line(size = 0.5),
              legend.title = element_text(size = 25),
              legend.text = element_text(size = 25),
              plot.margin = margin(1, 1, 1, 1), 
              line = element_line((size = 4), colour = 'black'),
              legend.spacing.x = unit(1, 'cm')) +
      ylim(0, 100)

ggsave(
    path = './',
    width = 15,
    height = 10,
    dpi = 300,
    plot = removed_paralog,
    filename = paste('./plot/library_genes_removed_paralog_hg38.png', sep = ''))


```

# Plot of sgRNAs that are not represented in the refined library

```{r}

my_palette <- brewer.pal(n = 4, name = "Set2")

percentage_bar_chart <- ggplot(df_percentage, aes(x = library, y = percentage_genes, fill = library)) +
    geom_bar(stat = "identity", position = position_dodge2(padding = 0.2), width = 0.7) +
    theme_minimal() +
    labs(title = "", x = "", y = "Genes targeted by < 3 sgRNAs (%)") +
    scale_x_discrete(labels = c("brunello" = "Brunello",
                                 "toronto_v3" = "Toronto \nKnockout V3",
                                 "yusa" = "Yusa \nProject Score",
                                 "avana" = "Avana")) +
    geom_text(aes(label = paste(percentage_genes, "%", sep = '')), position = position_dodge2(width = 0.7), vjust = -0.5, size = 12) +
    theme(axis.text.x = element_text(size = 40, vjust = 0.7, colour = 'black'),
          axis.text.y = element_text(size = 40),
          axis.title.y = element_text(size = 40, vjust = 0.5, margin = margin(r = 20)),
          panel.grid.major = element_blank(),
          panel.grid.minor = element_blank(),
          axis.line.x = element_line(size = 0.5),
          axis.line.y = element_line(size = 0.5),
          legend.title = element_text(size = 40),
          legend.text = element_text(size = 40),
          plot.margin = margin(1, 1, 1, 1), 
          line = element_line((size = 4), colour = 'black'),
          legend.spacing.x = unit(1, 'cm')) +
    scale_fill_manual(values = my_palette,
                      name = 'Library',
                      breaks = c("brunello", "toronto_v3", "yusa", "avana"),
                      labels = c("brunello" = "Brunello",
                                 "toronto_v3" = "Toronto Knockout V3",
                                 "yusa" = "Yusa Project Score",
                                 "avana" = "Avana")) +
    ylim(0, 100)

percentage_bar_chart

ggsave(
    path = './',
    width = 25,
    height = 15,
    dpi = 300,
    plot = percentage_bar_chart,
    filename = paste('./plot/library_genes_bar_chart_T2T.png', sep = ''))

```

