---
title: "Analysis of multi-target and off-target sgRNAs across libraries"
author: "Author: Stefanus Bernard"
date: "Date: `r format(Sys.time(), '%d %B %Y')`"
output: 
  html_document :
    theme : cerulean
    toc : true
    toc_float : true
    number_section : true
---

```{r setup, include=FALSE}
library(tidyverse)
library(VennDiagram)
library(ggvenn)
library(ggbreak)
library(BSgenome)
library(BSgenome.Hsapiens.UCSC.hg38)
library(BSgenome.Hsapiens.NCBI.T2T.CHM13v2.0)

source('../CRISPR-KO-library-refinement-pipeline/sgRNA_library_refinement_pipeline_functions.R')
source("./00_CRISPR-KO-off-target-paralog-functions.R")
```

# Import alignment data of each sgRNA libraries and library

```{r}
# import sgRNA library
# WE SHOULD DISPOSE ANY CONTROL FROM THE SGRNA LIBRARY AS THEY ARE NOT USED IN THIS ANALYSIS

terms <- c("CONTROL", "Control", "control", "INTRON", "Intron", "intron", "LacZ", "luciferase")
# lacZ and luciferase are the control in Toronto V3 library

terms_for_filtering <- paste(terms, collapse = "|")

avana_library_meyers <- import_sgrna_library("../CRISPR-KO-library-refinement-pipeline/public_crispr_library/avana_library.tsv")
brunello_library <- import_sgrna_library("../CRISPR-KO-library-refinement-pipeline/public_crispr_library/broadgpp-brunello-library-contents.tsv")
toronto_library <- import_sgrna_library("../CRISPR-KO-library-refinement-pipeline/public_crispr_library/tkov3_guide_sequence.tsv")
yusa_library <- import_sgrna_library("../CRISPR-KO-library-refinement-pipeline/public_crispr_library/yusa_hcrispr_ko_grnas.tsv")


# some parameters
annotation <- "../CRISPR-KO-library-refinement-pipeline/annotation_file/CCDS.20221027.txt"
genome_type_version <- "hg38"
bsgenome <- BSgenome.Hsapiens.UCSC.hg38


# hg38
avana_alignment_meyers <- import_sgrna_library_alignment("../CRISPR-KO-library-refinement-pipeline/object_intermediate/hg38/avana_library_aln.csv", genome_type_version)
brunello_alignment <- import_sgrna_library_alignment("../CRISPR-KO-library-refinement-pipeline/object_intermediate/hg38/broadgpp-brunello-library-contents_aln.csv", genome_type_version)
toronto_alignment <- import_sgrna_library_alignment("../CRISPR-KO-library-refinement-pipeline/object_intermediate/hg38/tkov3_guide_sequence_aln.csv", genome_type_version)
yusa_alignment <- import_sgrna_library_alignment("../CRISPR-KO-library-refinement-pipeline/object_intermediate/hg38/yusa_hcrispr_ko_grnas_aln.csv", genome_type_version)

# T2T-CHM13
# avana_alignment_meyers <- import_sgrna_library_alignment("../CRISPR-KO-library-refinement-pipeline/object_intermediate/T2T-CHM13_bsgenome/avana_library_aln.csv", "T2T-CHM13")
# brunello_alignment <- import_sgrna_library_alignment("../CRISPR-KO-library-refinement-pipeline/object_intermediate/T2T-CHM13_bsgenome/broadgpp-brunello-library-contents_aln.csv", "T2T-CHM13")
# toronto_alignment <- import_sgrna_library_alignment("../CRISPR-KO-library-refinement-pipeline/object_intermediate/T2T-CHM13_bsgenome/tkov3_guide_sequence_aln.csv", "T2T-CHM13")
# yusa_alignment <- import_sgrna_library_alignment("../CRISPR-KO-library-refinement-pipeline/object_intermediate/T2T-CHM13_bsgenome/yusa_hcrispr_ko_grnas_aln.csv", "T2T-CHM13")

```

## Import paralog data from Barbara's paper

```{r}

# import predicted paralog pairs from Barbara's paper

barbara_paralog <- list_paralog_gene_pairs()

paralog_pairs <- barbara_paralog[[1]]
paralog_individual <- barbara_paralog[[2]]

```


# Classify the sgRNAs in each libraries

```{r}

avana_classification <- off_target_classification(avana_alignment_meyers)
brunello_classification <- off_target_classification(brunello_alignment)
toronto_classification <- off_target_classification(toronto_alignment)
yusa_classification <- off_target_classification(yusa_alignment)

```

# Stratify multi-target sgRNAs (perfect alignment for each sgRNAs)

```{r}

avana_stratify_multi_target <- off_target_alignment_bin(avana_classification[[1]], c("perfect", "multi-target guides", "non-targeting"))

brunello_stratify_multi_target <- off_target_alignment_bin(brunello_classification[[1]], c("perfect", "multi-target guides", "non-targeting"))

toronto_stratify_multi_target <- off_target_alignment_bin(toronto_classification[[1]], c("perfect", "multi-target guides", "non-targeting"))

yusa_stratify_multi_target <- off_target_alignment_bin(yusa_classification[[1]], c("perfect", "multi-target guides", "non-targeting"))

```

# Export the bar chart showing stratification of multi-target sgRNAs

```{r}
export_images <- function(object, filename_dir, genome_type) {
  ggsave(
    path = './',
    width = 5,
    height = 6,
    dpi = 300,
    plot = object,
    filename = paste(filename_dir, genome_type, ".png", sep=""))
}

export_images(visualize_off_target_alignment(avana_stratify_multi_target[[2]], "Number of on-target alignments"),
              "./plot/avana_stratification_",
              genome_type_version)
export_images(visualize_off_target_alignment(brunello_stratify_multi_target[[2]], "Number of on-target alignments"),
              "./plot/brunello_stratification_",
              genome_type_version)
export_images(visualize_off_target_alignment(toronto_stratify_multi_target[[2]], "Number of on-target alignments"),
              "./plot/toronto_stratification_",
              genome_type_version)
export_images(visualize_off_target_alignment(yusa_stratify_multi_target[[2]], "Number of on-target alignments"),
              "./plot/yusa_stratification_",
              genome_type_version)
```


# What is percentage of sgRNAS that perfectly target multiple genomics sites and protein-coding genes perfectly?

## Identify the target location of multi-target sgRNAs (protein-coding genes/other genomic loci)

```{r}


list_guides <- avana_stratify_multi_target[[1]] %>% filter(alignment_bin %in% c(2,3,4,5,6,7,8,"> 8")) %>% pull(sgRNA)
avana_multi_protein_coding_genes <- off_target_annotation(avana_alignment_meyers,                                                           
                                                          annotation,
                                                          bsgenome,
                                                          genome_type_version,
                                                          avana_library_meyers, 
                                                          list_guides,
                                                          0,
                                                          "avana")

list_guides <- brunello_stratify_multi_target[[1]] %>% filter(alignment_bin %in% c(2,3,4,5,6,7,8,"> 8")) %>% pull(sgRNA)
brunello_multi_protein_coding_genes <- off_target_annotation(brunello_alignment, 
                                                             annotation,
                                                             bsgenome,
                                                             genome_type_version,
                                                             brunello_library, 
                                                             list_guides,
                                                             0,
                                                             "brunello")

list_guides <- toronto_stratify_multi_target[[1]] %>% filter(alignment_bin %in% c(2,3,4,5,6,7,8,"> 8")) %>% pull(sgRNA)
toronto_multi_protein_coding_genes <- off_target_annotation(toronto_alignment, 
                                                            annotation,
                                                            bsgenome,
                                                            genome_type_version,
                                                            toronto_library, 
                                                            list_guides,
                                                            0,
                                                            "toronto")

list_guides <- yusa_stratify_multi_target[[1]] %>% filter(alignment_bin %in% c(2,3,4,5,6,7,8,"> 8")) %>% pull(sgRNA)
yusa_multi_protein_coding_genes <- off_target_annotation(yusa_alignment, 
                                                         annotation,
                                                         bsgenome,
                                                         genome_type_version,
                                                         yusa_library, 
                                                         list_guides,
                                                         0,
                                                         "yusa")

```

## Percentage of multi-target sgRNAs in each libraries and their target locations

```{r}

table_multi_target <- rbind(brunello_multi_protein_coding_genes$freq_table,
                            toronto_multi_protein_coding_genes$freq_table,
                            yusa_multi_protein_coding_genes$freq_table,
                            avana_multi_protein_coding_genes$freq_table)
table_multi_target <- table_multi_target %>%
  arrange(target)


# Grouped bar chart
grouped_barplot <- ggplot(table_multi_target, aes(x = factor(library, levels = c('brunello', 'toronto', 'yusa', 'avana')), y = percent_sgrna_among_library, fill = target)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.9)) +
  theme_minimal() +
  labs(title = "", x = "sgRNA libraries", y = "Multi-target sgRNAs (%)", fill = "Target") +
  scale_x_discrete(labels = c('brunello' = 'Brunello',
                'yusa' = 'Yusa \nProject Score',
                'toronto' = 'Toronto \nKnockout V3',
                'avana' = 'Avana')) +
  scale_fill_manual(labels = c("other genomic loci" = "Other genomic loci",
                               "other protein-coding genes" = "Protein-coding genes"),
                    values = c("other genomic loci" = "#B3589A",
                              "other protein-coding genes" = "#E7F7D5")) +
  theme(axis.text.x = element_text(size = 30, color = "black"),
      axis.text.y = element_text(size = 30, colour = "black"),
      axis.title.x = element_text(size = 0, vjust = -1, margin = margin(r = 10), colour = "black"),
      axis.title.y = element_text(size = 30, margin = margin(r = 10), colour = "black"),
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      axis.line.x = element_line(size = 0.5),
      axis.line.y = element_line(size = 0.5),
      legend.title = element_text(size = 30),
      legend.text = element_text(size = 30)) +
  geom_text(aes(label = paste0(percent_sgrna_among_library, "%")), position = position_dodge(width = 0.8), vjust = -0.5, size = 6.5)  +
  ylim(0, 7)

grouped_barplot

ggsave(
    path = './',
    width = 18,
    height = 15,
    dpi = 300,
    plot = grouped_barplot,
    filename = paste('./plot/percentage_multi_target.png', sep = ''))

```

## Percentage of genes affected by multi-target sgRNAs across libraries

```{r}
# barchart for genes

grouped_barplot <- table_multi_target %>%
  filter(target == "other protein-coding genes") %>%
  ggplot(aes(x = factor(library, levels = c('brunello', 'toronto', 'yusa', 'avana')), y = percent_genes)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.8), fill = "#E7F7D5") +
  theme_minimal() +
  labs(title = "", x = "sgRNA libraries", y = "Genes affected by multi-target sgRNAs (%)") +
  scale_x_discrete(labels = c('brunello' = 'Brunello',
                'yusa' = 'Yusa \nProject Score',
                'toronto' = 'Toronto \nKnockout V3',
                'avana' = 'Avana')) +
  theme(axis.text.x = element_text(size = 15),
      axis.text.y = element_text(size = 15),
      axis.title.x = element_text(size = 18, vjust = -1, margin = margin(r = 10)),
      axis.title.y = element_text(size = 18, margin = margin(r = 10)),
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      axis.line.x = element_line(size = 0.5),
      axis.line.y = element_line(size = 0.5),
      legend.title = element_text(size = 15),
      legend.text = element_text(size = 13)) +
  geom_text(aes(label = paste0(percent_genes, "%")), position = position_dodge(width = 0.8), vjust = -0.5, size = 4.5)

grouped_barplot

ggsave(
    path = './',
    width = 7,
    height = 9,
    dpi = 300,
    plot = grouped_barplot,
    filename = paste('./plot/genes_multi_target.png', sep = ''))
```

# Question 1: Are multitargeting effects more likely to target pairs of paralogs?
# Question 2: Are genes with paralogs more likely to be subject to multi-targeting effects?

## Identify the double-target guides targeting paralog pairs and non-paralog pairs across libraries

```{r}

# Using annotated exons from GENCODE (comprehensive gene annotation, human v28), we found that 2503 (95.2%) of the double-target guides have both targets located in coding regions (Fortin et al, 2019) (totally different than my own findings)

# Among the 2503 double-target guides, 297 (11.9%) guides have their pair of targets annotated as being paralogs using the PANTHER database (Fortin et al, 2019)

                                         
# Avana paralog analysis for double-target guides
list_query_guides_paralog <- avana_stratify_multi_target[[1]] %>% filter(alignment_bin == 2) %>% pull(sgRNA)
avana_multi_paralog <- find_multi_target_paralog(avana_multi_protein_coding_genes[[1]], list_query_guides_paralog, paralog_pairs, avana_library_meyers, "avana")


# Brunello paralog analysis for double-target guides
list_query_guides_paralog <- brunello_stratify_multi_target[[1]] %>% filter(alignment_bin == 2) %>% pull(sgRNA)
brunello_multi_paralog <- find_multi_target_paralog(brunello_multi_protein_coding_genes[[1]], list_query_guides_paralog, paralog_pairs, brunello_library, "brunello")


# toronto paralog analysis for double-target guides
list_query_guides_paralog <- toronto_stratify_multi_target[[1]] %>% filter(alignment_bin == 2) %>% pull(sgRNA)
toronto_multi_paralog <- find_multi_target_paralog(toronto_multi_protein_coding_genes[[1]], list_query_guides_paralog, paralog_pairs, toronto_library, "toronto")


# yusa paralog analysis for double-target guides
list_query_guides_paralog <- yusa_stratify_multi_target[[1]] %>% filter(alignment_bin == 2) %>% pull(sgRNA)
yusa_multi_paralog <- find_multi_target_paralog(yusa_multi_protein_coding_genes[[1]], list_query_guides_paralog, paralog_pairs, yusa_library, "yusa")


# Summary table of double-target sgRNAs targeting paralogs
table_dbl_target_paralog <- rbind(brunello_multi_paralog$summary_count_df,
                                  toronto_multi_paralog$summary_count_df,
                                  yusa_multi_paralog$summary_count_df,
                                  avana_multi_paralog$summary_count_df)


# Barplot of percentage of sgRNA targeting paralog pairs and non-paralog pairs across libraries
barplot_paralog <- table_dbl_target_paralog %>%
  arrange(library_name) %>%
  select(-c(count_sgrna_paralog, count_sgrna_other_genes, count_sgrna_other_genomic_loci)) %>%
  pivot_longer(c(percentage_sgrna_paralog, percentage_sgrna_other_genes, percentage_sgrna_other_genomic_loci), names_to = 'names', values_to = 'percent') %>%
  ggplot(aes(library_name, percent, fill = names)) +
  geom_bar(position = "stack", stat = "identity") +
  theme_minimal() +
  labs(title = "", x = "sgRNA libraries", y = "Percentage of double-target guides (%)", fill = "Target") +
  scale_x_discrete(labels = c('brunello' = 'Brunello',
                'yusa' = 'Yusa \nProject Score',
                'toronto' = 'Toronto \nKnockout V3',
                'avana' = 'Avana')) +
  scale_fill_manual(labels = c("percentage_sgrna_paralog" = "Paralog pairs",
                               "percentage_sgrna_other_genes" = "Other protein-coding genes",
                               "percentage_sgrna_other_genomic_loci" = "Other genomic loci"),
                    values = c("percentage_sgrna_paralog" = "#B3589A",
                              "percentage_sgrna_other_genes" = "#E7F7D5",
                              "percentage_sgrna_other_genomic_loci" = "mistyrose")) +
  theme(axis.text.x = element_text(size = 15),
      axis.text.y = element_text(size = 15),
      axis.title.x = element_text(size = 18, vjust = -1, margin = margin(r = 10)),
      axis.title.y = element_text(size = 18, margin = margin(r = 10)),
      legend.title = element_text(size = 15),
      legend.text = element_text(size = 13)) +
  geom_text(aes(label = paste0(percent, "%")), position = position_stack(vjust = .5), size = 5)

barplot_paralog

ggsave(
    path = './',
    width = 10,
    height = 10,
    dpi = 300,
    plot = barplot_paralog,
    filename = paste('./plot/barplot_double_target_guides_T2T.png', sep = ''))


## Identify genes affected by double-target guides targeting paralog pairs across libraries

paralog_summary_df <- rbind(brunello_multi_paralog$paralog_affected_df,
                                  toronto_multi_paralog$paralog_affected_df,
                                  yusa_multi_paralog$paralog_affected_df,
                                  avana_multi_paralog$paralog_affected_df)



```


## Double-target guides and their sgRNA log-fold-change
## Find "clean" double-target guides
Clean double-target guides should be
1. Multi-target guides that strictly target 2 protein-coding genes
2. Not having additional single mismatches
3. Not having additional double mismatches


```{r}
# fetch perfect alignment at two locations
list_query_guides_paralog <- avana_stratify_multi_target[[1]] %>% filter(alignment_bin == 2) %>% pull(sgRNA)


# fetch double-target guides (without any additional single or double mismatches)
cleaned_double_target_guides <- avana_alignment_meyers %>%
      filter(sgRNA %in% list_query_guides_paralog) %>%
      group_by(sgRNA) %>%
      filter(all(n_mismatches == 0)) %>%
      distinct(sgRNA) %>%
      pull(sgRNA)
  
# annotate the cleaned double-target guides, fetch double-target guides targeting protein-coding genes only
annotate_cleaned_double_target_guides <- off_target_annotation(
                                          avana_alignment_meyers,                                                           
                                          annotation,
                                          bsgenome,
                                          genome_type_version,
                                          avana_library_meyers, 
                                          cleaned_double_target_guides,
                                          0,
                                          "avana")

# fetch double-target guides targeting paralogs
double_target_paralog <- find_multi_target_paralog(annotate_cleaned_double_target_guides[[1]], cleaned_double_target_guides, paralog_pairs, avana_library_meyers, "avana")

# list of cleaned double-target guides targeting paralogs
cleaned_double_target_paralog_guides <- double_target_paralog[[4]]

# dataframe containing the alignment of cleaned double-target guides targeting paralogs
aln_double_target_paralog_guides <- double_target_paralog[[1]] %>% filter(sgrna %in% cleaned_double_target_paralog_guides)


```










































































