---
title: "Backup Code of multi_target_sgrna_analysis"
output: github_document
---
# A venn diagram function to compare the similarity between three libraries

```{r}

venn_diagram <- function(set1, set2, name_set1, name_set2, color1, color2, name_venn) {
  only_set1 <- length(setdiff(set1, set2))
  only_set2 <- length(setdiff(set2, set1))
  both_sets  <- length(intersect(set1, set2))

  totalset1 <- only_set1 + both_sets
  totalset2 <- only_set2 + both_sets

  pct_set1 <- round((only_set1 / totalset1) * 100, 2)
  pct_set2 <- round((only_set2 / totalset2) * 100, 2)
  pct_overlap_set1 <- round((both_sets / totalset1) * 100, 2)
  pct_overlap_set2 <- round((both_sets / totalset2) * 100, 2)

  sgRNA_similarity <- venn.diagram(
        x = list(set1, set2),
        category.names = c(name_set1, name_set2),
        filename = NULL,  
        fill = c(color1, color2),
        alpha = 0.5,
        cex = 0,  # Set text size to 0 to hide numbers
        cat.cex = 0.75,  # Set category label size
        cat.pos = c(-2, 2),  # Adjust category label positions
        cat.dist = c(0.05, 0.05),  # Adjust category label distances
        lwd = 2
  )

  png(filename = name_venn, width = 1250, height = 1250, res = 300)
  grid.newpage()
  grid.draw(sgRNA_similarity)

  grid.text(label = paste0(only_set2, '\n', pct_set1, '%'), x = 0.25, y = 0.50, gp = gpar(fontsize = 10))
  grid.text(label = paste0(only_set1, '\n', pct_set2, '%'), x = 0.75, y = 0.50, gp = gpar(fontsize = 10))
  grid.text(label = paste0(pct_overlap_set1, '%', '\nor\n', pct_overlap_set2, '%'), x = 2.5, y = 0.50, gp = gpar(fontsize = 7))

  dev.off()
}


```

# compare the similarity of all sequences between three libraries

```{r}
venn_diagram(brunello$spacer, toronto$spacer, "Brunello \n 77K sgRNAs", "Toronto V3 \n 71K sgRNAs", 'powderblue', 'mistyrose', 'VennDiagram_sgRNAs_brunello_toronto.png')
venn_diagram(brunello$spacer, cellecta$spacer, "Brunello \n 77K sgRNAs", "Cellecta \n 150K sgRNAs", 'powderblue', 'lightgreen', 'VennDiagram_sgRNAs_brunello_cellecta.png') 
venn_diagram(toronto$spacer, cellecta$spacer, "Toronto V3 \n 77K sgRNAs", "Cellecta \n 150K sgRNAs", 'mistyrose', 'lightgreen', 'VennDiagram_sgRNAs_toronto_cellecta.png') 
```

# off-target guides

```{r}
# get list off target guides
list_off_target_guides <- function(alignment_data) {
  # start with multi-target sgRNA
  multi_target_library <- multi_target_sgrna(alignment_data)
  
  # removal of multi-target sgRNA from the library
  alignment_data_2 <- alignment_data %>%
    filter(!sgRNA %in% multi_target_library$list_guides)
  
  # continue with single mismatch sgRNA
  single_mismatch_library <- single_mismatch_sgrna(alignment_data_2)
  
  # removal of single mismatch sgRNA from the library
  alignment_data_3 <- alignment_data_2 %>%
    filter(!sgRNA %in% single_mismatch_library$list_guides)
  
  # continue with pam double mismatch sgRNA
  pam_double_mismatch_library <- pam_distal_double_mismatch(alignment_data_3)
  
  list_off_target_guides <- unique(union(multi_target_library$list_guides, unique(union(single_mismatch_library$list_guides, pam_double_mismatch_library$list_guides))))
  
  return(list(
    multi_target_guides = unique(multi_target_library$list_guides),
    single_mismatch_guides = unique(single_mismatch_library$list_guides), 
    pam_double_mismatch_guides = unique(pam_double_mismatch_library$list_guides),
    all_off_target_guides = list_off_target_guides))
}
```

# off target annotation

```{r}
# off target annotation (can be multi-target, single-mismatch, or pam-distal double mismatches) used to annotate whether the sgRNA target random genomic loci, protein-coding genes or intended target gene

off_target_annotation <- function(alignment_data, num_mismatch, library, library_name){
  gene_df <- annotate_off_target(alignment_data)
  gene_df <- gene_df %>%
    dplyr::rename('actual_chr' = 'chr',
                  'actual_gene' = 'gene')

  alignment_data <- add_cut_pos_pam_pos(alignment_data)
  alignment_data <- alignment_data %>%
    select(unique_aln_id, sgRNA, spacer, protospacer, gene, chr, n_mismatches, cut_pos) %>%
    dplyr::rename('sgrna' = 'sgRNA')
  
  # one version of alignment data for paralog analysis (you will need the actual gene to detect which multi-target sgRNA targeting only two protein-coding genes - one is the actual gene, the other one is the paralog pairs)
  alignment_data_gene_df_1 <- alignment_data %>%
    left_join(gene_df, by = join_by(unique_aln_id, sgrna, spacer, protospacer, cut_pos)) %>%
    select(unique_aln_id, sgrna, spacer, protospacer, gene, chr, n_mismatches, cut_pos, actual_chr, actual_gene) %>%
    filter(n_mismatches == num_mismatch) %>%
    mutate(target = case_when(gene == actual_gene ~ 'actual gene',
                              gene != actual_gene ~ 'protein-coding genes',
                              is.na(actual_gene) ~ 'other genomic loci'))
  
  # 2nd version of alignment data for frequency table and histogram plot
  alignment_data_gene_df_2 <- alignment_data_gene_df_1 %>%
    filter(target != 'actual gene')
  
  summary_target_df <- alignment_data_gene_df_2 %>%
    select(sgrna, target) %>%
    group_by(sgrna) %>%
    mutate(target = ifelse(n() > 1, 'other genomic loci or protein-coding genes', target)) %>%
    ungroup() %>%
    distinct()
  
  calculate_freq_table <- summary_target_df %>%
    select(target) %>%
    count(target) %>%
    dplyr::rename('count_sgrna' = 'n') %>%
    arrange(desc(count_sgrna)) %>%
    mutate(percent = round(count_sgrna/length(library$sgRNA)*100 ,2),
           library = library_name)

  return(list(
    aln_data_df_1 = alignment_data_gene_df_1,
    aln_data_df_2 = alignment_data_gene_df_2,
    summary_df = summary_target_df, 
    freq_table = calculate_freq_table))
}
```


# Count on-target alignments

```{r}
count_on_target_alignments <- library_alignment %>%
  count(sgRNA, protospacer) %>%
  mutate(n = if_else(is.na(protospacer), 0, n)) %>%
  dplyr::rename("number_of_alignments" = 'n') %>%
  select(-protospacer) %>%
  group_by(sgRNA) %>%
  summarise(number_of_alignments = sum(number_of_alignments), .groups = "drop")

```




## Is the proportion of sgRNAs targeting paralog pairs higher among multi-target sgRNAs compared to non-multi-target ones? (actual question to the Fisher exact test we performed)


```{r}

fisher_multi_target_paralog_genes <- function(off_target_paralog_genes, off_target_not_paralog_genes, normal_paralog_genes, normal_not_paralog_genes, sgrna_library) {

  print(paste('Paralog genes targeted by multi-target sgRNAs:', length(off_target_paralog_genes)))
  print(paste('Non paralog genes targeted by multi-target sgRNAs:', length(off_target_not_paralog_genes)))
  print(paste('Paralog genes not targeted by multi-target sgRNAs:', length(normal_paralog_genes)))
  print(paste('Non paralog genes not targeted by multi-target sgRNAs:', length(normal_not_paralog_genes)))

  # Create the contingency table
  contingency_table <- matrix(
    c(
      length(off_target_paralog_genes),  # Multi-target sgRNAs targeting paralog genes
      length(off_target_not_paralog_genes),  # Multi-target sgRNAs not targeting paralog genes
      length(normal_paralog_genes),  # Non-multi-target sgRNAs targeting paralog genes
      length(normal_not_paralog_genes)  # Non-multi-target sgRNAs not targeting paralog genes
    ),
    nrow = 2
  )

  print(contingency_table)

  # Check if the total matches the unique sgRNAs in alignment_data
  total_contingency_table <- sum(contingency_table)
  total_unique_gene <- length(unique(sgrna_library$gene))

  print(paste('total sgRNA count in contingency table:', total_contingency_table))
  print(paste('total original number of gene:', total_unique_gene))

  if (total_contingency_table == total_unique_gene) {
    print("The sum of the contingency table matches the total unique gene in sgRNA library.")
  } else {
    print("Mismatch detected between the contingency table sum and the total unique gene in sgRNA library")
  }

  # Perform Fisher's test
  fischer_library_multi_target <- fisher.test(contingency_table)
  print(fischer_library_multi_target)

}

gene_paralog_target <- function(sgrna_library, gene_paralog_multi_target, gene_not_paralog_multi_target, barbara_paralog_dataset) {
  
  # Load paralog gene pairs
  barbara_paralog <- read_csv(barbara_paralog_dataset)
  
  # Extract the vector of paralog genes from column A2
  paralog_genes <- pull(barbara_paralog, A2)
  
  # Genes that are paralogs but not multi-target or not in the not-paralog list
  gene_paralog_normal_target <- sgrna_library %>%
    filter(
      gene %in% paralog_genes,
      !gene %in% gene_paralog_multi_target,
      !gene %in% gene_not_paralog_multi_target
    )
  
  # Genes that are NOT paralogs, and also not multi-target or in not-paralog list
  gene_not_paralog_normal_target <- sgrna_library %>%
    filter(
      !gene %in% paralog_genes,
      !gene %in% gene_paralog_multi_target,
      !gene %in% gene_not_paralog_multi_target
    )
  
  return(list(
    paralog_normal_target = unique(gene_paralog_normal_target$gene),
    not_paralog_normal_target = unique(gene_not_paralog_normal_target$gene)
  ))
}


# Function to identify which sgRNAs among all libraries that are not targeting paralog pairs

find_not_multi_target_paralog <- function(sgrna_library, alignment_gene_df, paralog_dataset) {
  
  # read paralog dataset
  paralog <- read_csv(paralog_dataset)
  
  # non multi-target sgRNA in the library
  sgrna_library_not_multi_target <- sgrna_library %>%
    filter(!sgRNA %in% alignment_gene_df$sgrna)
  
  # normal sgRNA targeting paralog
  sgrna_library_paralog <- sgrna_library_not_multi_target %>%
    filter(gene %in% unique(c(paralog$A2, paralog$A1)))
  
  # normal sgRNA not targeting paralog
  sgrna_library_not_paralog <- setdiff(unique(sgrna_library_not_multi_target$sgRNA), unique(sgrna_library_paralog$sgRNA))
  
  return (list(normal_sgrna_paralog = unique(sgrna_library_paralog$sgRNA),
               normal_sgrna_not_paralog = sgrna_library_not_paralog))
  
}


brunello_paralog_normal <- gene_paralog_target(brunello, brunello_gene_paralog_multi_target, brunello_gene_not_paralog_multi_target, '../barbara_36K_paralog_pairs.csv')

brunello_gene_fisher <- fisher_multi_target_paralog_genes(brunello_gene_paralog_multi_target, 
                                                          brunello_gene_not_paralog_multi_target,
                                                          brunello_paralog_normal$paralog_normal_target,
                                                          brunello_paralog_normal$not_paralog_normal_target,
                                                          brunello)


```


### What is percentage of sgRNAs that target multiple genomic sites with mismatches?
### What is percentage of sgRNAs that target multiple protein-coding genes with mismatches?

```{r}

avana_stratify_single_mismatch <- off_target_alignment_bin(avana_classification[[1]], c("single mismatch"))
visualize_off_target_alignment(avana_stratify_single_mismatch[[2]], "Number of single mismatch off-target alignments")

brunello_stratify_single_mismatch <- off_target_alignment_bin(brunello_classification[[1]], c("single mismatch"))
visualize_off_target_alignment(brunello_stratify_single_mismatch[[2]], "Number of single mismatch off-target alignments")

toronto_stratify_single_mismatch <- off_target_alignment_bin(toronto_classification[[1]], c("single mismatch"))
visualize_off_target_alignment(toronto_stratify_single_mismatch[[2]], "Number of single mismatch off-target alignments")

yusa_stratify_single_mismatch <- off_target_alignment_bin(yusa_classification[[1]], c("single mismatch"))
visualize_off_target_alignment(yusa_stratify_single_mismatch[[2]], "Number of single mismatch off-target alignments")
  

```


```{r}
list_guides <- avana_stratify_single_mismatch[[1]] %>% filter(alignment_bin %in% c(1,2,3,4,5,6,7,8,"> 8")) %>% pull(sgRNA)
avana_single_mismatch_coding_genes <- off_target_annotation(avana_alignment_meyers, 
                                                            avana_library_meyers, 
                                                            list_guides,
                                                            1,
                                                            "avana")

list_guides <- brunello_stratify_single_mismatch[[1]] %>% filter(alignment_bin %in% c(1,2,3,4,5,6,7,8,"> 8")) %>% pull(sgRNA)
brunello_single_mismatch_coding_genes <- off_target_annotation(brunello_alignment, 
                                                            brunello_library, 
                                                            list_guides,
                                                            1,
                                                            "brunello")

list_guides <- toronto_stratify_single_mismatch[[1]] %>% filter(alignment_bin %in% c(1,2,3,4,5,6,7,8,"> 8")) %>% pull(sgRNA)
toronto_single_mismatch_coding_genes <- off_target_annotation(toronto_alignment, 
                                                            toronto_library, 
                                                            list_guides,
                                                            1,
                                                            "toronto")

list_guides <- yusa_stratify_single_mismatch[[1]] %>% filter(alignment_bin %in% c(1,2,3,4,5,6,7,8,"> 8")) %>% pull(sgRNA)
yusa_single_mismatch_coding_genes <- off_target_annotation(yusa_alignment, 
                                                            yusa_library, 
                                                            list_guides,
                                                            1,
                                                            "yusa")

table_single_mismatch_target <- rbind(brunello_single_mismatch_coding_genes$freq_table,
                            toronto_single_mismatch_coding_genes$freq_table,
                            yusa_single_mismatch_coding_genes$freq_table,
                            avana_single_mismatch_coding_genes$freq_table)
table_single_mismatch_target <- table_single_mismatch_target %>%
  arrange(target)

# Grouped bar chart
grouped_barplot <- ggplot(table_single_mismatch_target, aes(x = factor(library, levels = c('brunello', 'toronto', 'yusa', 'avana')), y = percent_sgrna_among_library, fill = target)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.9)) +
  theme_minimal() +
  labs(title = "", x = "sgRNA libraries", y = "Single mismatch off-target sgRNAs (%)", fill = "Target") +
  scale_x_discrete(labels = c('brunello' = 'Brunello',
                'yusa' = 'Yusa \nProject Score',
                'toronto' = 'Toronto \nKnockout V3',
                'avana' = 'Avana')) +
  scale_fill_manual(labels = c("other genomic loci" = "Other genomic loci",
                               "other protein-coding genes" = "Protein-coding genes"),
                    values = c("other genomic loci" = "#B3589A",
                              "other protein-coding genes" = "#E7F7D5")) +
  theme(axis.text.x = element_text(size = 30, color = "black"),
      axis.text.y = element_text(size = 30, colour = "black"),
      axis.title.x = element_text(size = 0, vjust = -1, margin = margin(r = 10), colour = "black"),
      axis.title.y = element_text(size = 30, margin = margin(r = 10), colour = "black"),
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      axis.line.x = element_line(size = 0.5),
      axis.line.y = element_line(size = 0.5),
      legend.title = element_text(size = 30),
      legend.text = element_text(size = 30)) +
  geom_text(aes(label = paste0(percent_sgrna_among_library, "%")), position = position_dodge(width = 0.8), vjust = -0.5, size = 6.5)  +
  ylim(0, 7)

grouped_barplot

ggsave(
    path = './',
    width = 18,
    height = 15,
    dpi = 300,
    plot = grouped_barplot,
    filename = paste('./plot/percentage_single_mismatch.png', sep = ''))

```

## Double-target guides and their sgRNA log-fold-change
## Find "clean" double-target guides
Clean double-target guides should be
1. Multi-target guides that strictly target 2 protein-coding genes
2. Not having additional single mismatches
3. Not having additional double mismatches

Afterwards we try to find where guide that co-targeting a pair of paralog genes (clean double-target guides) and compare it with at least one single-target guide for each of the two paralogs


```{r}

clean_double_target_guides <- function(list_guides, alignment_data, library_data, library_name, refined_library_dir) {
  
  # fetch double-target guides (without any additional single or double mismatches)
  cleaned_double_target_guides <- alignment_data %>%
        filter(sgRNA %in% list_guides) %>%
        group_by(sgRNA) %>%
        filter(all(n_mismatches == 0)) %>%
        distinct(sgRNA) %>%
        pull(sgRNA)
    
  # annotate the cleaned double-target guides, fetch double-target guides targeting protein-coding genes only
  annotate_cleaned_double_target_guides <- off_target_annotation(
                                            alignment_data,                                       
                                            annotation,
                                            bsgenome,
                                            genome_type_version,
                                            library_data, 
                                            cleaned_double_target_guides,
                                            0,
                                            library_name)
  
  # fetch double-target guides targeting paralogs
  double_target_paralog <- find_multi_target_paralog(annotate_cleaned_double_target_guides[[1]], 
                                                     cleaned_double_target_guides, 
                                                     paralog_pairs, 
                                                     library_data, 
                                                     library_name)
  
  # list of cleaned double-target guides targeting paralogs
  cleaned_double_target_paralog_guides <- double_target_paralog[[4]]
  
  # list of paralog genes affected by double-target guides
  paralog_affected_double_target_guides <- double_target_paralog[[7]]
  
  # import refined library and filter for gene paralog that affected by double-target guides
  # this is to filter the paralog gene that still has "clean" sgRNA upon refinement
  refined_library <- import_library(refined_library_dir) %>%
    filter(gene %in% paralog_affected_double_target_guides)
  
  # dataframe containing the alignment of cleaned double-target guides targeting paralogs
  # this is to filter for cleaned double-target-guides and gene in the refined library
  aln_double_target_paralog_guides <- double_target_paralog[[1]] %>% 
    filter(sgrna %in% cleaned_double_target_paralog_guides & gene %in% refined_library$gene)
    # select(sgrna, spacer, gene, actual_gene, target)
  
  # add another filter to fetch genes that match the criteria from aln_double_target_paralog_guides
  refined_library <- refined_library %>%
    filter(gene %in% unique(aln_double_target_paralog_guides$gene))
  
  combined_df <- bind_rows(refined_library, aln_double_target_paralog_guides)
  combined_df <- combined_df %>% 
    arrange(sgrna, gene)
    
  return(list(
    double_target_paralog = double_target_paralog,
    dbl_target_paralog_guides = cleaned_double_target_paralog_guides, 
    paralog_affected = paralog_affected_double_target_guides,
    df_single_target_guides = refined_library,
    df_co_targeting_guides = aln_double_target_paralog_guides,
    combined_df = combined_df))

}

# fetch perfect alignment at two locations
list_query_guides_paralog <- avana_stratify_multi_target[[1]] %>% filter(alignment_bin == 2) %>% pull(sgRNA)

avana_dbl_target_guides <- clean_double_target_guides(list_query_guides_paralog, 
                               avana_alignment_meyers, 
                               avana_library_meyers, 
                               "avana",
                               "../CRISPR-KO-library-refinement-pipeline/output_cleaning/hg38/avana_library_56K_refined.tsv")


list_query_guides_paralog <- brunello_stratify_multi_target[[1]] %>% filter(alignment_bin == 2) %>% pull(sgRNA)

brunello_dbl_target_guides <- clean_double_target_guides(list_query_guides_paralog, 
                               brunello_alignment, 
                               brunello_library, 
                               "brunello",
                               "../CRISPR-KO-library-refinement-pipeline/output_cleaning/hg38/broadgpp-brunello-library-contents_65K_refined.tsv")


list_query_guides_paralog <- toronto_stratify_multi_target[[1]] %>% filter(alignment_bin == 2) %>% pull(sgRNA)

toronto_dbl_target_guides <- clean_double_target_guides(list_query_guides_paralog, 
                               toronto_alignment, 
                               toronto_library, 
                               "toronto",
                               "../CRISPR-KO-library-refinement-pipeline/output_cleaning/hg38/tkov3_guide_sequence_62K_refined.tsv")


list_query_guides_paralog <- yusa_stratify_multi_target[[1]] %>% filter(alignment_bin == 2) %>% pull(sgRNA)

yusa_dbl_target_guides <- clean_double_target_guides(list_query_guides_paralog, 
                               yusa_alignment, 
                               yusa_library, 
                               "yusa",
                               "../CRISPR-KO-library-refinement-pipeline/output_cleaning/hg38/yusa_hcrispr_ko_grnas_76K_refined.tsv")



```


