---
title: "Analysis of gene hits change in Olivieri CRISPR-KO screen"
author: "Author: Stefanus Bernard"
date: "Date: `r format(Sys.time(), '%d %B %Y')`"
output: 
  html_document :
    theme : cerulean
    toc : true
    toc_float : true
    number_section : true
---

Import relevant library

```{r}
library(tidyverse)
library(dplyr)
library(readxl)
library(clusterProfiler)
library(enrichplot)
library(org.Hs.eg.db)
source("../00_CRISPR-KO-library-survey-functions.R")
```

Import TKOV3 sgRNA cleaning report and background genes

```{r}

TKOV3_report <- read_excel("../../../GuideRefine/output_cleaning/T2T-CHM13/tkov3_guide_sequence_full_report.xlsx")

original_library <- read_delim("../../data/library_data/original_library/tkov3_guide_sequence.tsv", col_names = c("sgRNA", "spacer", "gene")) %>% filter(!gene %in% c("LacZ", "luciferase", "EGFR")) %>% pull(gene) %>% unique()

refined_library <- read_delim("../../data/library_data/refined_library/tkov3_guide_sequence_61K_refined.tsv", col_names = c("sgRNA", "spacer", "gene")) %>% filter(!gene %in% c("LacZ", "luciferase", "EGFR")) %>% pull(gene) %>% unique()



# obtain alignment type data

pick_alignment_type <- function(classification_data, alignment_type, stratification_data) {
  data <- read_csv(classification_data) %>%
    filter(alignment %in% alignment_type)
  
  return(data)
}

selected_guides <- c("perfect", "single mismatch", "pam-distal single mismatch", "pam-distal double mismatch", "multi-target guides", "non-targeting")

tkov3_classification <- pick_alignment_type("../../data/sgrna_lfc_data/tkov3_data/tkov3_sgrna_alignment_classification.csv", selected_guides)
tkov3_classification <- count_alignment_bin(tkov3_classification, selected_guides)

```

Check rank of gene hits before and after

```{r}
hits_original <- read_tsv("../../data/result_gene_hits_change/tkov3_olivieri_paper/AZD6738_output_original.txt")
hits_refined <- read_tsv("../../data/result_gene_hits_change/tkov3_olivieri_paper/AZD6738_output_refined.txt")
```

Take an example of one of the hits MNNG

```{r}

AZD6738_original_only_hits <- read_excel("../../data/result_gene_hits_change/olivieri_hits_original_refined_approved_symbol.xlsx") %>%
  dplyr::select(AZD6738_original_only_hits) %>%
  drop_na() %>%
  pull(AZD6738_original_only_hits)

AZD6738_refined_only_hits <- read_excel("../../data/result_gene_hits_change/olivieri_hits_original_refined_approved_symbol.xlsx") %>%
  dplyr::select(AZD6738_refined_only_hits) %>%
  drop_na() %>%
  pull(AZD6738_refined_only_hits)

AZD6738_overlap_hits <- read_excel("../../data/result_gene_hits_change/olivieri_hits_original_refined_approved_symbol.xlsx") %>%
  dplyr::select(AZD6738_overlap_hits) %>%
  drop_na() %>%
  pull(AZD6738_overlap_hits)

```

Does the genes only represented in the original library removed because multi-target-/off-target or because of rank changes

```{r}
# check HGNC data

hgnc_protein_coding <- read_tsv("https://storage.googleapis.com/public-download-files/hgnc/tsv/tsv/hgnc_complete_set.txt") %>%
    filter(locus_group == "protein-coding gene" & !is.na(ensembl_gene_id)) %>%
    dplyr::rename("gene" = "symbol") %>%
    dplyr::select(gene, name, locus_group, ensembl_gene_id)

selected_gene <- hgnc_protein_coding %>%
  filter(gene %in% AZD6738_original_only_hits) %>%
  mutate(notes = case_when(gene %in% hits_refined$GENE ~ "rank change",
                           TRUE ~ "removed due to multi-target-/off-target sgrnas")) %>%
  dplyr::rename("GENE" = "gene") %>%
  left_join(hits_original, join_by(GENE))


```

# What are the genes removed because of multi-target-/off-target

```{r}

selected_gene_removed_multi_target <- selected_gene %>% filter(!notes == "rank change")

guides_observation <- tkov3_classification[[1]] %>% filter(gene %in% selected_gene_removed_multi_target$GENE)

guides_observation %>% 
  dplyr::select(gene, alignment) %>%
  group_by(gene) %>%
  count(alignment) %>%
  dplyr::rename("num_guides" = "n") %>%
  filter(!alignment == "perfect") %>%
  mutate(percentage = round((num_guides/4)*100, 2),
         alignment = factor(alignment, levels = c("pam-distal single mismatch", "single mismatch", "multi-target guides"))) %>%
  ggplot(aes(gene, percentage, fill = alignment)) +
    geom_bar(stat = "identity", position = "stack") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45)) +
    ylim(0, 100)

```

Does all the genes represented in original library actually driven by multi-target-/off-target effect

```{r}
# obtain all gene hits represented only in the original library
data_hits <- read_excel("../../data/result_gene_hits_change/olivieri_hits_original_refined_approved_symbol.xlsx") %>%
  dplyr::select(matches("_original_only_hits")) %>%
  pivot_longer(
    cols = matches("_original_only_hits"),
    names_to = "screens",
    values_to = "gene_hits"
  ) %>%
  drop_na()

```





