---
title: "Analysis of Brunello and Jacquere A375 cells hits"
author: "Author: Stefanus Bernard"
date: "Date: `r format(Sys.time(), '%d %B %Y')`"
output: 
  html_document :
    theme : cerulean
    toc : true
    toc_float : true
    number_section : true
---

# Import packages

```{r}
library(MAGeCKFlute)
library(tidyverse)
library(ggrepel)
library(ggExtra)
library(hgnc)
library(HGNChelper)
library(purrr)
library(xlsx)
library(readxl)
library(pheatmap)
library(stringr)
library(ggpubr)
source("../00_CRISPR-KO-library-survey-functions.R")
```

# Important function

```{r}

# Function to call gene hits from MAGeCK MLE (2*standard deviation)

CallMAGeCKHits <- function(gene_summary, scale = 2, beta_colnames) {

  # 1. Mean beta across replicates
  gene_summary$Diff <- rowMeans(
    gene_summary[, beta_colnames]
  )

  # 2. SD-based cutoff
  cc <- CutoffCalling(gene_summary$Diff, scale = scale)

  # 3. Initial classification
  gene_summary$hits_classification <- "Neutral"
  gene_summary$hits_classification[gene_summary$Diff <= -cc] <- "Essential genes"
  gene_summary$hits_classification[gene_summary$Diff >= cc] <- "Growth suppressor genes"

  # 4. Enforce replicate consistency
  gene_summary$hits_classification[
    gene_summary$hits_classification == "Essential genes" &
      !(gene_summary$repA < 0 & gene_summary$repB < 0)
  ] <- "Neutral"

  gene_summary$hits_classification[
    gene_summary$hits_classification == "Growth suppressor genes" &
      !(gene_summary$repA > 0 & gene_summary$repB > 0)
  ] <- "Neutral"

  return(gene_summary)
}


# pick alignment type

pick_alignment_type <- function(classification_data, alignment_type, stratification_data) {
  data <- read_csv(classification_data) %>%
    filter(alignment %in% alignment_type)
  
  return(data)
}


```

# Import MAGeCK MLE gene summary result

```{r}

brunello_original <- read_tsv("../../data/read_count_data/A375_brunello_original/A375_brunello_original.gene_summary.txt")
brunello_refined <- read_tsv("../../data/read_count_data/A375_brunello_refined/A375_brunello_refined.gene_summary.txt")

# jacquere_original <- read_tsv("../../data/read_count_data/A375_jacquere_original/A375_jacquere_original.gene_summary.txt")
# jacquere_refined <- read_tsv("../../data/read_count_data/A375_jacquere_refined/A375_jacquere_63K_refined.gene_summary.txt")

```

# Find hits

```{r}
brunello_original_hits <- CallMAGeCKHits(brunello_original, scale = 2, beta_colnames = c("RepA|beta", "RepB|beta", "RepC|beta"))
brunello_original_hits <- brunello_original_hits %>%
  filter(`RepA|wald-fdr` <= 0.05 & `RepB|wald-fdr` <= 0.05 & `RepC|wald-fdr` <= 0.05)

brunello_refined_hits <- CallMAGeCKHits(brunello_refined, scale = 2, beta_colnames = c("RepA|beta", "RepB|beta", "RepC|beta"))
brunello_refined_hits <- brunello_refined_hits %>%
  filter(`RepA|wald-fdr` <= 0.05 & `RepB|wald-fdr` <= 0.05 & `RepC|wald-fdr` <= 0.05)

# jacquere_original_hits <- CallMAGeCKHits(jacquere_original, scale = 2, beta_colnames = c("A375_CP0082_repA|beta", "A375_CP0082_repB|beta"))
# jacquere_original_hits <- jacquere_original_hits %>%
#   filter(`A375_CP0082_repA|wald-fdr` <= 0.05 & `A375_CP0082_repB|wald-fdr` <= 0.05)
# 
# jacquere_refined_hits <- CallMAGeCKHits(jacquere_refined, scale = 2, beta_colnames = c("A375_CP0082_repA|beta", "A375_CP0082_repB|beta"))
# jacquere_refined_hits <- jacquere_refined_hits %>%
#   filter(`A375_CP0082_repA|wald-fdr` <= 0.05 & `A375_CP0082_repB|wald-fdr` <= 0.05)

```

# Pull hits

```{r}

brunello_original_hits_list <- fix_gene_symbols(brunello_original_hits$Gene) %>% pull(Suggested.Symbol) %>% unique()
brunello_refined_hits_list <- fix_gene_symbols(brunello_refined_hits$Gene) %>% pull(Suggested.Symbol) %>% unique()

brunello_original_only_hits <- setdiff(brunello_original_hits_list, brunello_refined_hits_list)


# jacquere_original_hits_list <- fix_gene_symbols(jacquere_original_hits$Gene) %>% pull(Suggested.Symbol) %>% unique()
# jacquere_refined_hits_list <- jacquere_refined_hits %>% pull(Gene)
# 
# jacquere_original_only_hits <- setdiff(jacquere_original_hits_list, jacquere_refined_hits_list)
# 
# combined_vecs <- list(brunello_original_hits = brunello_original_hits_list,
#                          brunello_refined_hits = brunello_refined_hits_list,
#                          jacquere_original_hits = jacquere_original_hits_list,
#                          jacquere_refined_hits = jacquere_refined_hits_list)
# 
# max_len <- max(lengths(combined_vecs))
# 
# 
# final_df <- as.data.frame(
# lapply(combined_vecs, function(x) {
# length(x) <- max_len
# x
# })
# )

# write_csv(final_df, "../../data/result_gene_hits_change/brunello_jacquere_A375_hits_change.tsv")

```

## Survey of multi-targeting and off-targeting guides in Brunello and Jacquere

### Isolate CRISPR-KO screens that used A375 cells from DepMap

```{r}

# There are 4 A375 cell lines in Avana DepMap, however only one cell line ACH-000219 that use 

# import model screen data (cell line name and model ID)
model_data <- read_csv("../../data/hits_change_data/Model_25Q2_11022026.csv") %>%
  filter(str_detect(CellLineName, "A375|A-375")) %>%
  select(ModelID, CellLineName)

# import gene effect score from 25Q2
gene_effect <- read_csv("../../data/hits_change_data/CRISPRGeneEffect_25Q2_11022026.csv") %>%
  dplyr::rename("sample" = `...1`) %>%
  rename_with(~ str_remove(.x, "\\s*\\(\\d+\\)")) %>%
  filter(sample == "ACH-000219") %>%
  as.matrix() %>%
  t() %>%
  as.data.frame() %>%
  drop_na()

# select essential genes with gene effect score < -0.5
A375_essential <- gene_effect %>%
  dplyr::rename("ACH-000219" = "V1") %>%
  dplyr::slice(-1) %>%
  rownames_to_column("gene") %>%
  mutate(`ACH-000219` = as.numeric(`ACH-000219`)) %>%
  filter(`ACH-000219` < -0.5)

```

### Brunello analysis of genes with less than -0.5

```{r}

selected_guides <- c("perfect", "single mismatch", "pam-distal double mismatch", "multi-target guides", "non-targeting")

brunello_classification <- pick_alignment_type("../../data/sgrna_lfc_data/brunello_data/brunello_sgrna_alignment_classification.csv", selected_guides)
brunello_classification <- count_alignment_bin(brunello_classification, selected_guides)

# ------------------------------------------------------------------------------------------------------------------------------

# beta score of 893 original gene hits
brunello_essentiality_profiles <- brunello_original_hits %>%
  select(Gene, Diff, hits_classification) %>%
  dplyr::rename("gene" = "Gene")

# left join the sgRNA alignment profiles with the beta score (brunello_essentiality_profiles); total genes is 302
brunello_original_only_hits_analysis <- brunello_classification[[1]] %>%
  filter(gene %in% brunello_original_only_hits) %>%
  left_join(brunello_essentiality_profiles, join_by(gene)) %>%
  mutate(is_depmap_essential = ifelse(gene %in% A375_essential$gene, "A375_essential", "A375_not_essential")) %>%
  select(-spacer)

# Not essential genes but hits in brunello library (53 genes)
not_essential_but_hits_brunello <- brunello_original_only_hits_analysis %>%
  filter(is_depmap_essential == "A375_not_essential")

# Essential genes and hits in brunello library (249 genes)
essential_and_hits_brunello <- brunello_original_only_hits_analysis %>%
  filter(is_depmap_essential == "A375_essential")

# ------------------------------------------------------------------------------------------------------------------------------

# plot survey of essential and not essential genes but hits in Brunello library
survey <- brunello_original_only_hits_analysis %>%
  select(gene, alignment, alignment_bin, Diff) %>%
  group_by(gene) %>%
  count(alignment, alignment_bin) %>%
  dplyr::rename("num_guides" = "n")  %>%
  mutate(is_depmap_essential = ifelse(gene %in% A375_essential$gene, "A375_essential", "A375_not_essential"),
         is_depmap_essential = factor(is_depmap_essential, levels = c("A375_not_essential", "A375_essential"))) %>%
  ungroup(gene)

survey_plot <- survey %>%
  select(gene, is_depmap_essential) %>%
  distinct() %>%
  count(is_depmap_essential) %>%
  dplyr::rename("num_genes" = "n") %>%
  ggplot(aes(is_depmap_essential, num_genes)) +
     geom_col(fill = "#1E88E5") +
     theme_minimal() +
     labs(x = "A375 essential genes", y = "Number of gene hits") +
     scale_x_discrete(labels=c("A375 not essential", "A375 essential")) +
     geom_text(aes(label = paste("n= ", num_genes, sep="")),
                nudge_y = 10,
                nudge_x = 0,
                size = 8) +
  theme(axis.text.x = element_text(size = 25, vjust = 0.7, colour = 'black'),
          axis.text.y = element_text(size = 25, colour = "black"),
          axis.title.x = element_blank(),
          axis.title.y = element_text(size = 25, vjust = 0.5, margin = margin(r = 20), colour = "black"),
          axis.ticks.y = element_line(size = 1.5),
          axis.ticks.length = unit(0.3, "cm"),
          axis.line.y.right = element_blank(),
          axis.ticks.y.right = element_blank(), 
          axis.text.y.right = element_blank(), 
          panel.grid.major = element_blank(),
          panel.grid.minor = element_blank(),
          axis.line.x = element_line(size = 0.5, colour = "black"),
          axis.line.y = element_line(size = 0.5, colour = "black"),
          legend.position = "none",
          plot.margin = margin(1, 1, 1, 1),
          legend.spacing.x = unit(1, 'cm'))

survey_plot

ggsave(
    path = './',
    width = 8,
    height = 5,
    dpi = 1000,
    plot = survey_plot,
    filename = "../../figure/04_fig_analysis_gene_hits_change/brunello_not_essential_hits_barplot.png")

```

# Analysis of sgRNA Log2FC for gene hits

```{r}

brunello_library_and_lfc_data <- obtain_required_data("../../data/library_data/original_library/broadgpp-brunello-library-contents.tsv", 
                                                     "../../data/sgrna_lfc_data/output_normalized/brunello_a375_normalized_lfc.csv", 
                                                     brunello_classification[[1]])

# obtain only sgRNA and log2FC data
brunello_sgrna_and_lfc <- brunello_library_and_lfc_data %>%
  select(sgRNA, mean)

# TODO: some sgRNA Log2FC data for Brunello is missing because of T0 filtering
# Left join sgRNA alignment profiles + MAGeCK MLE score with brunello log2FC data
brunello_original_only_hits_analysis <- brunello_original_only_hits_analysis %>%
  left_join(brunello_sgrna_and_lfc, join_by("sgRNA")) %>%
  # drop sgrna with missing log2fc
  drop_na() %>%
  mutate(is_depmap_essential = paste0(is_depmap_essential, "_hits_only_in_brunello"),
         is_depmap_essential = factor(is_depmap_essential, levels = c("A375_not_essential_hits_only_in_brunello", "A375_essential_hits_only_in_brunello")))


brunello_original_only_hits_lfc <- brunello_original_only_hits_analysis %>%
  select(sgRNA, gene, mean, alignment, num_alignments, alignment_bin, is_depmap_essential)

# take non-essential genes Log2FC data as the control measurement; this is taken from whole brunello library and log2FC data and the data were filtered by taking sgRNA targeting original only hits (302 genes), genes that are essential in A375 and is perfect guides

brunello_non_essential_genes_control <- brunello_library_and_lfc_data %>%
  filter(!sgRNA %in% brunello_original_only_hits_analysis$sgRNA & !gene %in% A375_essential$gene & alignment == "perfect") %>%
  select(-spacer) %>%
  mutate(is_depmap_essential = "A375_not_essential_control")


# combine brunello data from original only hits (302 genes), and non essential genes that are control (16,395 genes)
combined_brunello <- rbind(brunello_original_only_hits_lfc, brunello_non_essential_genes_control) %>%
  mutate(is_depmap_essential = factor(is_depmap_essential, levels = c(
    "A375_not_essential_control",
    "A375_not_essential_hits_only_in_brunello", 
    "A375_essential_hits_only_in_brunello")))
  
```

## 1st Visualization: Boxplot Log2FC between essential hits and non essential but gene hits in Brunello

```{r}

boxplot_lfc <- combined_brunello %>%
  ggboxplot(x = "is_depmap_essential", 
            y = "mean",
            outliers = FALSE) +
    stat_compare_means(comparisons = list(c("A375_not_essential_control", "A375_not_essential_hits_only_in_brunello"),
                                          c("A375_not_essential_control", "A375_essential_hits_only_in_brunello")),
                       method = "wilcox.test",
                       method.args = list(alternative = "two.sided"),
                       size = 2.5,
                       step.increase = 0.05,
                       tip.length = 0.01,
                       label = "p.signif",
                       label.y = 1) +
  labs(title = "", 
         x = "Gene hits", 
         y = "Median sgRNA Log2FC") +
    theme(plot.margin = margin(10, 10, 25, 10),
          strip.text.x = element_text(size = 14),
          axis.text.y = element_text(size = 14, colour = "black"),
          axis.ticks.y = element_line(size = 1.5),
          axis.ticks.x = element_blank(), 
          axis.ticks.length = unit(0.1, "cm"),
          axis.title.x = element_blank(),
          axis.title.y = element_text(size = 14, vjust = 0.5, margin = margin(r = 20)),
          panel.grid.major = element_line(size = 0.5),
          panel.grid.minor = element_blank(),
          axis.line.x = element_line(size = 0.5),
          axis.line.y = element_line(size = 0.5), 
          line = element_line((size = 2), colour = 'black'),
          legend.position = "right",
          legend.text = element_text(size = 12)) +
  scale_x_discrete(labels = c(paste0(length(unique(brunello_non_essential_genes_control$gene))," genes\n(A375 not essential)"), 
                              paste0(length(unique(not_essential_but_hits_brunello$gene))," gene hits \n(A375 not essential)"), 
                              paste0(length(unique(essential_and_hits_brunello$gene))," gene hits \n(A375 essential)"))) +
  scale_y_continuous(limits = c(-3, 2), breaks = seq(-2, 2, by = 1))

 # get stats
  
  b <- ggplot_build(boxplot_lfc)
  stats <- b$data[[1]]
  
  label_df <- stats %>%
    transmute(
      x = x,
      y = ymax + 0.2,          # add padding
      label = round(ymax, 2)   # or any custom label
    )
  
  count <- combined_brunello %>%
    select(is_depmap_essential) %>%
    count(is_depmap_essential)
  
  count <- count %>% bind_cols(stats %>% select(x, ymin))
  
  boxplot_lfc <-  boxplot_lfc +
                                geom_text(
                                  data = count,
                                  aes(x = x, y = ymin - 0.5, label = paste0("n = ", n)),
                                  angle = 45,
                                  vjust = 0,
                                  hjust = 0.5,
                                  size = 5
                                )
  
boxplot_lfc

ggsave(
    path = './',
    width = 6,
    height = 5,
    dpi = 1000,
    plot = boxplot_lfc,
    filename = "../../figure/04_fig_analysis_gene_hits_change/brunello_not_essential_hits_lfc.png")

```

## 2nd Visualization: Multi bar chart of percentage of sgRNA type between essential gene hits and non essential but hits

```{r}

percentage_data <- combined_brunello %>%
  filter(is_depmap_essential %in% c("A375_not_essential_hits_only_in_brunello", "A375_essential_hits_only_in_brunello")) %>%
  group_by(is_depmap_essential) %>%
  count(alignment) %>%
  dplyr::rename("num_of_guides" = "n") %>%
  mutate(percentage = round(num_of_guides/sum(num_of_guides) * 100, 2)) %>%
  ungroup() %>%
  mutate(alignment = factor(alignment, levels = c("perfect", "multi-target guides", "single mismatch", "pam-distal double mismatch")))

# Define custom colors for the libraries
my_palette <- brewer.pal(n = 5, name = "Set2")

# Create the grouped multibar chart for each sgRNA libraries
multi_bar_chart <- ggplot(percentage_data, aes(x = is_depmap_essential, y = percentage, fill = alignment)) +
    geom_bar(stat = "identity", position = position_dodge2(padding = 0.2), width = 0.7) +
    theme_minimal() +
    labs(title = "", x = "", y = "sgRNA (%)") +
    scale_x_discrete(labels = c('A375_not_essential_hits_only_in_brunello' = 'A375 not essential \n gene hits',
                                  'A375_essential_hits_only_in_brunello' = 'A375 essential \n gene hits')) +
    geom_text(aes(label = percentage), position = position_dodge2(width = 0.7), vjust = -0.5, size = 6.5) +
    theme(axis.text.x = element_text(size = 25, vjust = 0.7, colour = "black"),
          axis.text.y = element_text(size = 25, colour = "black"),
          axis.title.y = element_text(size = 25, vjust = 0.5, margin = margin(r = 20)),
          panel.grid.major = element_blank(),
          panel.grid.minor = element_blank(),
          axis.line.x = element_line(size = 0.5),
          axis.line.y = element_line(size = 0.5),
          axis.ticks.y = element_line(size = 1.5),
          axis.ticks.length = unit(0.3, "cm"),
          legend.title = element_text(size = 25),
          legend.text = element_text(size = 25),
          plot.margin = margin(1, 1, 1, 1), 
          line = element_line((size = 4), colour = 'black'),
          legend.spacing.x = unit(1, 'cm')) +
    scale_fill_manual(values = my_palette,
                      name = 'sgRNA type',
                      labels = c('perfect' = 'On-target',
                                'multi-target guides' = 'Multi-target',
                                'single mismatch' = 'Single mismatch',
                                'pam-distal single mismatch' = 'PAM distal single mismatch',
                                'pam-distal double mismatch' = 'PAM distal double mismatch')) +
  scale_x_discrete(labels = c(paste0(length(unique(not_essential_but_hits_brunello$gene))," gene hits \n(A375 not essential)"), 
                              paste0(length(unique(essential_and_hits_brunello$gene))," gene hits \n(A375 essential)")))

multi_bar_chart

ggsave(
    path = './',
    width = 15,
    height = 8,
    dpi = 1000,
    plot = multi_bar_chart,
    filename = "../../figure/04_fig_analysis_gene_hits_change/brunello_multi_barchart_percentage_sgrna.png")
  
```

## 3rd Visualization: Scatterplot of Log2FC and Diff Beta score

```{r}
# Scatterplot of differential beta score and log2fc for A375 not essential gene hits but only represented in original Brunello library
combined_brunello %>%
  filter(is_depmap_essential == "A375_not_essential_hits_only_in_brunello") %>%
  ggplot(aes(mean, Diff, color = alignment)) +
    geom_point(size = 1.5, shape = 19)

```












# Comparison of Brunello only gene hits Log2FC with the Log2FC of avana library A375 cells
## 1st visualization: sgRNA Log2FC of gene hits identified in Brunello projected to Avana library

```{r}

# obtain alignment data

selected_guides <- c("perfect", "single mismatch", "pam-distal double mismatch", "multi-target guides", "non-targeting")
avana_classification <- pick_alignment_type("../../data/sgrna_lfc_data/avana_data/avana_sgrna_alignment_classification.csv", selected_guides)
avana_classification <- count_alignment_bin(avana_classification, selected_guides)
avana_classification[[1]] <- avana_classification[[1]] %>% select(-gene)

# obtain Log2FC data

# 11 genes from brunello only hits are missing in the avana A375 cells
avana_A375 <- read_csv("../../data/sgrna_lfc_data/output_normalized/avana_A375_normalized_lfc.csv") 

obtain_depmap_essential_brunello <- combined_brunello %>%
  select(gene, is_depmap_essential) %>%
  distinct()

avana_A375 <- avana_A375 %>%
  left_join(obtain_depmap_essential_brunello, join_by("gene")) %>%
  # we shouldnt do this but you can check back again later
  drop_na() %>%
  left_join(avana_classification[[1]], join_by(sgRNA, spacer)) %>%
  arrange(sgRNA)

avana_non_essential_genes_control <- avana_A375 %>% filter(is_depmap_essential == "A375_not_essential_control") %>% distinct() %>% pull(gene) %>% unique()
avana_not_essential_but_hits_brunello <- avana_A375 %>% filter(is_depmap_essential == "A375_not_essential_hits_only_in_brunello") %>% distinct() %>% pull(gene) %>% unique()
avana_essential_and_hits_brunello <- avana_A375 %>% filter(is_depmap_essential == "A375_essential_hits_only_in_brunello") %>% distinct() %>% pull(gene) %>% unique()


boxplot_lfc_avana <- avana_A375 %>%
  ggboxplot(x = "is_depmap_essential",
            y = "mean",
            outliers = FALSE) +
    stat_compare_means(comparisons = list(c("A375_not_essential_control", "A375_not_essential_hits_only_in_brunello"),
                                          c("A375_not_essential_control", "A375_essential_hits_only_in_brunello")),
                       method = "wilcox.test",
                       method.args = list(alternative = "two.sided"),
                       size = 2.5,
                       step.increase = 0.05,
                       tip.length = 0.01,
                       label = "p.signif",
                       label.y = 1) +
  labs(title = "", 
         x = "Gene hits", 
         y = "Median sgRNA Log2FC") +
    theme(plot.margin = margin(10, 10, 25, 10),
          strip.text.x = element_text(size = 14),
          axis.text.y = element_text(size = 14, colour = "black"),
          axis.ticks.y = element_line(size = 1.5),
          axis.ticks.x = element_blank(), 
          axis.ticks.length = unit(0.1, "cm"),
          axis.title.x = element_blank(),
          axis.title.y = element_text(size = 14, vjust = 0.5, margin = margin(r = 20)),
          panel.grid.major = element_line(size = 0.5),
          panel.grid.minor = element_blank(),
          axis.line.x = element_line(size = 0.5),
          axis.line.y = element_line(size = 0.5), 
          line = element_line((size = 2), colour = 'black'),
          legend.position = "right",
          legend.text = element_text(size = 12)) +
          scale_x_discrete(labels = c(paste0(length(unique(avana_non_essential_genes_control))," genes\n(A375 not essential)"), 
                                      paste0(length(unique(avana_not_essential_but_hits_brunello))," gene hits \n(A375 not essential)"), 
                                      paste0(length(unique(avana_essential_and_hits_brunello))," gene hits \n(A375 essential)")))
# get stats
  b <- ggplot_build(boxplot_lfc_avana)
  stats <- b$data[[1]]
  label_df <- stats %>%
    transmute(
      x = x,
      y = ymax + 0.2,          # add padding
      label = round(ymax, 2)   # or any custom label
    )
  
  count <- avana_A375 %>%
    select(is_depmap_essential) %>%
    count(is_depmap_essential)
  
  count <- count %>% bind_cols(stats %>% select(x, ymin))
  
  boxplot_lfc_avana <-  boxplot_lfc_avana +
                                geom_text(
                                  data = count,
                                  aes(x = x, y = ymin - 0.5, label = paste0("n = ", n)),
                                  angle = 45,
                                  vjust = 0,
                                  hjust = 0.5,
                                  size = 5
                                )
  
boxplot_lfc_avana

ggsave(
    path = './',
    width = 6,
    height = 5,
    dpi = 1000,
    plot = boxplot_lfc_avana,
    filename = "../../figure/04_fig_analysis_gene_hits_change/avana_brunello_not_essential_hits_lfc.png")

```

## 2nd Visualization: Multi bar chart of percentage of sgRNA type between essential gene hits and non essential but hits in Avana library

```{r}

percentage_data <- avana_A375 %>%
  filter(is_depmap_essential %in% c("A375_not_essential_hits_only_in_brunello", "A375_essential_hits_only_in_brunello")) %>%
  group_by(is_depmap_essential) %>%
  count(alignment) %>%
  dplyr::rename("num_of_guides" = "n") %>%
  mutate(percentage = round(num_of_guides/sum(num_of_guides) * 100, 2)) %>%
  ungroup() %>%
  mutate(alignment = factor(alignment, levels = c("perfect", "multi-target guides", "single mismatch", "pam-distal double mismatch")))

# Define custom colors for the libraries
my_palette <- brewer.pal(n = 5, name = "Set2")

# Create the grouped multibar chart for each sgRNA libraries
multi_bar_chart <- ggplot(percentage_data, aes(x = is_depmap_essential, y = percentage, fill = alignment)) +
    geom_bar(stat = "identity", position = position_dodge2(padding = 0.2), width = 0.7) +
    theme_minimal() +
    labs(title = "", x = "", y = "sgRNA (%)") +
    scale_x_discrete(labels = c('A375_not_essential_hits_only_in_brunello' = 'A375 not essential \n gene hits',
                                  'A375_essential_hits_only_in_brunello' = 'A375 essential \n gene hits')) +
    geom_text(aes(label = percentage), position = position_dodge2(width = 0.7), vjust = -0.5, size = 6.5) +
    theme(axis.text.x = element_text(size = 25, vjust = 0.7, colour = "black"),
          axis.text.y = element_text(size = 25, colour = "black"),
          axis.title.y = element_text(size = 25, vjust = 0.5, margin = margin(r = 20)),
          panel.grid.major = element_blank(),
          panel.grid.minor = element_blank(),
          axis.line.x = element_line(size = 0.5),
          axis.line.y = element_line(size = 0.5),
          axis.ticks.y = element_line(size = 1.5),
          axis.ticks.length = unit(0.3, "cm"),
          legend.title = element_text(size = 25),
          legend.text = element_text(size = 25),
          plot.margin = margin(1, 1, 1, 1), 
          line = element_line((size = 4), colour = 'black'),
          legend.spacing.x = unit(1, 'cm')) +
    scale_fill_manual(values = my_palette,
                      name = 'sgRNA type',
                      labels = c('perfect' = 'On-target',
                                'multi-target guides' = 'Multi-target',
                                'single mismatch' = 'Single mismatch',
                                'pam-distal single mismatch' = 'PAM distal single mismatch',
                                'pam-distal double mismatch' = 'PAM distal double mismatch')) +
  scale_x_discrete(labels = c(paste0(length(unique(avana_not_essential_but_hits_brunello))," gene hits \n(A375 not essential)"), 
                              paste0(length(unique(avana_essential_and_hits_brunello))," gene hits \n(A375 essential)")))

multi_bar_chart

ggsave(
    path = './',
    width = 15,
    height = 8,
    dpi = 1000,
    plot = multi_bar_chart,
    filename = "../../figure/04_fig_analysis_gene_hits_change/avana_multi_barchart_percentage_sgrna.png")
  
```




















### Jacquere analysis

```{r}
jacquere_classification <- pick_alignment_type("../../data/sgrna_lfc_data/jacquere_data/jacquere_sgrna_alignment_classification.csv", selected_guides)
jacquere_classification <- count_alignment_bin(jacquere_classification, selected_guides)

depmap_common_essential <- read_csv("../../data/CRISPRInferredCommonEssentials.csv") %>%
  mutate(Essentials = str_remove(Essentials, "\\s*\\(.*\\)")) %>%
  pull(Essentials)

jacquere_essentiality_profiles <- jacquere_original_hits %>%
  select(Gene, Diff, hits_classification) %>%
  dplyr::rename("gene" = "Gene")

jacquere_original_only_hits_analysis <- jacquere_classification[[1]] %>%
  filter(gene %in% jacquere_original_only_hits) %>%
  left_join(jacquere_essentiality_profiles, join_by(gene)) %>%
  mutate(is_depmap_essential = ifelse(gene %in% depmap_common_essential, TRUE, FALSE))

not_essential_but_hits_jacquere <- jacquere_original_only_hits_analysis %>%
  select(-spacer) %>%
  filter(is_depmap_essential == FALSE)

survey <- jacquere_original_only_hits_analysis %>%
  select(gene, alignment, alignment_bin, Diff) %>%
  group_by(gene) %>%
  count(alignment, alignment_bin) %>%
  dplyr::rename("num_guides" = "n")  %>%
  mutate(is_depmap_essential = ifelse(gene %in% depmap_common_essential, TRUE, FALSE)) %>%
  ungroup(gene)

survey %>%
  select(gene, is_depmap_essential) %>%
  distinct() %>%
  count(is_depmap_essential) %>%
  dplyr::rename("num_genes" = "n") %>%
  ggplot(aes(is_depmap_essential, num_genes)) +
     geom_col(fill = "#1E88E5") +
     theme_minimal() +
     labs(x = "DepMap common essential genes", y = "Number of gene hits") +
     geom_text(aes(label = paste("n= ", num_genes, sep="")),
                nudge_y = 10,
                nudge_x = 0,
                size = 8) +
  theme(axis.text.x = element_text(size = 25, vjust = 0.7, colour = 'black'),
          axis.text.y = element_text(size = 25, colour = "black"),
          axis.title.x = element_text(size = 25, vjust = 0.5, margin = margin(t = 20), colour = "black"),
          axis.title.y = element_text(size = 25, vjust = 0.5, margin = margin(r = 20), colour = "black"),
          axis.ticks.y = element_line(size = 1.5),
          axis.ticks.length = unit(0.3, "cm"),
          axis.line.y.right = element_blank(),
          axis.ticks.y.right = element_blank(), 
          axis.text.y.right = element_blank(), 
          panel.grid.major = element_blank(),
          panel.grid.minor = element_blank(),
          axis.line.x = element_line(size = 0.5, colour = "black"),
          axis.line.y = element_line(size = 0.5, colour = "black"),
          legend.position = "none",
          plot.margin = margin(1, 1, 1, 1),
          legend.spacing.x = unit(1, 'cm'))
```









