---
title: "Basic stats sgRNA library"
author: "Author: Stefanus Bernard"
date: "Date: `r format(Sys.time(), '%d %B %Y')`"
output: 
    html_document:
      theme: cerulean
      toc: true
      toc_float: true
      number_sections: true
      self_contained: true
---

```{r}
library(tidyverse)
library(readxl)
library(grid)
library(paletteer)
library(RColorBrewer)
library(UpSetR)
library(gprofiler2)
source('~/CRISPR-KO-GuideRefine/GuideRefine_functions.R')
source("00_CRISPR-KO-library-survey-functions.R")
```

# Import the data of sgRNA library report

```{r}

terms <- c("CONTROL", "Control", "control", "INTRON", "Intron", "intron", "LacZ", "luciferase", "NO_SITE", "ONE_INTERGENIC_SITE")
# lacZ and luciferase are the control in Toronto V3 library

terms_for_filtering <- paste(terms, collapse = "|")

brunello <- 'broadgpp-brunello-library-contents'
toronto <- 'tkov3_guide_sequence'
avana <- 'avana_library'
yusa <- 'yusa_hcrispr_ko_grnas'
jacquere <- 'Jacquere_PerGuideAnnotations_Quota4'

import_report_without_control <- function(file_dir, control_terms) {
  
  data <- read_excel(paste(file_dir))
  data <- data %>% filter(!str_detect(gene, terms_for_filtering))
  return(data)
  
}

brunello_report <- import_report_without_control(paste0('~/CRISPR-KO-GuideRefine/output_cleaning/T2T-CHM13/', brunello, '_full_report.xlsx'))
toronto_report <- import_report_without_control(paste0('~/CRISPR-KO-GuideRefine/output_cleaning/T2T-CHM13/', toronto, '_full_report.xlsx'))
avana_report <- import_report_without_control(paste0('~/CRISPR-KO-GuideRefine/output_cleaning/T2T-CHM13/', avana, '_full_report.xlsx'))
yusa_report <- import_report_without_control(paste0('~/CRISPR-KO-GuideRefine/output_cleaning/T2T-CHM13/', yusa, '_full_report.xlsx'))
jacquere_report <- import_report_without_control(paste0('~/CRISPR-KO-GuideRefine/output_cleaning/T2T-CHM13/', jacquere, '_full_report.xlsx'))

```

## Import paralog data from Barbara's paper

```{r}
# import predicted paralog pairs from Barbara's paper

# barbara_paralog <- barbara_paralog_gene_pairs()
ensembl_paralog <- ensembl_paralog_gene_pairs()

paralog_pairs <- ensembl_paralog[[1]]
paralog_individual <- ensembl_paralog[[2]]

```

# Calculate the percentage of multi-target, single-mismatch, and PAM-distal double mismatch sgRNA in each libraries

```{r}

brunello_multi_target <- round(calculate_percentage(brunello_report, "multi-target sgRNA", "sgRNA number"), 2)
toronto_multi_target <- round(calculate_percentage(toronto_report, "multi-target sgRNA", "sgRNA number"), 2)
avana_multi_target <- round(calculate_percentage(avana_report, "multi-target sgRNA", "sgRNA number"), 2)
yusa_multi_target <- round(calculate_percentage(yusa_report, "multi-target sgRNA", "sgRNA number"), 2)
jacquere_multi_target <- round(calculate_percentage(jacquere_report, "multi-target sgRNA", "sgRNA number"), 2)

brunello_pam_distal_single <- round(calculate_percentage(brunello_report, "PAM distal single mismatch", "sgRNA number"), 2)
toronto_pam_distal_single <- round(calculate_percentage(toronto_report, "PAM distal single mismatch", "sgRNA number"), 2)
avana_pam_distal_single <- round(calculate_percentage(avana_report, "PAM distal single mismatch", "sgRNA number"), 2)
yusa_pam_distal_single <- round(calculate_percentage(yusa_report, "PAM distal single mismatch", "sgRNA number"), 2)
jacquere_pam_distal_single <- round(calculate_percentage(jacquere_report, "PAM distal single mismatch", "sgRNA number"), 2)

brunello_pam_distal_double <- round(calculate_percentage(brunello_report, "PAM distal double mismatch", "sgRNA number"), 2)
toronto_pam_distal_double <- round(calculate_percentage(toronto_report, "PAM distal double mismatch", "sgRNA number"), 2)
avana_pam_distal_double <- round(calculate_percentage(avana_report, "PAM distal double mismatch", "sgRNA number"), 2)
yusa_pam_distal_double <- round(calculate_percentage(yusa_report, "PAM distal double mismatch", "sgRNA number"), 2)
jacquere_pam_distal_double <- round(calculate_percentage(jacquere_report, "PAM distal double mismatch", "sgRNA number"), 2)

# Create dataframe for multi-target, single-mismatch, and pam-distal from all library

multi_target <- c(brunello_multi_target, toronto_multi_target, yusa_multi_target, avana_multi_target, jacquere_multi_target)
pam_distal_single <- c(brunello_pam_distal_single, toronto_pam_distal_single, yusa_pam_distal_single, avana_pam_distal_single, jacquere_pam_distal_single)
pam_distal_double <- c(brunello_pam_distal_double, toronto_pam_distal_double, yusa_pam_distal_double, avana_pam_distal_double, jacquere_pam_distal_double)

library <- c("brunello", "toronto_v3", "yusa", "avana", "jacquere")
factor <- c(1,2,3,4,5)

sum_dataframe <- data.frame(library, multi_target, pam_distal_single, pam_distal_double)

# Reorder the library factor to ensure the bars are ordered as Brunello, Toronto, Yusa, Avana
sum_dataframe_long <- sum_dataframe %>%
    pivot_longer(cols = c("multi_target", "pam_distal_single", "pam_distal_double"), 
                 names_to = "category", 
                 values_to = "percentage") %>%
    mutate(category = factor(category, levels = c("multi_target", "pam_distal_single", "pam_distal_double")),
           library = factor(library, levels = c("brunello", "toronto_v3", "yusa", "avana", "jacquere")))


# Define custom colors for the libraries
my_palette <- brewer.pal(n = 5, name = "Set2")


# Create the grouped multibar chart for each sgRNA libraries
multi_bar_chart <- ggplot(sum_dataframe_long, aes(x = category, y = percentage, fill = library)) +
    geom_bar(stat = "identity", position = position_dodge2(padding = 0.2), width = 0.7) +
    theme_minimal() +
    labs(title = "", x = "", y = "sgRNA (%)") +
    scale_x_discrete(labels = c('multi_target' = 'Multi-target sgRNA',
                                'pam_distal_single' = 'PAM distal \n single mismatch sgRNA',
                                'pam_distal_double' = 'PAM distal \n double mismatch sgRNA')) +
    geom_text(aes(label = percentage), position = position_dodge2(width = 0.7), vjust = -0.5, size = 6.5) +
    theme(axis.text.x = element_text(size = 25, vjust = 0.7, colour = "black"),
          axis.text.y = element_text(size = 25, colour = "black"),
          axis.title.y = element_text(size = 25, vjust = 0.5, margin = margin(r = 20)),
          panel.grid.major = element_blank(),
          panel.grid.minor = element_blank(),
          axis.line.x = element_line(size = 0.5),
          axis.line.y = element_line(size = 0.5),
          axis.ticks.y = element_line(size = 1.5),
          axis.ticks.length = unit(0.3, "cm"),
          legend.title = element_text(size = 25),
          legend.text = element_text(size = 25),
          plot.margin = margin(1, 1, 1, 1), 
          line = element_line((size = 4), colour = 'black'),
          legend.spacing.x = unit(1, 'cm')) +
    scale_fill_manual(values = my_palette,
                      name = 'Library',
                      breaks = c("brunello", "toronto_v3", "yusa", "avana", "jacquere"),
                      labels = c("brunello" = "Brunello",
                                 "toronto_v3" = "TKOv3",
                                 "yusa" = "Yusa (Project Score)",
                                 "avana" = "Avana",
                                 "jacquere" = "Jacquere")) + 
  scale_y_continuous(breaks=seq(0, 10, 2), limits=c(0, 10))

multi_bar_chart

ggsave(
    path = './',
    width = 20,
    height = 10,
    dpi = 1000,
    plot = multi_bar_chart,
    filename = paste('../figure/01_fig_analysis_off-target-sgrna/percentage_off_target_T2T.png', sep = ''))
```

# Percentage of sgRNAs not represented after library refinement
# Percentage of genes not represented after library refinement

```{r}

avana_result_gene <- percentages_and_list_genes(avana_report, paralog_individual)
brunello_result_gene <- percentages_and_list_genes(brunello_report, paralog_individual)
toronto_result_gene <- percentages_and_list_genes(toronto_report, paralog_individual)
yusa_result_gene <- percentages_and_list_genes(yusa_report, paralog_individual)
jacquere_result_gene <- percentages_and_list_genes(jacquere_report, paralog_individual)

# Export the removed genes from each library (input for script 04_analysis_removed_genes_biogrid.Rmd)

max_len <- max(length(avana_result_gene[["failed_genes"]]), 
               length(avana_result_gene[["failed_paralog"]]), 
               length(avana_result_gene[["failed_singleton"]]),
               length(brunello_result_gene[["failed_genes"]]), 
               length(brunello_result_gene[["failed_paralog"]]), 
               length(brunello_result_gene[["failed_singleton"]]),
               length(toronto_result_gene[["failed_genes"]]), 
               length(toronto_result_gene[["failed_paralog"]]), 
               length(toronto_result_gene[["failed_singleton"]]),
               length(yusa_result_gene[["failed_genes"]]), 
               length(yusa_result_gene[["failed_paralog"]]), 
               length(yusa_result_gene[["failed_singleton"]]),
               length(jacquere_result_gene[["failed_genes"]]), 
               length(jacquere_result_gene[["failed_paralog"]]), 
               length(jacquere_result_gene[["failed_singleton"]])
               )

pad <- function(x, n) {
  length(x) <- n 
  return(x)
}

df_removed_genes <- data.frame(
  avana_removed_genes_all = pad(avana_result_gene[["failed_genes"]], max_len),
  avana_removed_paralog = pad(avana_result_gene[["failed_paralog"]], max_len),
  avana_removed_singleton = pad(avana_result_gene[["failed_singleton"]], max_len),
  brunello_removed_genes_all = pad(brunello_result_gene[["failed_genes"]], max_len),
  brunello_removed_paralog = pad(brunello_result_gene[["failed_paralog"]], max_len),
  brunello_removed_singleton = pad(brunello_result_gene[["failed_singleton"]], max_len),
  toronto_removed_genes_all = pad(toronto_result_gene[["failed_genes"]], max_len),
  toronto_removed_paralog = pad(toronto_result_gene[["failed_paralog"]], max_len),
  toronto_removed_singleton = pad(toronto_result_gene[["failed_singleton"]], max_len),
  yusa_removed_genes_all = pad(yusa_result_gene[["failed_genes"]], max_len),
  yusa_removed_paralog = pad(yusa_result_gene[["failed_paralog"]], max_len),
  yusa_removed_singleton = pad(yusa_result_gene[["failed_singleton"]], max_len),
  jacquere_removed_genes_all = pad(jacquere_result_gene[["failed_genes"]], max_len),
  jacquere_removed_paralog = pad(jacquere_result_gene[["failed_paralog"]], max_len),
  jacquere_removed_singleton = pad(jacquere_result_gene[["failed_singleton"]], max_len),
  stringsAsFactors = FALSE
)

write_csv(df_removed_genes, "../data/biogrid_crispr_ko_data/important_data/removed_genes_all_library.csv")

```

# Data to plot genes (paralog) that are not well represented in the refined library

```{r}

overall_failed_paralog <- unique(c(avana_result_gene[["failed_paralog"]], 
                                   brunello_result_gene[["failed_paralog"]], 
                                   toronto_result_gene[["failed_paralog"]], 
                                   yusa_result_gene[["failed_paralog"]],
                                   jacquere_result_gene[["failed_paralog"]]))
overall_failed_singleton <- unique(c(avana_result_gene[["failed_singleton"]], 
                                   brunello_result_gene[["failed_singleton"]], 
                                   toronto_result_gene[["failed_singleton"]], 
                                   yusa_result_gene[["failed_singleton"]],
                                   jacquere_result_gene[["failed_singleton"]]))

main_df_percentage <- data.frame(library = c("brunello", "toronto_v3", "yusa", "avana", "jacquere"),
                                 paralog = c(brunello_result_gene[[1]], toronto_result_gene[[1]], yusa_result_gene[[1]], avana_result_gene[[1]], jacquere_result_gene[[1]]),
                                 singleton = c(brunello_result_gene[[2]], toronto_result_gene[[2]], yusa_result_gene[[2]], avana_result_gene[[2]], jacquere_result_gene[[2]]),
                                 total_percentage_failed_genes = c(brunello_result_gene[[3]], toronto_result_gene[[3]], yusa_result_gene[[3]], avana_result_gene[[3]], jacquere_result_gene[[3]]),
                                 total_percentage_failed_sgrna = c(brunello_result_gene[[4]], toronto_result_gene[[4]], yusa_result_gene[[4]], avana_result_gene[[4]], jacquere_result_gene[[4]])
                            )

failed_paralog_percentage <- main_df_percentage %>%
  select(!c(total_percentage_failed_genes, total_percentage_failed_sgrna)) %>%
  add_row(library = "overall", paralog = ensembl_paralog[[3]]$percentage[ensembl_paralog[[3]]$has_paralogs == TRUE], singleton = ensembl_paralog[[3]]$percentage[ensembl_paralog[[3]]$has_paralogs == FALSE]) %>%
  mutate(library = factor(library, levels = c("brunello", "toronto_v3", "yusa", "avana", "jacquere", " ","overall"))) %>%
  pivot_longer(!library, names_to = "not_represented_genes", values_to = "percentage") %>% 
  mutate(not_represented_genes = factor(not_represented_genes, levels = c("singleton", "paralog"), labels = c("Singleton", "Paralog")))

```

# Plot of genes (paralog) that are not well represented in the refined library

```{r}

removed_paralog <- ggplot(failed_paralog_percentage, aes(x = library, y = percentage, fill = not_represented_genes)) +
      geom_bar(stat = "identity", position = "stack") +
      scale_fill_discrete(
          name = "",
          labels = c("Singleton", "Paralog")) +
        theme_minimal() +
        labs(title = "", x = "", y = "Genes targeted by < 3 sgRNAs (%)") +
        scale_x_discrete(labels = c("brunello" = "Brunello",
                                     "toronto_v3" = "TKOv3",
                                     "yusa" = "Yusa \n(Project Score)",
                                     "avana" = "Avana",
                                     "jacquere" = "Jacquere",
                                     "overall" = "Overall")) +
        geom_text(aes(label = percentage), position = position_stack(vjust = 0.5), vjust = 0, size = 8) +
        geom_hline(yintercept = ensembl_paralog[[3]]$percentage[ensembl_paralog[[3]]$has_paralogs == TRUE], linetype = "solid", color = "black", size = 0.5) +
        theme(axis.text.x = element_text(size = 25, vjust = 0.7, colour = "black"),
              axis.text.y = element_text(size = 25, colour = "black"),
              axis.title.y = element_text(size = 25, vjust = 0.5, margin = margin(r = 20)),
              panel.grid.major = element_blank(),
              panel.grid.minor = element_blank(),
              axis.line.x = element_line(size = 0.5),
              axis.line.y = element_line(size = 0.5),
              axis.ticks.y = element_line(size = 1.5),
              axis.ticks.length = unit(0.3, "cm"),
              legend.title = element_text(size = 25),
              legend.text = element_text(size = 25),
              plot.margin = margin(1, 1, 1, 1), 
              line = element_line((size = 4), colour = "black"),
              legend.spacing.x = unit(1, 'cm')) +
      ylim(0, 101)

removed_paralog

ggsave(
    path = './',
    width = 17,
    height = 10,
    dpi = 300,
    plot = removed_paralog,
    filename = paste('../figure/01_fig_analysis_off-target-sgrna/library_genes_removed_paralog_T2T.png', sep = ''))


```

# Plot of genes that are not represented in the refined library

```{r}

my_palette <- brewer.pal(n = 5, name = "Set2")

failed_genes_percentage <- main_df_percentage %>%
  select(-c(paralog, singleton)) %>%
  mutate(library = factor(library, levels = c("brunello", "toronto_v3", "yusa", "avana", "jacquere")))

percentage_bar_chart <- ggplot(failed_genes_percentage, aes(x = library, y = total_percentage_failed_genes, fill = library)) +
    geom_bar(stat = "identity", position = "stack") +
        theme_minimal() +
        labs(title = "", x = "", y = "Genes targeted by < 3 sgRNAs (%)") +
        scale_x_discrete(labels = c("brunello" = "Brunello",
                                     "toronto_v3" = "TKOv3",
                                     "yusa" = "Yusa \n(Project Score)",
                                     "avana" = "Avana",
                                     "jacquere" = "Jacquere")) +
        geom_text(aes(label = total_percentage_failed_genes), position = position_stack(), vjust = -0.5, size = 7) +
        theme(axis.text.x = element_text(size = 25, vjust = 0.7, colour = 'black'),
              axis.text.y = element_text(size = 25, colour = 'black'),
              axis.title.y = element_text(size = 25, vjust = 0.5, margin = margin(r = 20)),
              panel.grid.major = element_blank(),
              panel.grid.minor = element_blank(),
              axis.line.x = element_line(size = 0.5),
              axis.line.y = element_line(size = 0.5),
              axis.ticks.y = element_line(size = 1.5),
              axis.ticks.length = unit(0.3, "cm"),
              legend.title = element_text(size = 25),
              legend.text = element_text(size = 25),
              plot.margin = margin(1, 1, 1, 1), 
              line = element_line((size = 4), colour = 'black'),
              legend.spacing.x = unit(1, 'cm')) +
    scale_fill_manual(values = my_palette,
                      name = 'Library',
                      breaks = c("brunello", "toronto_v3", "yusa", "avana", "jacquere"),
                      labels = c("brunello" = "Brunello",
                                 "toronto_v3" = "TKOv3",
                                 "yusa" = "Yusa (Project Score)",
                                 "avana" = "Avana",
                                 "jacquere" = "Jacquere")) +
      ylim(0, 100)

percentage_bar_chart

ggsave(
    path = './',
    width = 16,
    height = 10,
    dpi = 1000,
    plot = percentage_bar_chart,
    filename = paste('../figure/01_fig_analysis_off-target-sgrna/library_genes_bar_chart_T2T.png', sep = ''))

```

# plot of sgRNA that is not represented in the refined library

```{r}

my_palette <- brewer.pal(n = 5, name = "Set2")

failed_sgrna_percentage <- main_df_percentage %>%
  select(-c(paralog, singleton, total_percentage_failed_genes)) %>%
  mutate(library = factor(library, levels = c("brunello", "toronto_v3", "yusa", "avana", "jacquere")))

percentage_bar_chart <- ggplot(failed_sgrna_percentage, aes(x = library, y = total_percentage_failed_sgrna, fill = library)) +
    geom_bar(stat = "identity", position = "stack") +
        theme_minimal() +
        labs(title = "", x = "", y = "Removed sgRNAs (%)") +
        scale_x_discrete(labels = c("brunello" = "Brunello",
                                     "toronto_v3" = "TKOv3",
                                     "yusa" = "Yusa \n(Project Score)",
                                     "avana" = "Avana",
                                     "jacquere" = "Jacquere")) +
        geom_text(aes(label = total_percentage_failed_sgrna), position = position_stack(), vjust = -0.5, size = 7) +
        theme(axis.text.x = element_text(size = 25, vjust = 0.7, colour = 'black'),
              axis.text.y = element_text(size = 25, colour = 'black'),
              axis.title.y = element_text(size = 25, vjust = 0.5, margin = margin(r = 20)),
              panel.grid.major = element_blank(),
              panel.grid.minor = element_blank(),
              axis.line.x = element_line(size = 0.5),
              axis.line.y = element_line(size = 0.5),
              axis.ticks.y = element_line(size = 1.5),
              axis.ticks.length = unit(0.3, "cm"),
              legend.title = element_text(size = 25),
              legend.text = element_text(size = 25),
              plot.margin = margin(1, 1, 1, 1), 
              line = element_line((size = 4), colour = 'black'),
              legend.spacing.x = unit(1, 'cm')) +
    scale_fill_manual(values = my_palette,
                      name = 'Library',
                      breaks = c("brunello", "toronto_v3", "yusa", "avana", "jacquere"),
                      labels = c("brunello" = "Brunello",
                                 "toronto_v3" = "TKOv3",
                                 "yusa" = "Yusa (Project Score)",
                                 "avana" = "Avana",
                                 "jacquere" = "Jacquere")) +
      ylim(0, 100)

percentage_bar_chart

ggsave(
    path = './',
    width = 16,
    height = 10,
    dpi = 1000,
    plot = percentage_bar_chart,
    filename = paste('../figure/01_fig_analysis_off-target-sgrna/library_sgrna_bar_chart_T2T.png', sep = ''))


```

# UpSet plot for commonality of removed genes between 5 libraries

```{r}

removed_gene_list_combined <- list(
  Avana = avana_result_gene[["failed_genes"]],
  Brunello = brunello_result_gene[["failed_genes"]],
  Toronto = toronto_result_gene[["failed_genes"]],
  Yusa = yusa_result_gene[["failed_genes"]],
  Jacquere = jacquere_result_gene[["failed_genes"]]
)

png("../figure/01_fig_analysis_off-target-sgrna/removed_genes_all_libraries.png", width = 12000, height = 7500, res = 1000)

par(mar = c(5, 20, 4, 2)) 
upset(fromList(removed_gene_list_combined), 
      order.by = "degree",
      main.bar.color = "#4B0092",
      sets.bar.color = "#1AFF1A",
      text.scale = 1.5)

dev.off()

```

# Genes removed from all libraries

```{r}

removed_genes_all_libraries <- Reduce(intersect, removed_gene_list_combined)





```








