---
title: "Analysis of commonly removed genes across libraries"
author: "Author: Stefanus Bernard"
date: "Date: `r format(Sys.time(), '%d %B %Y')`"
output: 
  html_document :
    theme : cerulean
    toc : true
    toc_float : true
    number_section : true
---

```{r setup, include=FALSE}
library(tidyverse)
library(stringr)
library(UpSetR)
library(GenomicFeatures)
library(HGNChelper)
library(BSgenome.Hsapiens.NCBI.T2TCHM13v2.0)
library(GenomeInfoDbData)
library(txdbmaker)
library(crisprVerse)
library(pheatmap)

source("./00_CRISPR-KO-library-survey-functions.R")
```

## Import data containing removed genes from each library

```{r}

removed_genes <- read_csv("../data/biogrid_crispr_ko_data/important_data/removed_genes_all_library.csv")

removed_gene_list_combined <- list(
  Avana = na.omit(removed_genes[["avana_removed_genes_all"]]),
  Brunello = na.omit(removed_genes[["brunello_removed_genes_all"]]),
  Toronto = na.omit(removed_genes[["toronto_removed_genes_all"]]),
  Yusa = na.omit(removed_genes[["yusa_removed_genes_all"]]),
  Jacquere = na.omit(removed_genes[["jacquere_removed_genes_all"]])
)

# List of 520 protein-coding genes are removed across CRISPR-KO libraries, what are these genes?
removed_genes_all_libraries <- unique(Reduce(intersect, removed_gene_list_combined))
all_genes_but_not_removed <- read_tsv("https://storage.googleapis.com/public-download-files/hgnc/tsv/tsv/hgnc_complete_set.txt") %>%
  filter(locus_group == "protein-coding gene" & !symbol %in% removed_genes_all_libraries) %>%
  pull(symbol)

```

## UpSet plot for commonality of removed genes between 5 libraries

```{r}

png("../figure/01_fig_analysis_off-target-sgrna/removed_genes_all_libraries.png", width = 12000, height = 7500, res = 1000)

par(mar = c(5, 20, 4, 2))
upset(fromList(removed_gene_list_combined),
      order.by = "degree",
      main.bar.color = "#4B0092",
      sets.bar.color = "#1AFF1A",
      text.scale = 1.5)

dev.off()

```

## There are 520 genes removed across libraries
we asked why there are 520 genes failed/represented by fewer than 3 guides across libraries

## First we survey for each genes, what are the percentage of genes removed by specific multi-target-/off-target guides

```{r}

terms <- c("CONTROL", "Control", "control", "INTRON", "Intron", "intron", "LacZ", "luciferase", "NO_SITE", "ONE_INTERGENIC_SITE")
# lacZ and luciferase are the control in Toronto V3 library

terms_for_filtering <- paste(terms, collapse = "|")

brunello <- 'broadgpp-brunello-library-contents'
toronto <- 'tkov3_guide_sequence'
avana <- 'avana_library'
yusa <- 'yusa_hcrispr_ko_grnas'
jacquere <- 'Jacquere_PerGuideAnnotations_Quota4'

# import sgRNA refining report

avana_report <- import_and_process_report(paste0('../../GuideRefine/output_cleaning/T2T-CHM13/', avana, '_full_report.xlsx')) %>% filter(gene_hgnc %in% removed_genes_all_libraries)
brunello_report <- import_and_process_report(paste0('../../GuideRefine/output_cleaning/T2T-CHM13/', brunello, '_full_report.xlsx')) %>% filter(gene_hgnc %in% removed_genes_all_libraries)
toronto_report <- import_and_process_report(paste0('../../GuideRefine/output_cleaning/T2T-CHM13/', toronto, '_full_report.xlsx')) %>% filter(gene_hgnc %in% removed_genes_all_libraries) %>% dplyr::select(-`PAM distal single mismatch`)
yusa_report <- import_and_process_report(paste0('../../GuideRefine/output_cleaning/T2T-CHM13/', yusa, '_full_report.xlsx')) %>% filter(gene_hgnc %in% removed_genes_all_libraries)
jacquere_report <- import_and_process_report(paste0('../../GuideRefine/output_cleaning/T2T-CHM13/', jacquere, '_full_report.xlsx')) %>% filter(gene_hgnc %in% removed_genes_all_libraries)

combined_report <- rbind(avana_report, brunello_report, toronto_report, yusa_report, jacquere_report) %>%
  arrange(gene_hgnc) %>%
  dplyr::select(-c(`Additional sgRNA (Corrected)`, `counted total sgRNA`, `actual total sgRNA`, `number match`, `symbol change`, `protein-coding gene`, `quality check`))


total_col <- "sgRNA number"
part_cols  <- c("multi-target sgRNA", "single mismatch sgRNA", "sgRNA not target exon")
median_cols <- c("median_rate_multi-target sgRNA", "median_rate_single mismatch sgRNA", "median_rate_sgRNA not target exon")

# normalize sgrna
normalize_sgrna <- combined_report %>%
  dplyr::select(-c(gene, Approved)) %>%
  group_by(gene_hgnc) %>%
  mutate(
    across(
      all_of(part_cols),
      ~ ifelse(`sgRNA number` > 0, .x / `sgRNA number`, NA_real_),
      .names = "rate_{.col}"
    )
  )

# dataset sanity check
normalize_sgrna %>%
  summarise(
    across(
      starts_with("rate_"),
      ~ median(.x, na.rm = TRUE)
    )
  )


# Combine dataset at gene level by median
gene_summary <- normalize_sgrna %>%
  group_by(gene_hgnc) %>%
  summarise(
    across(
      starts_with("rate_"),
      ~ median(.x, na.rm = TRUE),
      .names = "median_{.col}"
    ),
    n_datasets_present = sum(`sgRNA number` > 0, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  # filter only genes where original number of sgRNA intact in the original library
  filter(n_datasets_present > 4)

mat <- as.matrix(gene_summary[, median_cols])
rownames(mat) <- gene_summary$gene_hgnc

png("../figure/01_fig_analysis_off-target-sgrna/heatmap_cluster_removed_genes.png", width = 500, height = 1000, res = 150)
pheatmap(
  mat,
  cluster_rows = TRUE,
  cluster_cols = TRUE,
  angle_col = 45,
  show_rownames = FALSE,
  color = colorRampPalette(c("blue", "white", "red"))(100),
  labels_col = c("Median rate \nmulti-target sgRNA", "Median rate \nsingle mismatch sgRNA", "Median rate \nsgRNA not target exon")
)

dev.off()

```





## Hypothesis 1: maybe those genes are hard to target

First, we prepare the T2T-CHM13 txdb file

```{r}
# create txdb sqlite file for T2T-CHM13

T2T_txdb <- txdbmaker::makeTxDbFromGFF(file = "../../GuideRefine/annotation_file/GCF_009914755.1_T2T-CHM13v2.0_genomic.gff.gz",
                            dataSource = "NCBI T2T Assembly GCF_009914755.1_T2T-CHM13v2.0",
                            organism = "Homo sapiens")
saveDb(T2T_txdb, "../data/T2T_data/TxDb_T2T-CHM13v2.sqlite")

```

## We asked whether the size of genes and limited number of protospacer site contributes to the off-target activity of sgRNAs ?

```{r}
# Load TXdb file

T2T_txdb <- loadDb("../data/T2T_data/TxDb_T2T-CHM13v2.sqlite")
columns(T2T_txdb)

T2T_txdb

refseq_to_chr <- c(
  "NC_060925.1" = "1",
  "NC_060926.1" = "2",
  "NC_060927.1" = "3",
  "NC_060928.1" = "4",
  "NC_060929.1" = "5",
  "NC_060930.1" = "6",
  "NC_060931.1" = "7",
  "NC_060932.1" = "8",
  "NC_060933.1" = "9",
  "NC_060934.1" = "10",
  "NC_060935.1" = "11",
  "NC_060936.1" = "12",
  "NC_060937.1" = "13",
  "NC_060938.1" = "14",
  "NC_060939.1" = "15",
  "NC_060940.1" = "16",
  "NC_060941.1" = "17",
  "NC_060942.1" = "18",
  "NC_060943.1" = "19",
  "NC_060944.1" = "20",
  "NC_060945.1" = "21",
  "NC_060946.1" = "22",
  "NC_060947.1" = "X",
  "NC_060948.1" = "Y"
)



# exons per gene
cds_by_gene <- cdsBy(T2T_txdb, by = "gene")
cds_gr <- cds_by_gene[names(cds_by_gene) %in% all_genes_but_not_removed]

# only filter canonical cds (not predicted XM or else)
cds_gr <- endoapply(cds_gr, function(gr) {
  gr[str_detect(gr$cds_name, "NP")]
})

# rename seqlevels
seqlevels(cds_gr) <- refseq_to_chr[seqlevels(cds_gr)]

# drop anything not in the genome
cds_gr <- keepSeqlevels(
  cds_gr,
  seqlevels(BSgenome.Hsapiens.NCBI.T2TCHM13v2.0),
  pruning.mode = "coarse"
)

```

## obtain possible protospacer sites from CDS data

```{r}
# flatten cds granges (that contain ~500 genes), with identity attached
cds_flat <- unlist(cds_gr, use.names = FALSE)
cds_flat

cds_flat$gene_id <- rep(
  names(cds_gr),
  elementNROWS(cds_gr)
)

# put the flat data as a dataframe to join later
df_cds_flat <- as.data.frame(cds_flat)


# reduced cds within each gene, the function of reduced cds is to filtered out the duplicate from canonical and isoform transcript that has the same start-end exonic region
cds_reduced <- reduce(
  split(cds_flat, cds_flat$gene_id)
)

cds_reduced <- unlist(cds_reduced, use.names = FALSE)

# find all possible spacers site from the cds of ~500 removed genes
data(SpCas9, package="crisprBase")
cds_find_spacer <- findSpacers(cds_reduced, bsgenome = BSgenome.Hsapiens.NCBI.T2TCHM13v2.0, crisprNuclease = SpCas9)

hits <- findOverlaps(cds_find_spacer, cds_flat, type = "within")
hits

df_all_spacers <- data.frame(
  gene_id = cds_flat$gene_id[subjectHits(hits)],
  cds_id = cds_flat$cds_id[subjectHits(hits)],
  spacer_chr = as.character(seqnames(cds_find_spacer)[queryHits(hits)]),
  spacer_strand = as.character(strand(cds_find_spacer)[queryHits(hits)]),
  pam = cds_find_spacer$pam[queryHits(hits)],
  pam_site = cds_find_spacer$pam_site[queryHits(hits)],
  cut_site = cds_find_spacer$cut_site[queryHits(hits)],
  spacer_seq = cds_find_spacer$protospacer[queryHits(hits)],
  stringsAsFactors = FALSE
)

all_spacers_exon_data <- df_all_spacers %>%
  left_join(df_cds_flat, join_by(gene_id, cds_id)) %>%
  dplyr::select(-c(seqnames, strand)) %>%
  dplyr::rename("start_cds" = "start",
                "end_cds" = "end",
                "chromosome" = "spacer_chr",
                "strand" = "spacer_strand") %>%
  relocate(cds_name, .before = start_cds)

write_csv("all_spacers_exon_data_all_genes.csv")

```

## Does each genes have limited number of protospacer site ?

```{r}

count_protospacer_per_gene <- all_spacers_exon_data %>%
  dplyr::select(gene_id, pam_site) %>%
  distinct() %>%
  group_by(gene_id) %>%
  summarise(num_pam_site = n()) %>%
  ungroup()

count_protospacer_per_gene$count_bin <- cut(count_protospacer_per_gene$num_pam_site,
                                            breaks = c(0, 10, 20, 30, 40, 50, 60, 70, 80, 90, 100, Inf),
                                            labels = c(
                                              "0–10",
                                              "10–20",
                                              "20–30",
                                              "30–40",
                                              "40–50",
                                              "50-60",
                                              "60–70",
                                              "70–80",
                                              "80–90",
                                              "90-100",
                                              ">100"),
                                            right = TRUE,
                                            include.lowest = TRUE
                                          )

df_barplot <- count_protospacer_per_gene %>%
  dplyr::select(count_bin) %>%
  count(count_bin) %>%
  dplyr::rename("num_genes" = "n")


ggplot(df_barplot, aes(x = count_bin, y = num_genes)) +
    geom_col(fill = "#1E88E5") +
    theme_minimal() +
    labs(x = "Number of NGG PAM-site at CDS", y = "Number of genes") +
    geom_text(aes(label = paste("n=", num_genes, sep="")),
              nudge_y = 5,
              nudge_x = 0,
              size = 5) +
    theme(axis.text.x = element_text(size = 12, vjust = 0.7, angle = 45, colour = 'black'),
          axis.text.y = element_text(size = 12, colour = "black"),
          axis.title.x = element_text(size = 12, vjust = 0.5, margin = margin(r = 20), colour = "black"),
          axis.title.y = element_text(size = 12, vjust = 0.5, margin = margin(r = 20), colour = "black"),
          axis.ticks.y = element_line(size = 1),
          axis.ticks.length = unit(0.15, "cm"),
          axis.line.y.right = element_blank(),
          axis.ticks.y.right = element_blank(), 
          axis.text.y.right = element_blank(), 
          panel.grid.major = element_blank(),
          panel.grid.minor = element_blank(),
          axis.line.x = element_line(size = 0.5, colour = "black"),
          axis.line.y = element_line(size = 0.5, colour = "black"),
          legend.position = "none",
          plot.margin = margin(1, 1, 1, 1),
          legend.spacing.x = unit(1, 'cm'))


```

```{r}
# Another code to get the summary of exon length from all removed genes

# Extract gene coordinates
genes_gr <- genes(T2T_txdb)

# Filter for your list
genes_gr <- genes_gr[names(genes_gr) %in% removed_genes_all_libraries]
genes_gr$span_length <- width(genes_gr)

exon_lengths <- lapply(exons_gr, width)
exon_lengths_df <- stack(exons_gr)
exon_lengths_df$exon_length <- width(exon_lengths_df)

data_exons_all_removed_genes <- as.data.frame(exon_lengths_df) %>%
  filter(str_detect(exon_name, "NM")) %>%
  mutate(exon_nm_accession = str_extract(exon_name, "^[^.]+"))

# dataframe of total exon length in basepairs (of only 500 genes)
total_exon_length <- data_exons_all_removed_genes %>%
  dplyr::select(name, exon_nm_accession, exon_length) %>%
  arrange(name) %>%
  group_by(name, exon_nm_accession) %>%
  summarise(total_exon_length = sum(exon_length))

```











