---
title: "Analysis of commonly removed genes across libraries"
author: "Author: Stefanus Bernard"
date: "Date: `r format(Sys.time(), '%d %B %Y')`"
output: 
  html_document :
    theme : cerulean
    toc : true
    toc_float : true
    number_section : true
---

```{r setup, include=FALSE}
library(tidyverse)
library(UpSetR)
library(GenomicFeatures)
library(HGNChelper)
library(BSgenome.Hsapiens.NCBI.T2TCHM13v2.0)

source("./00_CRISPR-KO-library-survey-functions.R")
```

## Import data containing removed genes from each library

```{r}

removed_genes <- read_csv("../data/biogrid_crispr_ko_data/important_data/removed_genes_all_library.csv")

removed_gene_list_combined <- list(
  Avana = na.omit(removed_genes[["avana_removed_genes_all"]]),
  Brunello = na.omit(removed_genes[["brunello_removed_genes_all"]]),
  Toronto = na.omit(removed_genes[["toronto_removed_genes_all"]]),
  Yusa = na.omit(removed_genes[["yusa_removed_genes_all"]]),
  Jacquere = na.omit(removed_genes[["jacquere_removed_genes_all"]])
)

# List of 520 protein-coding genes are removed across CRISPR-KO libraries, what are these genes?
removed_genes_all_libraries <- unique(Reduce(intersect, removed_gene_list_combined))

```

## UpSet plot for commonality of removed genes between 5 libraries

```{r}

png("../figure/01_fig_analysis_off-target-sgrna/removed_genes_all_libraries.png", width = 12000, height = 7500, res = 1000)

par(mar = c(5, 20, 4, 2))
upset(fromList(removed_gene_list_combined),
      order.by = "degree",
      main.bar.color = "#4B0092",
      sets.bar.color = "#1AFF1A",
      text.scale = 1.5)

dev.off()

```

## Import more relevant data

```{r}

# Import HGNC dataset

hgnc <- read_tsv(url("https://storage.googleapis.com/public-download-files/hgnc/tsv/tsv/hgnc_complete_set.txt")) %>%
  filter(symbol %in% removed_genes_all_libraries) %>%
  dplyr::select(symbol, name, locus_group, ensembl_gene_id) %>%
  dplyr::rename("gene" = "symbol")

# Import paralog data
ensembl_paralog <- ensembl_paralog_gene_pairs()

paralog_pairs <- ensembl_paralog[[1]]
paralog_individual <- ensembl_paralog[[2]]

# Import the data of sgRNA library report (looking on how many guides actually multi-targeting, etc)

terms <- c("CONTROL", "Control", "control", "INTRON", "Intron", "intron", "LacZ", "luciferase", "NO_SITE", "ONE_INTERGENIC_SITE")
# lacZ and luciferase are the control in Toronto V3 library

terms_for_filtering <- paste(terms, collapse = "|")

avana <- 'avana_library'
brunello <- 'broadgpp-brunello-library-contents'
toronto <- 'tkov3_guide_sequence'
yusa <- 'yusa_hcrispr_ko_grnas'
jacquere <- 'Jacquere_PerGuideAnnotations_Quota4'


avana_report <- import_and_process_report(paste0('~/CRISPR-KO-GuideRefine/output_cleaning/T2T-CHM13/', avana, '_full_report.xlsx')) %>% filter(gene_hgnc %in% removed_genes_all_libraries) %>% mutate(library = "avana")

brunello_report <- import_and_process_report(paste0('~/CRISPR-KO-GuideRefine/output_cleaning/T2T-CHM13/', brunello, '_full_report.xlsx')) %>% filter(gene_hgnc %in% removed_genes_all_libraries) %>% mutate(library = "brunello")

toronto_report <- import_and_process_report(paste0('~/CRISPR-KO-GuideRefine/output_cleaning/T2T-CHM13/', toronto, '_full_report.xlsx')) %>% filter(gene_hgnc %in% removed_genes_all_libraries) %>% mutate(library = "toronto")

yusa_report <- import_and_process_report(paste0('~/CRISPR-KO-GuideRefine/output_cleaning/T2T-CHM13/', yusa, '_full_report.xlsx')) %>% filter(gene_hgnc %in% removed_genes_all_libraries) %>% mutate(library = "yusa")

jacquere_report <- import_and_process_report(paste0('~/CRISPR-KO-GuideRefine/output_cleaning/T2T-CHM13/', jacquere, '_full_report.xlsx')) %>% filter(gene_hgnc %in% removed_genes_all_libraries) %>% mutate(library = "jacquere")

combined_report <- rbind(avana_report, brunello_report, toronto_report, yusa_report, jacquere_report) %>% arrange(gene_hgnc)

rm(avana_report, brunello_report, toronto_report, yusa_report, jacquere_report)

# Import the alignment classification data

avana_alignment <- read_csv("../data/sgrna_lfc_data/avana_data/avana_sgrna_alignment_classification.csv") %>% update_gene_symbol(., removed_genes_all_libraries) %>% mutate(library = "avana")

brunello_alignment <- read_csv("../data/sgrna_lfc_data/brunello_data/brunello_sgrna_alignment_classification.csv") %>% update_gene_symbol(., removed_genes_all_libraries) %>% mutate(library = "brunello")

toronto_alignment <- read_csv("../data/sgrna_lfc_data/tkov3_data/tkov3_sgrna_alignment_classification.csv") %>% update_gene_symbol(., removed_genes_all_libraries) %>% mutate(library = "toronto")

yusa_alignment <- read_csv("../data/sgrna_lfc_data/yusa_data/yusa_sgrna_alignment_classification.csv") %>% update_gene_symbol(., removed_genes_all_libraries) %>% mutate(library = "yusa")

jacquere_alignment <- read_csv("../data/sgrna_lfc_data/jacquere_data/jacquere_sgrna_alignment_classification.csv") %>% update_gene_symbol(., removed_genes_all_libraries) %>% mutate(library = "jacquere")

combined_alignment <- rbind(avana_alignment, brunello_alignment, toronto_alignment, yusa_alignment, jacquere_alignment) %>% 
  arrange(gene_hgnc) %>%
  # remove perfect guides because we dont need it
  filter(!alignment == "perfect")

rm(avana_alignment, brunello_alignment, toronto_alignment, yusa_alignment, jacquere_alignment)

```

## There are 520 genes removed across libraries
we asked why there are 520 genes failed/represented by fewer than 3 guides across libraries

1. Our first hypothesis is maybe the same sgRNA design were used across libraries for some genes. Which led to the failure of on-target activity 

```{r}

duplicated_sequence <- combined_alignment %>%
  # filter(duplicated(spacer) | duplicated(spacer, fromLast = TRUE)) %>%
  dplyr::select(spacer, library) %>%
  group_by(library) %>%
  summarise(seqs = list(unique(spacer))) %>%
  deframe()
  
upset(fromList(duplicated_sequence),
      order.by = "degree",
      main.bar.color = "#4B0092",
      sets.bar.color = "#1AFF1A",
      text.scale = 1.5)

```

However, only 30 sgRNA sequences, that failed and shared among the libraries (except Yusa Project Score). Therefore, sgRNA sequences is not a contributing factor for multi-targeting-/off-targeting sgRNAs.

We asked whether paralogs are the major contributing factor in causing multi-target-/off-targeting activity in sgRNAs. 

```{r}

# count how many 520 genes has paralogs
df <- hgnc %>%
  left_join(ensembl_paralog[[4]], join_by(gene, locus_group, ensembl_gene_id)) %>%
  mutate(has_paralogs = replace_na(has_paralogs, FALSE))
  
# count percentage of paralogs
df_percent_paralogs <- df %>%
  dplyr::select(gene, has_paralogs) %>%
  distinct() %>%
  count(has_paralogs) %>%
  dplyr::rename("count" = "n") %>%
  mutate(percent = count / sum(count) * 100)

# 377 genes out of 520 that were removed across libraries are paralogs (72.5%), indicating significant number of enrichment


# we focused our attention to the 377 paralog genes that were removed across libraries. Furthermore, we discussed the 177 singleton genes separately

removed_paralog <- df %>% 
  dplyr::select(gene, has_paralogs) %>%
  group_by(gene) %>%
  summarise(count_paralog = if_else(has_paralogs == TRUE, n(), 0)) %>%
  arrange(count_paralog) %>%
  ungroup() %>%
  distinct()

removed_paralog$paralogs_bin <- cut(removed_paralog$count_paralog,
                                    breaks = c(-0.5, 0.5:5.5, Inf),  # -0.5 to 5.5 for 0–5, then Inf for >5
                                    labels = c(as.character(0:5), "> 5"),
                                    right = TRUE)

removed_paralog %>% count(paralogs_bin)

# 0	143			
# 1	67			
# 2	38			
# 3	25			
# 4	11			
# 5	11			
# > 5	225	

removed_paralog_bin <- removed_paralog %>% 
  count(paralogs_bin) %>% 
  dplyr::rename("count_paralog" = "n")

boxplot_paralog <- ggplot(removed_paralog_bin, aes(x = paralogs_bin, y = count_paralog)) +
    geom_col(fill = "#1E88E5") +
    theme_minimal() +
  labs(x = "Number of paralogs", y = "Removed genes across 5 libraries") +
  geom_text(aes(label = paste("n=", count_paralog, sep="")),
            nudge_y = 10,
            nudge_x = 0,
            size = 5) +
  theme(axis.text.x = element_text(size = 15, vjust = 0.7, colour = 'black'),
        axis.text.y = element_text(size = 15, colour = "black"),
        axis.title.x = element_text(size = 15, vjust = 0.5, margin = margin(t = 10), colour = "black"),
        axis.title.y = element_text(size = 15, vjust = 0.5, margin = margin(r = 10), colour = "black"),
        axis.ticks.y = element_line(size = 1.5),
        axis.ticks.length = unit(0.15, "cm"),
        axis.line.y.right = element_blank(),
        axis.ticks.y.right = element_blank(), 
        axis.text.y.right = element_blank(), 
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.line.x = element_line(size = 0.5, colour = "black"),
        axis.line.y = element_line(size = 0.5, colour = "black"),
        legend.position = "none",
        plot.margin = margin(1, 1, 1, 1),
        legend.spacing.x = unit(1, 'cm'))

ggsave(
  path = '../figure/01_fig_analysis_off-target-sgrna/',
  width = 6.5,
  height = 5,
  dpi = 1000,
  plot = boxplot_paralog,
  filename = "removed_genes_paralogs.png")


# we found most removed genes are containing paralogs family, with more than >1 paralogs

# since paralogs came from gene duplication event, we asked whether the genes with more paralogs have their sequences more identical with the original genes

removed_paralog_max <- df %>% 
  dplyr::select(gene, has_paralogs, percent_identity_human_gene_identical_to_query) %>%
  group_by(gene) %>%
  # count paralog NA because they dont have paralog gene
  summarise(count_paralog = if_else(has_paralogs == TRUE, n(), NA),
            max_percent_identity = max(percent_identity_human_gene_identical_to_query, na.rm = TRUE)) %>%
  arrange(count_paralog) %>%
  ungroup() %>%
  distinct() %>%
  # drop NA/removed genes without paralogs
  drop_na()

removed_paralog_max$paralogs_bin <- cut(removed_paralog_max$count_paralog,
                                    breaks = c(-0.5, 0.5:5.5, Inf),  # -0.5 to 5.5 for 0–5, then Inf for >5
                                    labels = c(as.character(0:5), "> 5"),
                                    right = TRUE)


annotation_df <- removed_paralog_max %>%
    count(paralogs_bin)

boxplot_percent_identity_paralog <- removed_paralog_max %>%
    ggplot(aes(paralogs_bin, max_percent_identity, group = paralogs_bin)) +
    geom_boxplot(outlier.shape = NA, size = 0.5) +
    theme_minimal() +
    labs(title = "", 
         x = "Number of paralogs for a single gene", 
         y = "% of protein sequences \nmatched in closest paralog") +
    theme(
      axis.text.x = element_text(size = 15, vjust = 0.7, colour = "black"),
      axis.text.y = element_text(size = 15, colour = "black"),
      axis.title.x = element_text(size = 15, margin = margin(t = 10)),
      axis.title.y = element_text(size = 15, margin = margin(r = 10)),
      axis.ticks = element_line(size = 1.5),
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      axis.line.x = element_line(size = 0.5),
      axis.line.y = element_line(size = 0.5))

  b <- ggplot_build(boxplot_percent_identity_paralog)
  boxplot_whisker <- b$data[[1]]$ymax
  
  boxplot_percent_identity_paralog <- boxplot_percent_identity_paralog +
    geom_text(
      data = annotation_df,
      aes(x = paralogs_bin, y = boxplot_whisker + 0.1, label = paste0("n= ", n)),
      size = 3.5,
      angle = 45,
      vjust = -0.2,
      hjust = -0.2
    ) +
    coord_cartesian(clip = "off")


ggsave(
  path = '../figure/01_fig_analysis_off-target-sgrna/',
  width = 5,
  height = 6,
  dpi = 1000,
  plot = boxplot_percent_identity_paralog,
  filename = "removed_genes_paralogs_percent_identity.png")



```


We asked whether multi-targeting-/off-targeting sgRNAs tend to target specific paralogs
























## Does genes size is small ?

```{r}

# T2T_txdb <- makeTxDbFromGFF(file = "~/CRISPR-KO-GuideRefine/annotation_file/GCF_009914755.1_T2T-CHM13v2.0_genomic.gff.gz",
#                             dataSource = "NCBI T2T Assembly GCF_009914755.1_T2T-CHM13v2.0",
#                             organism = "Homo sapiens")

T2T_txdb <- loadDb("../data/T2T_data/TxDb_CHM13.sqlite")
columns(T2T_txdb)

# Extract gene coordinates
genes_gr <- genes(T2T_txdb)

# Filter for your list
genes_gr <- genes_gr[names(genes_gr) %in% removed_genes_all_libraries]

genes_gr$span_length <- width(genes_gr)

test <- as.data.frame(genes_gr)

```















