---
title: "Analysis of multi-target and off-target sgRNAs across libraries"
author: "Author: Stefanus Bernard"
date: "Date: `r format(Sys.time(), '%d %B %Y')`"
output: 
  html_document :
    theme : cerulean
    toc : true
    toc_float : true
    number_section : true
---

```{r setup, include=FALSE}
library(tidyverse)
library(VennDiagram)
library(ggvenn)
library(ggbreak)
library(BSgenome)
library(BSgenome.Hsapiens.UCSC.hg38)
library(BSgenome.Hsapiens.NCBI.T2TCHM13v2.0)

source('~/CRISPR-KO-GuideRefine/GuideRefine_functions.R')
source("./00_CRISPR-KO-library-survey-functions.R")
```

# Import alignment data of each sgRNA libraries and library

```{r}
# import sgRNA library
# WE SHOULD DISPOSE ANY CONTROL FROM THE SGRNA LIBRARY AS THEY ARE NOT USED IN THIS ANALYSIS

terms <- c("CONTROL", "Control", "control", "INTRON", "Intron", "intron", "LacZ", "luciferase", "NO_SITE", "ONE_INTERGENIC_SITE")
# lacZ and luciferase are the control in Toronto V3 library
terms_for_filtering <- paste(terms, collapse = "|")

avana_library_meyers <- import_sgrna_library("~/CRISPR-KO-GuideRefine/public_crispr_library/avana_library.tsv")
brunello_library <- import_sgrna_library("~/CRISPR-KO-GuideRefine/public_crispr_library/broadgpp-brunello-library-contents.tsv")
toronto_library <- import_sgrna_library("~/CRISPR-KO-GuideRefine/public_crispr_library/tkov3_guide_sequence.tsv")
yusa_library <- import_sgrna_library("~/CRISPR-KO-GuideRefine/public_crispr_library/yusa_hcrispr_ko_grnas.tsv")
jacquere_library <- import_sgrna_library("~/CRISPR-KO-GuideRefine/public_crispr_library/Jacquere_PerGuideAnnotations_Quota4.tsv")

annotation <- "~/CRISPR-KO-GuideRefine/annotation_file/T2T-CHM13v2.0_gene_annot_granges.rds"
genome_type_version <- "T2T-CHM13"
bsgenome <- BSgenome.Hsapiens.NCBI.T2TCHM13v2.0

# T2T-CHM13
avana_alignment_meyers <- import_sgrna_library_alignment("~/CRISPR-KO-GuideRefine/object_intermediate/T2T-CHM13_bsgenome/avana_library_aln.csv", "T2T-CHM13")
brunello_alignment <- import_sgrna_library_alignment("~/CRISPR-KO-GuideRefine/object_intermediate/T2T-CHM13_bsgenome/broadgpp-brunello-library-contents_aln.csv", "T2T-CHM13")
toronto_alignment <- import_sgrna_library_alignment("~/CRISPR-KO-GuideRefine/object_intermediate/T2T-CHM13_bsgenome/tkov3_guide_sequence_aln.csv", "T2T-CHM13")
yusa_alignment <- import_sgrna_library_alignment("~/CRISPR-KO-GuideRefine/object_intermediate/T2T-CHM13_bsgenome/yusa_hcrispr_ko_grnas_aln.csv", "T2T-CHM13")
jacquere_alignment <- import_sgrna_library_alignment("~/CRISPR-KO-GuideRefine/object_intermediate/T2T-CHM13_bsgenome/Jacquere_PerGuideAnnotations_Quota4_aln.csv", "T2T-CHM13")

# list_removed_sgrna sgRNAs
disposed_avana <- import_list_off_target_guides("~/CRISPR-KO-GuideRefine/output_cleaning/T2T-CHM13/avana_library_disposed_sgRNAs.tsv")
disposed_brunello <- import_list_off_target_guides("~/CRISPR-KO-GuideRefine/output_cleaning/T2T-CHM13/broadgpp-brunello-library-contents_disposed_sgRNAs.tsv")
disposed_toronto <- import_list_off_target_guides("~/CRISPR-KO-GuideRefine/output_cleaning/T2T-CHM13/tkov3_guide_sequence_disposed_sgRNAs.tsv")
disposed_yusa <- import_list_off_target_guides("~/CRISPR-KO-GuideRefine/output_cleaning/T2T-CHM13/yusa_hcrispr_ko_grnas_disposed_sgRNAs.tsv")
disposed_jacquere <- import_list_off_target_guides("~/CRISPR-KO-GuideRefine/output_cleaning/T2T-CHM13/Jacquere_PerGuideAnnotations_Quota4_disposed_sgRNAs.tsv")

```

## Import paralog data from ENSEMBL 115

```{r}

# import predicted paralog pairs from ENSEMBL 115
ensembl_paralog <- ensembl_paralog_gene_pairs()

paralog_pairs <- ensembl_paralog[[1]]
paralog_individual <- ensembl_paralog[[2]]

```

# Classify the sgRNAs in each libraries

```{r}

avana_classification <- off_target_classification(avana_alignment_meyers, disposed_avana["pam_distal_double_mismatch_guides"])
brunello_classification <- off_target_classification(brunello_alignment, disposed_brunello["pam_distal_double_mismatch_guides"])
toronto_classification <- off_target_classification(toronto_alignment, disposed_toronto["pam_distal_double_mismatch_guides"])
yusa_classification <- off_target_classification(yusa_alignment, disposed_yusa["pam_distal_double_mismatch_guides"])
jacquere_classification <- off_target_classification(jacquere_alignment, disposed_jacquere["pam_distal_double_mismatch_guides"])

```

# Stratify multi-target sgRNAs (perfect alignment for each sgRNAs)

to investigate multi-target guides alignment to cell fitness, we exclude any single-mismatch guides, pam-distal double mismatch guides. To prevent confounding our analysis

```{r}

avana_stratify_multi_target <- multi_target_alignment_bin(avana_classification[[1]], c("perfect", "multi-target guides", "non-targeting"))
write_csv(avana_stratify_multi_target[[1]], "../data/sgrna_lfc_data/avana_data/avana_stratify_multi_target.csv")

brunello_stratify_multi_target <- multi_target_alignment_bin(brunello_classification[[1]], c("perfect", "multi-target guides", "non-targeting"))
write_csv(brunello_stratify_multi_target[[1]], "../data/sgrna_lfc_data/brunello_data/brunello_stratify_multi_target.csv")

toronto_stratify_multi_target <- multi_target_alignment_bin(toronto_classification[[1]], c("perfect", "multi-target guides", "non-targeting"))
write_csv(toronto_stratify_multi_target[[1]], "../data/sgrna_lfc_data/tkov3_data/tkov3_stratify_multi_target.csv")

yusa_stratify_multi_target <- multi_target_alignment_bin(yusa_classification[[1]], c("perfect", "multi-target guides", "non-targeting"))
write_csv(yusa_stratify_multi_target[[1]], "../data/sgrna_lfc_data/yusa_data/yusa_stratify_multi_target.csv")

jacquere_stratify_multi_target <- multi_target_alignment_bin(jacquere_classification[[1]], c("perfect", "multi-target guides", "non-targeting"))
write_csv(jacquere_stratify_multi_target[[1]], "../data/sgrna_lfc_data/jacquere_data/jacquere_stratify_multi_target.csv")


```

# Stratify single-mismatch sgRNAs (sgRNA align with single mismatch)

to investigate single-mismatch off-target alignment to cell fitness, we exclude any multi-targeting guides, pam-distal double mismatch guides. To prevent confounding our analysis

```{r}

avana_stratify_single_mismatch <- single_mismatch_off_target_alignment_bin(avana_classification[[1]], c("perfect", "single mismatch", "non-targeting"))
write_csv(avana_stratify_single_mismatch[[1]], "../data/sgrna_lfc_data/avana_data/avana_stratify_single_mismatch.csv")

brunello_stratify_single_mismatch <- single_mismatch_off_target_alignment_bin(brunello_classification[[1]], c("perfect", "single mismatch", "non-targeting"))
write_csv(brunello_stratify_single_mismatch[[1]], "../data/sgrna_lfc_data/brunello_data/brunello_stratify_single_mismatch.csv")

toronto_stratify_single_mismatch <- single_mismatch_off_target_alignment_bin(toronto_classification[[1]], c("perfect", "single mismatch", "non-targeting"))
write_csv(toronto_stratify_single_mismatch[[1]], "../data/sgrna_lfc_data/tkov3_data/tkov3_stratify_single_mismatch.csv")

yusa_stratify_single_mismatch <- single_mismatch_off_target_alignment_bin(yusa_classification[[1]], c("perfect", "single mismatch", "non-targeting"))
write_csv(yusa_stratify_single_mismatch[[1]], "../data/sgrna_lfc_data/yusa_data/yusa_stratify_single_mismatch.csv")

jacquere_stratify_single_mismatch <- single_mismatch_off_target_alignment_bin(jacquere_classification[[1]], c("perfect", "single mismatch", "non-targeting"))
write_csv(jacquere_stratify_single_mismatch[[1]], "../data/sgrna_lfc_data/jacquere_data/jacquere_stratify_single_mismatch.csv")


```

# Export the bar chart showing stratification of multi-target sgRNAs

```{r}

export_images <- function(object, filename_dir, genome_type) {
  ggsave(
    path = './',
    width = 13,
    height = 12,
    dpi = 300,
    plot = object,
    filename = paste(filename_dir, genome_type, ".png", sep=""))
}

# stratification of multi-target sgRNAs

export_images(visualize_off_target_alignment(avana_stratify_multi_target[[2]], "Number of on-target alignments", aln_break_2 = 2, aln_break_1 = 1),
              "../figure/02_fig_analysis_paralog_off-target-sgrna/avana_stratification_multi_target_",
              genome_type_version)

export_images(visualize_off_target_alignment(brunello_stratify_multi_target[[2]], "Number of on-target alignments", aln_break_2 = 2, aln_break_1 = 1),
              "../figure/02_fig_analysis_paralog_off-target-sgrna/brunello_stratification_multi_target_",
              genome_type_version)

export_images(visualize_off_target_alignment(toronto_stratify_multi_target[[2]], "Number of on-target alignments", aln_break_2 = 2, aln_break_1 = 1),
              "../figure/02_fig_analysis_paralog_off-target-sgrna/toronto_stratification_multi_target_",
              genome_type_version)

export_images(visualize_off_target_alignment(yusa_stratify_multi_target[[2]], "Number of on-target alignments", aln_break_2 = 2, aln_break_1 = 1),
              "../figure/02_fig_analysis_paralog_off-target-sgrna/yusa_stratification_multi_target_",
              genome_type_version)

export_images(visualize_off_target_alignment(jacquere_stratify_multi_target[[2]], "Number of on-target alignments", aln_break_2 = 2, aln_break_1 = 1),
              "../figure/02_fig_analysis_paralog_off-target-sgrna/jacquere_stratification_multi_target_",
              genome_type_version)

# stratification of single mismatch sgRNAs

export_images(visualize_off_target_alignment(avana_stratify_single_mismatch[[2]], "Number of single mismatch off-target alignments", aln_break_2 = 1, aln_break_1 = 0),
              "../figure/02_fig_analysis_paralog_off-target-sgrna/avana_stratification_single_mismatch_",
              genome_type_version)

export_images(visualize_off_target_alignment(brunello_stratify_single_mismatch[[2]], "Number of single mismatch off-target alignments", aln_break_2 = 1, aln_break_1 = 0),
              "../figure/02_fig_analysis_paralog_off-target-sgrna/brunello_stratification_single_mismatch_",
              genome_type_version)

export_images(visualize_off_target_alignment(toronto_stratify_single_mismatch[[2]], "Number of single mismatch off-target alignments", aln_break_2 = 1, aln_break_1 = 0),
              "../figure/02_fig_analysis_paralog_off-target-sgrna/toronto_stratification_single_mismatch_",
              genome_type_version)

export_images(visualize_off_target_alignment(yusa_stratify_single_mismatch[[2]], "Number of single mismatch off-target alignments", aln_break_2 = 1, aln_break_1 = 0),
              "../figure/02_fig_analysis_paralog_off-target-sgrna/yusa_stratification_single_mismatch_",
              genome_type_version)

export_images(visualize_off_target_alignment(jacquere_stratify_single_mismatch[[2]], "Number of single mismatch off-target alignments", aln_break_2 = 1, aln_break_1 = 0),
              "../figure/02_fig_analysis_paralog_off-target-sgrna/jacquere_stratification_single_mismatch_",
              genome_type_version)

```


# What is percentage of sgRNAs that perfectly target multiple genomics sites and protein-coding genes perfectly?

## Identify the target location of multi-target sgRNAs (protein-coding genes/other genomic loci)

```{r}


list_guides <- avana_stratify_multi_target[[1]] %>% filter(alignment_bin %in% c(2,3,4,5,"> 5")) %>% pull(sgRNA)
avana_annotated <- annotate_all_sgrna(avana_alignment_meyers,                                       
                                      annotation,
                                      bsgenome,
                                      genome_type_version,
                                      list_guides,
                                      paralog_individual,
                                      paralog_pairs)

list_guides <- brunello_stratify_multi_target[[1]] %>% filter(alignment_bin %in% c(2,3,4,5,"> 5")) %>% pull(sgRNA)
brunello_annotated <- annotate_all_sgrna(brunello_alignment,
                                           annotation,
                                           bsgenome,
                                           genome_type_version,
                                           list_guides,
                                           paralog_individual,
                                           paralog_pairs)

list_guides <- toronto_stratify_multi_target[[1]] %>% filter(alignment_bin %in% c(2,3,4,5,"> 5")) %>% pull(sgRNA)
toronto_annotated <- annotate_all_sgrna(toronto_alignment,
                                         annotation,
                                         bsgenome,
                                         genome_type_version,
                                         list_guides,
                                         paralog_individual,
                                         paralog_pairs)

list_guides <- yusa_stratify_multi_target[[1]] %>% filter(alignment_bin %in% c(2,3,4,5,"> 5")) %>% pull(sgRNA)
yusa_annotated <- annotate_all_sgrna(yusa_alignment,
                                       annotation,
                                       bsgenome,
                                       genome_type_version,
                                       list_guides,
                                       paralog_individual,
                                       paralog_pairs)

list_guides <- jacquere_stratify_multi_target[[1]] %>% filter(alignment_bin %in% c(2,3,4,5,"> 5")) %>% pull(sgRNA)
jacquere_annotated <- annotate_all_sgrna(jacquere_alignment,
                                       annotation,
                                       bsgenome,
                                       genome_type_version,
                                       list_guides,
                                       paralog_individual,
                                       paralog_pairs)

```

# Question 1: Are multitargeting effects more likely to target pairs of paralogs?
# Question 2: Are genes with paralogs more likely to be subject to multi-targeting effects?

## Identify the double-target guides targeting paralog pairs and non-paralog pairs across libraries

```{r}

dbl_target_avana <- avana_stratify_multi_target[[1]] %>% filter(alignment_bin == 2) %>% pull(sgRNA)
dbl_target_brunello <- brunello_stratify_multi_target[[1]] %>% filter(alignment_bin == 2) %>% pull(sgRNA)
dbl_target_toronto <- toronto_stratify_multi_target[[1]] %>% filter(alignment_bin == 2) %>% pull(sgRNA)
dbl_target_yusa <- yusa_stratify_multi_target[[1]] %>% filter(alignment_bin == 2) %>% pull(sgRNA)
dbl_target_jacquere <- jacquere_stratify_multi_target[[1]] %>% filter(alignment_bin == 2) %>% pull(sgRNA)

# list_query_guides_paralog <- summarize_paralog_targets(dbl_target_avana, paralog_individual, "avana")[[2]] %>% filter(paralog_indicator %in% c("paralog")) %>% pull(sgRNA)
avana_multi_paralog <- find_multi_target_paralog(avana_annotated, dbl_target_avana, "avana")

# list_query_guides_paralog <- summarize_paralog_targets(dbl_target_brunello, paralog_individual, "brunello")[[2]] %>% filter(paralog_indicator == "paralog") %>% pull(sgRNA)
brunello_multi_paralog <- find_multi_target_paralog(brunello_annotated, dbl_target_brunello, "brunello")

# list_query_guides_paralog <- summarize_paralog_targets(dbl_target_toronto, paralog_individual, "toronto_v3")[[2]] %>% filter(paralog_indicator == "paralog") %>% pull(sgRNA)
toronto_multi_paralog <- find_multi_target_paralog(toronto_annotated, dbl_target_toronto, "toronto")

# list_query_guides_paralog <- summarize_paralog_targets(dbl_target_yusa, paralog_individual, "yusa")[[2]] %>% filter(paralog_indicator == "paralog") %>% pull(sgRNA)
yusa_multi_paralog <- find_multi_target_paralog(yusa_annotated, dbl_target_yusa, "yusa")

jacquere_multi_paralog <- find_multi_target_paralog(jacquere_annotated, dbl_target_jacquere, "jacquere")

# Summary table of double-target sgRNAs targeting paralogs
table_dbl_target_paralog <- rbind(brunello_multi_paralog,
                                  toronto_multi_paralog,
                                  yusa_multi_paralog,
                                  avana_multi_paralog,
                                  jacquere_multi_paralog)

my_palette <- brewer.pal(n = 3, name = "Set2")

# Barplot of percentage of sgRNA targeting paralog pairs and non-paralog pairs across libraries
barplot_paralog <- table_dbl_target_paralog %>%
  arrange(library_name) %>%
  select(-c(total_dbl_target_sgrna, count_sgrna_excluded, count_sgrna_paralog, count_sgrna_other_genes, count_sgrna_non_coding)) %>%
  pivot_longer(c(percentage_sgrna_excluded, percentage_sgrna_paralog, percentage_sgrna_other_genes, percentage_sgrna_non_coding), names_to = 'names', values_to = 'percent') %>%
  mutate(
    names = factor(
      names,
      levels = c("percentage_sgrna_excluded",
                 "percentage_sgrna_non_coding",
                 "percentage_sgrna_other_genes",
                 "percentage_sgrna_paralog")),
    library_name = factor(library_name, levels = c("brunello", "toronto", "yusa", "avana", "jacquere"))) %>%
  ggplot(aes(library_name, percent, fill = names)) +
  geom_bar(position = "stack", stat = "identity") +
  theme_minimal() +
  labs(title = "", x = "", y = "Off-target location of double-target guides (%)", fill = "") +
  scale_x_discrete(labels = c('brunello' = 'Brunello',
                'yusa' = 'Yusa \n(Project Score)',
                'toronto' = 'TKOv3',
                'avana' = 'Avana',
                'jacquere' = 'Jacquere')) +
  scale_fill_manual(name = "Target location", 
                    labels = c("percentage_sgrna_paralog" = "Paralog + paralog (ENSEMBL 115 pairs)",
                               "percentage_sgrna_other_genes" = "Both target protein-coding genes",
                               "percentage_sgrna_non_coding" = "Protein-coding gene + non-coding region",
                               "percentage_sgrna_excluded" = "Guides excluded because of additional mismatches"),
                    values = c("percentage_sgrna_paralog" = "#009E73",
                              "percentage_sgrna_other_genes" = "#0072B2",
                              "percentage_sgrna_non_coding" = "#E69F00",
                              "percentage_sgrna_excluded" = "lightgrey")) +
  theme(axis.text.x = element_text(size = 20, colour = 'black'),
      axis.text.y = element_text(size = 20, colour = 'black'),
      axis.title.x = element_text(size = 20, margin = margin(r = 10)),
      axis.title.y = element_text(size = 20, margin = margin(r = 10)),
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      legend.title = element_text(size = 20),
      legend.text = element_text(size = 20),
      axis.line.x = element_line(size = 0.5),
      axis.line.y = element_line(size = 0.5),
      line = element_line((size = 4), colour = 'black')) +
  geom_text(aes(label = ifelse(percent > 2, percent, "")), position = position_stack(vjust = 0.5), size = 7)

barplot_paralog

ggsave(
    path = './',
    width = 18,
    height = 10,
    dpi = 300,
    plot = barplot_paralog,
    filename = paste('../figure/02_fig_analysis_paralog_off-target-sgrna/barplot_double_target_guides_T2T.png', sep = ''))

```


From 2,760 double-target guides in Avana library:
1. 2009 double-target guides originally target gene with paralog pairs
2. 751 double-target guides originally target singleton gene

From 2009 double-target guides that originally target a gene that has paralog pairs:
1. 1357 guides are excluded because it has additional single or double mismatches (67.51%)
2. 212 guides have their pair target annotated as paralog (paralog + paralog) (10.55%)
3. 14 guides have their pair target as singleton (paralog + singleton) (0.7%)
4. 427 guides target non-coding have their pair target non-coding (paralog + non-coding) (21.24%)






























































