---
title: "Analysis of multi-target and off-target sgRNAs across libraries"
author: "Author: Stefanus Bernard"
date: "Date: `r format(Sys.time(), '%d %B %Y')`"
output: 
  html_document :
    theme : cerulean
    toc : true
    toc_float : true
    number_section : true
---

```{r setup, include=FALSE}
library(tidyverse)
library(VennDiagram)
library(ggvenn)
library(ggbreak)
library(BSgenome)
library(BSgenome.Hsapiens.UCSC.hg38)
library(BSgenome.Hsapiens.NCBI.T2TCHM13v2.0)

source('~/CRISPR-KO-GuideRefine/GuideRefine_functions.R')
source("./00_CRISPR-KO-library-survey-functions.R")
```

# Import alignment data of each sgRNA libraries and library

```{r}
# import sgRNA library
# WE SHOULD DISPOSE ANY CONTROL FROM THE SGRNA LIBRARY AS THEY ARE NOT USED IN THIS ANALYSIS

terms <- c("CONTROL", "Control", "control", "INTRON", "Intron", "intron", "LacZ", "luciferase")
# lacZ and luciferase are the control in Toronto V3 library

terms_for_filtering <- paste(terms, collapse = "|")

avana_library_meyers <- import_sgrna_library("~/CRISPR-KO-GuideRefine/public_crispr_library/avana_library.tsv")
brunello_library <- import_sgrna_library("~/CRISPR-KO-GuideRefine/public_crispr_library/broadgpp-brunello-library-contents.tsv")
toronto_library <- import_sgrna_library("~/CRISPR-KO-GuideRefine/public_crispr_library/tkov3_guide_sequence.tsv")
yusa_library <- import_sgrna_library("~/CRISPR-KO-GuideRefine/public_crispr_library/yusa_hcrispr_ko_grnas.tsv")


# some parameters
# annotation <- "../CRISPR-KO-GuideRefine/annotation_file/CCDS.20221027.txt"
# genome_type_version <- "hg38"
# bsgenome <- BSgenome.Hsapiens.UCSC.hg38

annotation <- "~/CRISPR-KO-GuideRefine/annotation_file/T2T-CHM13v2.0_gene_annot_granges.rds"
genome_type_version <- "T2T-CHM13"
bsgenome <- BSgenome.Hsapiens.NCBI.T2TCHM13v2.0


# hg38
# avana_alignment_meyers <- import_sgrna_library_alignment("~/CRISPR-KO-GuideRefine/object_intermediate/hg38/avana_library_aln.csv", genome_type_version)
# brunello_alignment <- import_sgrna_library_alignment("~/CRISPR-KO-GuideRefine/object_intermediate/hg38/broadgpp-brunello-library-contents_aln.csv", genome_type_version)
# toronto_alignment <- import_sgrna_library_alignment("~/CRISPR-KO-GuideRefine/object_intermediate/hg38/tkov3_guide_sequence_aln.csv", genome_type_version)
# yusa_alignment <- import_sgrna_library_alignment("~/CRISPR-KO-GuideRefine/object_intermediate/hg38/yusa_hcrispr_ko_grnas_aln.csv", genome_type_version)

# T2T-CHM13
avana_alignment_meyers <- import_sgrna_library_alignment("~/CRISPR-KO-GuideRefine/object_intermediate/T2T-CHM13_bsgenome/avana_library_aln.csv", "T2T-CHM13")
brunello_alignment <- import_sgrna_library_alignment("~/CRISPR-KO-GuideRefine/object_intermediate/T2T-CHM13_bsgenome/broadgpp-brunello-library-contents_aln.csv", "T2T-CHM13")
toronto_alignment <- import_sgrna_library_alignment("~/CRISPR-KO-GuideRefine/object_intermediate/T2T-CHM13_bsgenome/tkov3_guide_sequence_aln.csv", "T2T-CHM13")
yusa_alignment <- import_sgrna_library_alignment("~/CRISPR-KO-GuideRefine/object_intermediate/T2T-CHM13_bsgenome/yusa_hcrispr_ko_grnas_aln.csv", "T2T-CHM13")

```

## Import paralog data from Barbara's paper

```{r}

# import predicted paralog pairs from Barbara's paper

# barbara_paralog <- list_paralog_gene_pairs()
ensembl_paralog <- ensembl_paralog_gene_pairs()

paralog_pairs <- ensembl_paralog[[1]]
paralog_individual <- ensembl_paralog[[2]]

```

# Classify the sgRNAs in each libraries

```{r}

avana_classification <- off_target_classification(avana_alignment_meyers)
brunello_classification <- off_target_classification(brunello_alignment)
toronto_classification <- off_target_classification(toronto_alignment)
yusa_classification <- off_target_classification(yusa_alignment)

```

# Stratify multi-target sgRNAs (perfect alignment for each sgRNAs)

```{r}

avana_stratify_multi_target <- off_target_alignment_bin(avana_classification[[1]], c("perfect", "multi-target guides", "non-targeting"))
write_csv(avana_stratify_multi_target[[1]], "../data/sgrna_lfc_data/avana_data/avana_stratify_multi_target.csv")

brunello_stratify_multi_target <- off_target_alignment_bin(brunello_classification[[1]], c("perfect", "multi-target guides", "non-targeting"))
write_csv(brunello_stratify_multi_target[[1]], "../data/sgrna_lfc_data/brunello_data/brunello_stratify_multi_target.csv")

toronto_stratify_multi_target <- off_target_alignment_bin(toronto_classification[[1]], c("perfect", "multi-target guides", "non-targeting"))
write_csv(toronto_stratify_multi_target[[1]], "../data/sgrna_lfc_data/tkov3_data/tkov3_stratify_multi_target.csv")

yusa_stratify_multi_target <- off_target_alignment_bin(yusa_classification[[1]], c("perfect", "multi-target guides", "non-targeting"))
write_csv(yusa_stratify_multi_target[[1]], "../data/sgrna_lfc_data/yusa_data/yusa_stratify_multi_target.csv")

```

# Export the bar chart showing stratification of multi-target sgRNAs

```{r}

export_images <- function(object, filename_dir, genome_type) {
  ggsave(
    path = './',
    width = 13,
    height = 11,
    dpi = 300,
    plot = object,
    filename = paste(filename_dir, genome_type, ".png", sep=""))
}

export_images(visualize_off_target_alignment(avana_stratify_multi_target[[2]], "Number of on-target alignments"),
              "../figure/02_fig_analysis_paralog_off-target-sgrna/avana_stratification_",
              genome_type_version)
export_images(visualize_off_target_alignment(brunello_stratify_multi_target[[2]], "Number of on-target alignments"),
              "../figure/02_fig_analysis_paralog_off-target-sgrna/brunello_stratification_",
              genome_type_version)
export_images(visualize_off_target_alignment(toronto_stratify_multi_target[[2]], "Number of on-target alignments"),
              "../figure/02_fig_analysis_paralog_off-target-sgrna/toronto_stratification_",
              genome_type_version)
export_images(visualize_off_target_alignment(yusa_stratify_multi_target[[2]], "Number of on-target alignments"),
              "../figure/02_fig_analysis_paralog_off-target-sgrna/yusa_stratification_",
              genome_type_version)
```


# What is percentage of sgRNAs that perfectly target multiple genomics sites and protein-coding genes perfectly?

## Identify the target location of multi-target sgRNAs (protein-coding genes/other genomic loci)

```{r}


list_guides <- avana_stratify_multi_target[[1]] %>% filter(alignment_bin %in% c(2,3,4,5,"> 5")) %>% pull(sgRNA)
avana_multi_protein_coding_genes <- off_target_annotation(avana_alignment_meyers,                                       
                                                          annotation,
                                                          bsgenome,
                                                          genome_type_version,
                                                          avana_library_meyers, 
                                                          list_guides,
                                                          0,
                                                          "avana")

list_guides <- brunello_stratify_multi_target[[1]] %>% filter(alignment_bin %in% c(2,3,4,5,"> 5")) %>% pull(sgRNA)
brunello_multi_protein_coding_genes <- off_target_annotation(brunello_alignment, 
                                                             annotation,
                                                             bsgenome,
                                                             genome_type_version,
                                                             brunello_library, 
                                                             list_guides,
                                                             0,
                                                             "brunello")

list_guides <- toronto_stratify_multi_target[[1]] %>% filter(alignment_bin %in% c(2,3,4,5,"> 5")) %>% pull(sgRNA)
toronto_multi_protein_coding_genes <- off_target_annotation(toronto_alignment, 
                                                            annotation,
                                                            bsgenome,
                                                            genome_type_version,
                                                            toronto_library, 
                                                            list_guides,
                                                            0,
                                                            "toronto")

list_guides <- yusa_stratify_multi_target[[1]] %>% filter(alignment_bin %in% c(2,3,4,5,"> 5")) %>% pull(sgRNA)
yusa_multi_protein_coding_genes <- off_target_annotation(yusa_alignment, 
                                                         annotation,
                                                         bsgenome,
                                                         genome_type_version,
                                                         yusa_library, 
                                                         list_guides,
                                                         0,
                                                         "yusa")

```

# Question 1: Are multitargeting effects more likely to target pairs of paralogs?
# Question 2: Are genes with paralogs more likely to be subject to multi-targeting effects?

## How many double-target guides originally target paralog and singleton ?

```{r}

dbl_target_avana <- avana_stratify_multi_target[[1]] %>% filter(alignment_bin == 2) %>% pull(sgRNA)
dbl_target_brunello <- brunello_stratify_multi_target[[1]] %>% filter(alignment_bin == 2) %>% pull(sgRNA)
dbl_target_toronto <- toronto_stratify_multi_target[[1]] %>% filter(alignment_bin == 2) %>% pull(sgRNA)
dbl_target_yusa <- yusa_stratify_multi_target[[1]] %>% filter(alignment_bin == 2) %>% pull(sgRNA)


summarize_paralog_targets <- function(sgRNA, paralogs, library) {
  # extract gene names
  genes <- str_extract(sgRNA, "(?<=sg_?)[A-Za-z0-9]+")
  
  # classify paralog vs non-paralog
  paralog_indicator <- ifelse(genes %in% paralogs, "paralog", "singleton")
  
  df_dbl_target_paralog <- data.frame(sgRNA, genes, paralog_indicator, library)
  
  # summary counts + percentages
  df <- data.frame(category = paralog_indicator) %>%
    count(category) %>%
    dplyr::rename("count" = "n") %>%
    mutate(percentage = round(count / sum(count) * 100, 2),
           library = library)
  
  return(list(df, df_dbl_target_paralog))
}

avana_dbl_target_genes_paralog <- summarize_paralog_targets(dbl_target_avana, paralog_individual, "avana")[[1]]
brunello_dbl_target_genes_paralog <- summarize_paralog_targets(dbl_target_brunello, paralog_individual, "brunello")[[1]]
toronto_dbl_target_genes_paralog <- summarize_paralog_targets(dbl_target_toronto, paralog_individual, "toronto_v3")[[1]]
yusa_dbl_target_genes_paralog <- summarize_paralog_targets(dbl_target_yusa, paralog_individual, "yusa")[[1]]

df_summary = rbind(avana_dbl_target_genes_paralog, 
                   brunello_dbl_target_genes_paralog, 
                   toronto_dbl_target_genes_paralog, 
                   yusa_dbl_target_genes_paralog)

df_summary <- df_summary %>% mutate(library = factor(library, levels = c("brunello", "toronto_v3", "yusa", "avana")),
                                    category = factor(category, levels = c("singleton", "paralog")))


df_paralog_dbl_target_genes <- ggplot(df_summary, aes(x = library, y = percentage, fill = category)) +
      geom_bar(stat = "identity", position = "stack") +
      scale_fill_discrete(
          name = "",
          labels = c("Singleton", "Paralog")) +
        theme_minimal() +
        labs(title = "", x = "", y = "Intended target gene of double-target guides (%)", fill = "Target") +
        scale_x_discrete(labels = c("brunello" = "Brunello",
                                     "toronto_v3" = "TKOv3",
                                     "yusa" = "Yusa \n(Project Score)",
                                     "avana" = "Avana")) +
        geom_text(aes(label = percentage), position = position_stack(), vjust = 2, size = 8) +
        theme(axis.text.x = element_text(size = 25, vjust = 0.7, colour = "black"),
              axis.text.y = element_text(size = 25, colour = "black"),
              axis.title.y = element_text(size = 25, vjust = 0.5, margin = margin(r = 20)),
              panel.grid.major = element_blank(),
              panel.grid.minor = element_blank(),
              axis.line.x = element_line(size = 0.5),
              axis.line.y = element_line(size = 0.5),
              legend.title = element_text(size = 25),
              legend.text = element_text(size = 25),
              plot.margin = margin(1, 1, 1, 1), 
              line = element_line((size = 4), colour = "black"),
              legend.spacing.x = unit(1, 'cm')) +
      ylim(0, 100)

df_paralog_dbl_target_genes

ggsave(
    path = './',
    width = 12,
    height = 10,
    dpi = 300,
    plot = df_paralog_dbl_target_genes,
    filename = paste('../figure/02_fig_analysis_paralog_off-target-sgrna/dbl_target_guides_original_genes_paralog_T2T.png', sep = ''))

```



## Identify the double-target guides targeting paralog pairs and non-paralog pairs across libraries

```{r}

# Using annotated exons from GENCODE (comprehensive gene annotation, human v28), we found that 2503 (95.2%) of the double-target guides have both targets located in coding regions (Fortin et al, 2019) (totally different than my own findings)

# Among the 2503 double-target guides, 297 (11.9%) guides have their pair of targets annotated as being paralogs using the PANTHER database (Fortin et al, 2019)

# ALL NUMBERS BELOW ARE IN HG38

# Avana paralog analysis for double-target guides
# list_query_guides_paralog <- avana_stratify_multi_target[[1]] %>% filter(alignment_bin == 2) %>% pull(sgRNA)

list_query_guides_paralog <- summarize_paralog_targets(dbl_target_avana, paralog_individual, "avana")[[2]] %>% filter(paralog_indicator == "paralog") %>% pull(sgRNA)

avana_multi_paralog <- find_multi_target_paralog(avana_multi_protein_coding_genes[[1]], list_query_guides_paralog, paralog_pairs, avana_library_meyers, "avana")

# the number of query guides (2878) and the total number of paralog classification (694+244+2017=2955) does not match, kindly check this function
# corrected number is 2818

# Brunello paralog analysis for double-target guides
# list_query_guides_paralog <- brunello_stratify_multi_target[[1]] %>% filter(alignment_bin == 2) %>% pull(sgRNA)

list_query_guides_paralog <- summarize_paralog_targets(dbl_target_brunello, paralog_individual, "brunello")[[2]] %>% filter(paralog_indicator == "paralog") %>% pull(sgRNA)

brunello_multi_paralog <- find_multi_target_paralog(brunello_multi_protein_coding_genes[[1]], list_query_guides_paralog, paralog_pairs, brunello_library, "brunello")

# the number of query guides (1747) and the total number of paralog classification (810+167+703=1680) does not match, kindly check this function
# corrected number is 1615

# toronto paralog analysis for double-target guides
# list_query_guides_paralog <- toronto_stratify_multi_target[[1]] %>% filter(alignment_bin == 2) %>% pull(sgRNA)

list_query_guides_paralog <- summarize_paralog_targets(dbl_target_toronto, paralog_individual, "toronto_v3")[[2]] %>% filter(paralog_indicator == "paralog") %>% pull(sgRNA)

toronto_multi_paralog <- find_multi_target_paralog(toronto_multi_protein_coding_genes[[1]], list_query_guides_paralog, paralog_pairs, toronto_library, "toronto")

# the number of query guides (1773) and the total number of paralog classification (57+124+1688=1869) does not match, kindly check this function
# corrected number is 1771

# yusa paralog analysis for double-target guides
# list_query_guides_paralog <- yusa_stratify_multi_target[[1]] %>% filter(alignment_bin == 2) %>% pull(sgRNA)

list_query_guides_paralog <- summarize_paralog_targets(dbl_target_yusa, paralog_individual, "yusa")[[2]] %>% filter(paralog_indicator == "paralog") %>% pull(sgRNA)

yusa_multi_paralog <- find_multi_target_paralog(yusa_multi_protein_coding_genes[[1]], list_query_guides_paralog, paralog_pairs, yusa_library, "yusa")

# the number of query guides (3054) and the total number of paralog classification (605+262+2253=3120) does not match, kindly check this function
# corrected number is 2967

# Summary table of double-target sgRNAs targeting paralogs
table_dbl_target_paralog <- rbind(brunello_multi_paralog$summary_count_df,
                                  toronto_multi_paralog$summary_count_df,
                                  yusa_multi_paralog$summary_count_df,
                                  avana_multi_paralog$summary_count_df)

my_palette <- brewer.pal(n = 3, name = "Set2")

# Barplot of percentage of sgRNA targeting paralog pairs and non-paralog pairs across libraries
barplot_paralog <- table_dbl_target_paralog %>%
  arrange(library_name) %>%
  select(-c(count_sgrna_paralog, count_sgrna_other_genes, count_sgrna_other_genomic_loci)) %>%
  pivot_longer(c(percentage_sgrna_paralog, percentage_sgrna_other_genes, percentage_sgrna_other_genomic_loci), names_to = 'names', values_to = 'percent') %>%
  mutate(
    names = factor(
      names,
      levels = c("percentage_sgrna_other_genomic_loci",
                 "percentage_sgrna_other_genes",
                 "percentage_sgrna_paralog")),
    library_name = factor(library_name, levels = c("brunello", "toronto", "yusa", "avana"))) %>%
  ggplot(aes(library_name, percent, fill = names)) +
  geom_bar(position = "stack", stat = "identity") +
  theme_minimal() +
  labs(title = "", x = "", y = "Paralog + off-target location of double-target guides (%)", fill = "") +
  scale_x_discrete(labels = c('brunello' = 'Brunello',
                'yusa' = 'Yusa \n(Project Score)',
                'toronto' = 'TKOv3',
                'avana' = 'Avana')) +
  scale_fill_manual(labels = c("percentage_sgrna_paralog" = "Paralog + paralog (ENSEMBL pairs)",
                               "percentage_sgrna_other_genes" = "Paralog + singleton",
                               "percentage_sgrna_other_genomic_loci" = "Paralog + non-coding region"),
                    values = c("percentage_sgrna_paralog" = "#009E73",
                              "percentage_sgrna_other_genes" = "#0072B2",
                              "percentage_sgrna_other_genomic_loci" = "#E69F00")) +
  theme(axis.text.x = element_text(size = 20, colour = 'black'),
      axis.text.y = element_text(size = 20, colour = 'black'),
      axis.title.x = element_text(size = 20, margin = margin(r = 10)),
      axis.title.y = element_text(size = 20, margin = margin(r = 10)),
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      legend.title = element_text(size = 20),
      legend.text = element_text(size = 20),
      axis.line.x = element_line(size = 0.5),
      axis.line.y = element_line(size = 0.5),
      line = element_line((size = 4), colour = 'black')) +
  geom_text(aes(label = percent), position = position_stack(vjust = .5), size = 7)

barplot_paralog

ggsave(
    path = './',
    width = 12,
    height = 10,
    dpi = 300,
    plot = barplot_paralog,
    filename = paste('../figure/02_fig_analysis_paralog_off-target-sgrna/barplot_double_target_guides_T2T.png', sep = ''))


## Identify genes affected by double-target guides targeting paralog pairs across libraries

paralog_summary_df <- rbind(brunello_multi_paralog$paralog_affected_df,
                                  toronto_multi_paralog$paralog_affected_df,
                                  yusa_multi_paralog$paralog_affected_df,
                                  avana_multi_paralog$paralog_affected_df)

```


































































