---
title: "Analysis of commonly removed genes across libraries"
author: "Author: Stefanus Bernard"
date: "Date: `r format(Sys.time(), '%d %B %Y')`"
output: 
  html_document :
    theme : cerulean
    toc : true
    toc_float : true
    number_section : true
---

```{r setup, include=FALSE}
library(tidyverse)
library(stringr)
library(UpSetR)
library(GenomicFeatures)
library(HGNChelper)
library(BSgenome.Hsapiens.NCBI.T2TCHM13v2.0)
library(GenomeInfoDbData)
library(txdbmaker)
library(crisprVerse)
library(pheatmap)
library(ggpubr)

source("./00_CRISPR-KO-library-survey-functions.R")
```

## Import data containing removed genes from each library

```{r}

removed_genes <- read_csv("../data/biogrid_crispr_ko_data/important_data/removed_genes_all_library.csv")

removed_gene_list_combined <- list(
  Avana = na.omit(removed_genes[["avana_removed_genes_all"]]),
  Brunello = na.omit(removed_genes[["brunello_removed_genes_all"]]),
  Toronto = na.omit(removed_genes[["toronto_removed_genes_all"]]),
  Yusa = na.omit(removed_genes[["yusa_removed_genes_all"]]),
  Jacquere = na.omit(removed_genes[["jacquere_removed_genes_all"]])
)

# List of 520 protein-coding genes are removed across CRISPR-KO libraries, what are these genes?
removed_genes_all_libraries <- unique(Reduce(intersect, removed_gene_list_combined))
all_genes_but_not_removed <- read_tsv("https://storage.googleapis.com/public-download-files/hgnc/tsv/tsv/hgnc_complete_set.txt") %>%
  filter(locus_group == "protein-coding gene" & !symbol %in% removed_genes_all_libraries) %>%
  pull(symbol)

```

## UpSet plot for commonality of removed genes between 5 libraries

```{r}

png("../figure/01_fig_analysis_off-target-sgrna/removed_genes_all_libraries.png", width = 12000, height = 7500, res = 1000)

par(mar = c(5, 20, 4, 2))
upset(fromList(removed_gene_list_combined),
      order.by = "degree",
      main.bar.color = "#4B0092",
      sets.bar.color = "#1AFF1A",
      text.scale = 1.5)

dev.off()

```

## There are 520 genes removed across libraries
we asked why there are 520 genes failed/represented by fewer than 3 guides across libraries

## Hypothesis 1: maybe those genes are hard to target

First, we prepare the T2T-CHM13 txdb file

```{r}
# create txdb sqlite file for T2T-CHM13

T2T_txdb <- txdbmaker::makeTxDbFromGFF(file = "../../GuideRefine/annotation_file/GCF_009914755.1_T2T-CHM13v2.0_genomic.gff.gz",
                            dataSource = "NCBI T2T Assembly GCF_009914755.1_T2T-CHM13v2.0",
                            organism = "Homo sapiens")
saveDb(T2T_txdb, "../data/T2T_data/TxDb_T2T-CHM13v2.sqlite")

```

## We asked whether the size of genes and limited number of protospacer site contributes to the off-target activity of sgRNAs ?

```{r}
# Load TXdb file

T2T_txdb <- loadDb("../data/T2T_data/TxDb_T2T-CHM13v2.sqlite")
columns(T2T_txdb)

T2T_txdb

refseq_to_chr <- c(
  "NC_060925.1" = "1",
  "NC_060926.1" = "2",
  "NC_060927.1" = "3",
  "NC_060928.1" = "4",
  "NC_060929.1" = "5",
  "NC_060930.1" = "6",
  "NC_060931.1" = "7",
  "NC_060932.1" = "8",
  "NC_060933.1" = "9",
  "NC_060934.1" = "10",
  "NC_060935.1" = "11",
  "NC_060936.1" = "12",
  "NC_060937.1" = "13",
  "NC_060938.1" = "14",
  "NC_060939.1" = "15",
  "NC_060940.1" = "16",
  "NC_060941.1" = "17",
  "NC_060942.1" = "18",
  "NC_060943.1" = "19",
  "NC_060944.1" = "20",
  "NC_060945.1" = "21",
  "NC_060946.1" = "22",
  "NC_060947.1" = "X",
  "NC_060948.1" = "Y"
)



# cds per gene
cds_by_gene <- cdsBy(T2T_txdb, by = "gene")
cds_gr <- cds_by_gene[names(cds_by_gene) %in% removed_genes_cds_gene_list]

# only filter canonical cds (not predicted XM or else)
cds_gr <- endoapply(cds_gr, function(gr) {
  gr[str_detect(gr$cds_name, "NP")]
})

# rename seqlevels
seqlevels(cds_gr) <- refseq_to_chr[seqlevels(cds_gr)]

# drop anything not in the genome
cds_gr <- keepSeqlevels(
  cds_gr,
  seqlevels(BSgenome.Hsapiens.NCBI.T2TCHM13v2.0),
  pruning.mode = "coarse"
)

```

## obtain possible protospacer sites from CDS data

```{r}
# flatten cds granges (that contain ~500 genes), with identity attached
cds_flat <- unlist(cds_gr, use.names = FALSE)
cds_flat

cds_flat$gene_id <- rep(
  names(cds_gr),
  elementNROWS(cds_gr)
)

# put the flat data as a dataframe to join later
df_cds_flat <- as.data.frame(cds_flat)


# reduced cds within each gene, the function of reduced cds is to filtered out the duplicate from canonical and isoform transcript that has the same start-end exonic region
cds_reduced <- reduce(
  split(cds_flat, cds_flat$gene_id)
)

cds_reduced <- unlist(cds_reduced, use.names = FALSE)

# find all possible spacers site from the cds of ~500 removed genes
data(SpCas9, package="crisprBase")
cds_find_spacer <- findSpacers(cds_reduced, bsgenome = BSgenome.Hsapiens.NCBI.T2TCHM13v2.0, crisprNuclease = SpCas9)

hits <- findOverlaps(cds_find_spacer, cds_flat, type = "within")
hits

df_all_spacers <- data.frame(
  gene_id = cds_flat$gene_id[subjectHits(hits)],
  cds_id = cds_flat$cds_id[subjectHits(hits)],
  spacer_chr = as.character(seqnames(cds_find_spacer)[queryHits(hits)]),
  spacer_strand = as.character(strand(cds_find_spacer)[queryHits(hits)]),
  pam = cds_find_spacer$pam[queryHits(hits)],
  pam_site = cds_find_spacer$pam_site[queryHits(hits)],
  cut_site = cds_find_spacer$cut_site[queryHits(hits)],
  spacer_seq = cds_find_spacer$protospacer[queryHits(hits)],
  stringsAsFactors = FALSE
)

all_spacers_exon_data <- df_all_spacers %>%
  left_join(df_cds_flat, join_by(gene_id, cds_id)) %>%
  dplyr::select(-c(seqnames, strand)) %>%
  dplyr::rename("start_cds" = "start",
                "end_cds" = "end",
                "chromosome" = "spacer_chr",
                "strand" = "spacer_strand") %>%
  relocate(cds_name, .before = start_cds)

write_csv(all_spacers_exon_data, "../data/removed_genes_survey/all_spacers_exon_data_499_removed_genes.csv")

```

## Does each genes have limited number of protospacer site ?

```{r}

count_protospacer_per_gene <- all_spacers_exon_data %>%
  dplyr::select(gene_id, pam_site) %>%
  distinct() %>%
  group_by(gene_id) %>%
  summarise(num_pam_site = n()) %>%
  ungroup()

count_protospacer_per_gene$count_bin <- cut(count_protospacer_per_gene$num_pam_site,
                                            breaks = c(0, 10, 20, 30, 40, 50, 60, 70, 80, 90, 100, Inf),
                                            labels = c(
                                              "0–10",
                                              "10–20",
                                              "20–30",
                                              "30–40",
                                              "40–50",
                                              "50-60",
                                              "60–70",
                                              "70–80",
                                              "80–90",
                                              "90-100",
                                              ">100"),
                                            right = TRUE,
                                            include.lowest = TRUE
                                          )

df_barplot <- count_protospacer_per_gene %>%
  dplyr::select(count_bin) %>%
  count(count_bin) %>%
  dplyr::rename("num_genes" = "n")


ggplot(df_barplot, aes(x = count_bin, y = num_genes)) +
    geom_col(fill = "#1E88E5") +
    theme_minimal() +
    labs(x = "Number of NGG PAM-site at CDS", y = "Number of genes") +
    geom_text(aes(label = paste("n=", num_genes, sep="")),
              nudge_y = 5,
              nudge_x = 0,
              size = 5) +
    theme(axis.text.x = element_text(size = 12, vjust = 0.7, angle = 45, colour = 'black'),
          axis.text.y = element_text(size = 12, colour = "black"),
          axis.title.x = element_text(size = 12, vjust = 0.5, margin = margin(r = 20), colour = "black"),
          axis.title.y = element_text(size = 12, vjust = 0.5, margin = margin(r = 20), colour = "black"),
          axis.ticks.y = element_line(size = 1),
          axis.ticks.length = unit(0.15, "cm"),
          axis.line.y.right = element_blank(),
          axis.ticks.y.right = element_blank(), 
          axis.text.y.right = element_blank(), 
          panel.grid.major = element_blank(),
          panel.grid.minor = element_blank(),
          axis.line.x = element_line(size = 0.5, colour = "black"),
          axis.line.y = element_line(size = 0.5, colour = "black"),
          legend.position = "none",
          plot.margin = margin(1, 1, 1, 1),
          legend.spacing.x = unit(1, 'cm'))


```

## We asked whether the exon size is small as it may contributes to the off-target activity of sgRNAs ?

```{r}
# Another code to get the summary of exon length from all removed genes

T2T_txdb <- loadDb("../data/T2T_data/TxDb_T2T-CHM13v2.sqlite")
columns(T2T_txdb)

T2T_txdb

refseq_to_chr <- c(
  "NC_060925.1" = "1",
  "NC_060926.1" = "2",
  "NC_060927.1" = "3",
  "NC_060928.1" = "4",
  "NC_060929.1" = "5",
  "NC_060930.1" = "6",
  "NC_060931.1" = "7",
  "NC_060932.1" = "8",
  "NC_060933.1" = "9",
  "NC_060934.1" = "10",
  "NC_060935.1" = "11",
  "NC_060936.1" = "12",
  "NC_060937.1" = "13",
  "NC_060938.1" = "14",
  "NC_060939.1" = "15",
  "NC_060940.1" = "16",
  "NC_060941.1" = "17",
  "NC_060942.1" = "18",
  "NC_060943.1" = "19",
  "NC_060944.1" = "20",
  "NC_060945.1" = "21",
  "NC_060946.1" = "22",
  "NC_060947.1" = "X",
  "NC_060948.1" = "Y"
)

calculate_exon_length_and_gene_span <- function(txdb_files, list_of_genes) {
  
  # ---------------------------------------------- Gene span data -------------------------------------------------------

  # Get gene ranges
  genes_gr <- genes(T2T_txdb)
  
  # Subset the genes data only for the removed genes across libraries (520 genes)
  # or select all the rest of the genes (18776 genes)
  genes_gr_subset <- genes_gr[mcols(genes_gr)$gene_id %in% list_of_genes]
  
  # Compute span per gene
  gene_span_per_gene <- width(genes_gr_subset)
  
  # Convert to dataframe
  gene_span_df <- data.frame(
    gene = mcols(genes_gr_subset)$gene_id,
    total_gene_length_bp = gene_span_per_gene
  )
  
  # "KRTAP5-1"  "KRTAP5-10" "PLA2G10"   "TAS2R43"   "UPK3BL1" is not mapped to the T2T-CHM13 gene ranges data, annotation issues
    
  # ---------------------------------------------- Exon data per gene -------------------------------------------------------

  # Obtain exons per gene
  exons_by_gene <- exonsBy(T2T_txdb, by = "gene")
  
  # Subset exons by genes removed across libraries (520 genes)
  exons_gr <- exons_by_gene[names(exons_by_gene) %in% list_of_genes]
  
  # Reduce overlapping exons within each gene
  reduced_exons <- reduce(exons_gr)
  
  # Compute non-redundant exon length per gene and gene span
  total_exon_length_per_gene <- sum(width(reduced_exons))
  
  # Convert the total exon length per gene to dataframe
  total_exon_length_df <- data.frame(
    gene = names(total_exon_length_per_gene),
    total_exon_length_bp = total_exon_length_per_gene
  )
  
  rownames(total_exon_length_df) <- NULL
  
  # ------------------------------------------ Left join exon data and gene span ---------------------------------------------
  
  total_exon_length_df <- total_exon_length_df %>%
    left_join(gene_span_df, join_by("gene"))
  
  return(total_exon_length_df)
  
}



calculate_cds_length_and_gene_span <- function(txdb_files, list_of_genes) {
  
  # ---------------------------------------------- Gene span data -------------------------------------------------------

  # Get gene ranges
  genes_gr <- genes(T2T_txdb)
  
  # Subset the genes data only for the removed genes across libraries (520 genes)
  # or select all the rest of the genes (18776 genes)
  genes_gr_subset <- genes_gr[mcols(genes_gr)$gene_id %in% list_of_genes]
  
  # Compute span per gene
  gene_span_per_gene <- width(genes_gr_subset)
  
  # Convert to dataframe
  gene_span_df <- data.frame(
    gene = mcols(genes_gr_subset)$gene_id,
    total_gene_length_bp = gene_span_per_gene
  )
  
  # "KRTAP5-1"  "KRTAP5-10" "PLA2G10"   "TAS2R43"   "UPK3BL1" is not mapped to the T2T-CHM13 gene ranges data, annotation issues
    
  # ---------------------------------------------- Exon data per gene -------------------------------------------------------

  # Obtain exons per gene
  cds_by_gene <- cdsBy(T2T_txdb, by = "gene")
  
  # Subset exons by genes removed across libraries (520 genes)
  cds_gr <- cds_by_gene[names(cds_by_gene) %in% list_of_genes]
  
  # Reduce overlapping CDS regions within each gene
  reduced_cds <- reduce(cds_gr)
  
  # Compute non-redundant exon length per gene and gene span
  total_cds_length_per_gene <- sum(width(reduced_cds))
  
  # Convert the total exon length per gene to dataframe
  total_cds_length_df <- data.frame(
    gene = names(total_cds_length_per_gene),
    total_cds_length_bp = total_cds_length_per_gene
  )
  
  rownames(total_cds_length_df) <- NULL
  
  # ------------------------------------------ Left join exon data and gene span ---------------------------------------------
  
  total_cds_length_df <- total_cds_length_df %>%
    left_join(gene_span_df, join_by("gene"))
  
  return(total_cds_length_df)
  
}

# 515 genes out of 520 genes mapped into the T2T_txdb file
removed_genes_exons <- calculate_exon_length_and_gene_span(T2T_txdb, removed_genes_all_libraries)
removed_genes_cds <- calculate_cds_length_and_gene_span(T2T_txdb, removed_genes_all_libraries)
removed_genes_cds_gene_list <- removed_genes_cds %>% pull(gene) %>% unique()


# 18,632 genes out of 18,776 genes mapped into the T2T_txdb file
control_genes_exons <- calculate_exon_length_and_gene_span(T2T_txdb, all_genes_but_not_removed)
control_genes_cds <- calculate_cds_length_and_gene_span(T2T_txdb, all_genes_but_not_removed)
control_genes_cds_gene_list <- control_genes_cds %>% pull(gene) %>% unique()

```

from 520 removed genes across libraries, there are 5 genes that were not mapped to T2T-CHM13 data: KRTAP5-1, KRTAP5-10, PLA2G10, TAS2R43, UPK3BL1 , Most likely due to annotation issues or symbol naming

### Visualization 1: Barplot of exon length of removed genes or control genes

```{r}

# Select either removed genes or control genes

# ------------------------------------------ Bin exon length by 2000 bp ---------------------------------------------

control_genes_cds$cds_length_bin <- cut(control_genes_cds$total_cds_length_bp,
                                            breaks = c(0, 2000, 4000, 6000, 8000, 10000, Inf),
                                            labels = c(
                                              "0–2000",
                                              "2000–4000",
                                              "4000–6000",
                                              "6000–8000",
                                              "8000–10000",
                                              ">10000"),
                                            right = TRUE,
                                            include.lowest = TRUE
                                          )

# --------------------------------------- Plot the total exon length bin ----------------------------------------------------

df_barplot <- control_genes_cds %>%
  dplyr::select(cds_length_bin) %>%
  count(cds_length_bin) %>%
  dplyr::rename("num_genes" = "n") %>%
  mutate(percentage = round((num_genes/sum(num_genes)*100), 2))


barplot_cds_length <- ggplot(df_barplot, aes(x = cds_length_bin, y = percentage)) +
                          geom_col(fill = "#1E88E5") +
                          theme_minimal() +
                          labs(x = "Total CDS length (bp)", y = "Percentage of genes") +
                          geom_text(aes(label = paste0(percentage, " %")),
                                    nudge_y = 2,
                                    nudge_x = 0,
                                    size = 5) +
                          theme(axis.text.x = element_text(size = 15, vjust = 0.7, angle = 0, colour = 'black'),
                                axis.text.y = element_text(size = 15, colour = "black"),
                                axis.title.x = element_text(size = 15, vjust = 0.5, margin = margin(t = 20), colour = "black"),
                                axis.title.y = element_text(size = 15, vjust = 0.5, margin = margin(r = 20), colour = "black"),
                                axis.ticks.x = element_line(size = 1),
                                axis.ticks.y = element_line(size = 1),
                                axis.ticks.length = unit(0.15, "cm"),
                                axis.line.y.right = element_blank(),
                                axis.ticks.y.right = element_blank(), 
                                axis.text.y.right = element_blank(), 
                                panel.grid.major = element_blank(),
                                panel.grid.minor = element_blank(),
                                axis.line.x = element_line(size = 0.5, colour = "black"),
                                axis.line.y = element_line(size = 0.5, colour = "black"),
                                legend.position = "none",
                                plot.margin = margin(1, 1, 1, 1),
                                legend.spacing.x = unit(1, 'cm'))

ggsave(
    path = './',
    width = 9,
    height = 5,
    dpi = 1000,
    plot = barplot_cds_length,
    filename = "../figure/01_fig_analysis_off-target-sgrna/cds_length_18586_control_genes.png")

```

### Visualization 2: Boxplot of exon length of removed genes or control genes

```{r}

removed_genes_cds <- removed_genes_cds %>% mutate(gene_status = "removed genes",
                                          log_transform_cds_length = log10(total_cds_length_bp))
control_genes_cds <- control_genes_cds %>% mutate(gene_status = "control genes",
                                          log_transform_cds_length = log10(total_cds_length_bp))

combined_df_cds_length <- rbind(removed_genes_cds, control_genes_cds) %>% mutate(gene_status = factor(gene_status, levels = c("control genes", "removed genes")))

cds_length_boxplot <- ggboxplot(combined_df_cds_length, 
                         x = "gene_status", 
                         y = "log_transform_cds_length",
                         fill = "gene_status",
                         outliers = FALSE) +
                    stat_compare_means(comparisons = list(c("removed genes", "control genes")),
                                       method = "wilcox.test",
                                       method.args = list(alternative = "two.sided"),
                                       size = 2.5,
                                       step.increase = 0.05,
                                       tip.length = 0.01,
                                       label = "p.signif",
                                       label.y = 4.5) +
  scale_fill_manual(
      name = "",
      values = c("control genes" = "powderblue",
                 "removed genes" = "mistyrose"),
      labels = c("Control Genes", "Removed Genes")) +
  labs(title = "", 
       x = "",
       y = "log10(Total CDS length in all transcripts [bp])") +
    theme(plot.margin = margin(10, 10, 25, 10),
          strip.text.x = element_text(size = 14),
          axis.text.x = element_blank(),
          axis.text.y = element_text(size = 14, colour = "black"),
          axis.ticks.y = element_line(size = 1.5),
          axis.ticks.x = element_blank(), 
          axis.ticks.length = unit(0.1, "cm"),
          axis.title.y = element_text(size = 14, vjust = 0.5, margin = margin(r = 20)),
          panel.grid.major = element_line(size = 0.5),
          panel.grid.minor = element_blank(),
          axis.line.x = element_line(size = 0.5),
          axis.line.y = element_line(size = 0.5), 
          line = element_line((size = 2), colour = 'black'),
          legend.position = "right",
          legend.text = element_text(size = 12)) +
  ylim(0, 5)

# get stats

b <- ggplot_build(cds_length_boxplot)
stats <- b$data[[1]]

label_df <- stats %>%
  transmute(
    x = x,
    y = ymax + 0.2,          # add padding
    label = round(ymax, 2)   # or any custom label
  )

count <- combined_df_cds_length %>%
  distinct() %>%
  count(gene_status)

count <- count %>% bind_cols(stats %>% dplyr::select(x, ymin))

cds_length_boxplot <-  cds_length_boxplot +
                              geom_text(
                                data = count,
                                aes(x = x, y = ymin - 0.5, label = paste0("n= ", n)),
                                angle = 45,
                                vjust = 0,
                                hjust = 0.5,
                                size = 3.5
                              )
cds_length_boxplot

ggsave(
    path = '../figure/01_fig_analysis_off-target-sgrna/',
    width = 5,
    height = 5,
    dpi = 1000,
    plot = cds_length_boxplot,
    filename = "log10_cds_length_removed_genes.png")

```





