---
title: "Analysis of commonly removed genes across libraries"
author: "Author: Stefanus Bernard"
date: "Date: `r format(Sys.time(), '%d %B %Y')`"
output: 
  html_document :
    theme : cerulean
    toc : true
    toc_float : true
    number_section : true
---

```{r setup, include=FALSE}
library(tidyverse)
library(stringr)
library(UpSetR)
library(GenomicFeatures)
library(HGNChelper)
library(BSgenome.Hsapiens.NCBI.T2TCHM13v2.0)
library(GenomeInfoDbData)
library(txdbmaker)
library(crisprVerse)

source("./00_CRISPR-KO-library-survey-functions.R")
```

## Import data containing removed genes from each library

```{r}

removed_genes <- read_csv("../data/biogrid_crispr_ko_data/important_data/removed_genes_all_library.csv")

removed_gene_list_combined <- list(
  Avana = na.omit(removed_genes[["avana_removed_genes_all"]]),
  Brunello = na.omit(removed_genes[["brunello_removed_genes_all"]]),
  Toronto = na.omit(removed_genes[["toronto_removed_genes_all"]]),
  Yusa = na.omit(removed_genes[["yusa_removed_genes_all"]]),
  Jacquere = na.omit(removed_genes[["jacquere_removed_genes_all"]])
)

# List of 520 protein-coding genes are removed across CRISPR-KO libraries, what are these genes?
removed_genes_all_libraries <- unique(Reduce(intersect, removed_gene_list_combined))

```

## UpSet plot for commonality of removed genes between 5 libraries

```{r}

png("../figure/01_fig_analysis_off-target-sgrna/removed_genes_all_libraries.png", width = 12000, height = 7500, res = 1000)

par(mar = c(5, 20, 4, 2))
upset(fromList(removed_gene_list_combined),
      order.by = "degree",
      main.bar.color = "#4B0092",
      sets.bar.color = "#1AFF1A",
      text.scale = 1.5)

dev.off()

```

## Import more relevant data

```{r}

# Import HGNC dataset

hgnc <- read_tsv(url("https://storage.googleapis.com/public-download-files/hgnc/tsv/tsv/hgnc_complete_set.txt")) %>%
  filter(symbol %in% removed_genes_all_libraries) %>%
  dplyr::select(symbol, name, locus_group, ensembl_gene_id) %>%
  dplyr::rename("gene" = "symbol")

# Import paralog data
ensembl_paralog <- ensembl_paralog_gene_pairs()

paralog_pairs <- ensembl_paralog[[1]]
paralog_individual <- ensembl_paralog[[2]]

# Import the data of sgRNA library report (looking on how many guides actually multi-targeting, etc)

terms <- c("CONTROL", "Control", "control", "INTRON", "Intron", "intron", "LacZ", "luciferase", "NO_SITE", "ONE_INTERGENIC_SITE")
# lacZ and luciferase are the control in Toronto V3 library

terms_for_filtering <- paste(terms, collapse = "|")

avana <- 'avana_library'
brunello <- 'broadgpp-brunello-library-contents'
toronto <- 'tkov3_guide_sequence'
yusa <- 'yusa_hcrispr_ko_grnas'
jacquere <- 'Jacquere_PerGuideAnnotations_Quota4'


avana_report <- import_and_process_report(paste0('~/CRISPR-KO-GuideRefine/output_cleaning/T2T-CHM13/', avana, '_full_report.xlsx')) %>% filter(gene_hgnc %in% removed_genes_all_libraries) %>% mutate(library = "avana")

brunello_report <- import_and_process_report(paste0('~/CRISPR-KO-GuideRefine/output_cleaning/T2T-CHM13/', brunello, '_full_report.xlsx')) %>% filter(gene_hgnc %in% removed_genes_all_libraries) %>% mutate(library = "brunello")

toronto_report <- import_and_process_report(paste0('~/CRISPR-KO-GuideRefine/output_cleaning/T2T-CHM13/', toronto, '_full_report.xlsx')) %>% filter(gene_hgnc %in% removed_genes_all_libraries) %>% mutate(library = "toronto")

yusa_report <- import_and_process_report(paste0('~/CRISPR-KO-GuideRefine/output_cleaning/T2T-CHM13/', yusa, '_full_report.xlsx')) %>% filter(gene_hgnc %in% removed_genes_all_libraries) %>% mutate(library = "yusa")

jacquere_report <- import_and_process_report(paste0('~/CRISPR-KO-GuideRefine/output_cleaning/T2T-CHM13/', jacquere, '_full_report.xlsx')) %>% filter(gene_hgnc %in% removed_genes_all_libraries) %>% mutate(library = "jacquere")

combined_report <- rbind(avana_report, brunello_report, toronto_report, yusa_report, jacquere_report) %>% arrange(gene_hgnc)

rm(avana_report, brunello_report, toronto_report, yusa_report, jacquere_report)

# Import the alignment classification data

avana_alignment <- read_csv("../data/sgrna_lfc_data/avana_data/avana_sgrna_alignment_classification.csv") %>% update_gene_symbol(., removed_genes_all_libraries) %>% mutate(library = "avana")

brunello_alignment <- read_csv("../data/sgrna_lfc_data/brunello_data/brunello_sgrna_alignment_classification.csv") %>% update_gene_symbol(., removed_genes_all_libraries) %>% mutate(library = "brunello")

toronto_alignment <- read_csv("../data/sgrna_lfc_data/tkov3_data/tkov3_sgrna_alignment_classification.csv") %>% update_gene_symbol(., removed_genes_all_libraries) %>% mutate(library = "toronto")

yusa_alignment <- read_csv("../data/sgrna_lfc_data/yusa_data/yusa_sgrna_alignment_classification.csv") %>% update_gene_symbol(., removed_genes_all_libraries) %>% mutate(library = "yusa")

jacquere_alignment <- read_csv("../data/sgrna_lfc_data/jacquere_data/jacquere_sgrna_alignment_classification.csv") %>% update_gene_symbol(., removed_genes_all_libraries) %>% mutate(library = "jacquere")

combined_alignment <- rbind(avana_alignment, brunello_alignment, toronto_alignment, yusa_alignment, jacquere_alignment) %>% 
  arrange(gene_hgnc) %>%
  # remove perfect guides because we dont need it
  filter(!alignment == "perfect")

rm(avana_alignment, brunello_alignment, toronto_alignment, yusa_alignment, jacquere_alignment)

```

## There are 520 genes removed across libraries
we asked why there are 520 genes failed/represented by fewer than 3 guides across libraries

## We asked whether the size of genes and limited number of protospacer site contributes to the off-target activity of sgRNAs ?

```{r}
# create txdb sqlite file for T2T-CHM13

T2T_txdb <- txdbmaker::makeTxDbFromGFF(file = "../../GuideRefine/annotation_file/GCF_009914755.1_T2T-CHM13v2.0_genomic.gff.gz",
                            dataSource = "NCBI T2T Assembly GCF_009914755.1_T2T-CHM13v2.0",
                            organism = "Homo sapiens")
saveDb(T2T_txdb, "../data/T2T_data/TxDb_T2T-CHM13v2.sqlite")

```


```{r}
# Load TXdb file

T2T_txdb <- loadDb("../data/T2T_data/TxDb_T2T-CHM13v2.sqlite")
columns(T2T_txdb)

T2T_txdb

# list of columns you can fetch
#[1] "CDSCHROM"   "CDSEND"     "CDSID"      "CDSNAME"    "CDSPHASE"   "CDSSTART"   "CDSSTRAND"  "EXONCHROM" 
#  [9] "EXONEND"    "EXONID"     "EXONNAME"   "EXONRANK"   "EXONSTART"  "EXONSTRAND" "GENEID"     "TXCHROM"   
# [17] "TXEND"      "TXID"       "TXNAME"     "TXSTART"    "TXSTRAND"   "TXTYPE" 

# Extract gene coordinates
genes_gr <- genes(T2T_txdb)

# Filter for your list
genes_gr <- genes_gr[names(genes_gr) %in% removed_genes_all_libraries]
genes_gr$span_length <- width(genes_gr)
test <- as.data.frame(genes_gr)


# exons per gene
exons_by_gene <- exonsBy(T2T_txdb, by = "gene")
exons_gr <- exons_by_gene[names(exons_by_gene) %in% removed_genes_all_libraries]

exon_lengths <- lapply(exons_gr, width)
exon_lengths_df <- stack(exons_gr)
exon_lengths_df$exon_length <- width(exon_lengths_df)

data_exons_all_removed_genes <- as.data.frame(exon_lengths_df) %>%
  filter(str_detect(exon_name, "NM")) %>%
  mutate(exon_nm_accession = str_extract(exon_name, "^[^.]+"))


# dataframe of total exon length in basepairs (of only 500 genes)
total_exon_length <- data_exons_all_removed_genes %>%
  dplyr::select(name, exon_nm_accession, exon_length) %>%
  arrange(name) %>%
  group_by(name, exon_nm_accession) %>%
  summarise(total_exon_length = sum(exon_length))

```

# obtain possible protospacer sites from exonic data

```{r}

unlist_exons_gr <- unlist(exons_gr, use.names = FALSE)

data(SpCas9, package="crisprBase")
spacer_site <- findSpacers(unlist_exons_gr, bsgenome = BSgenome.Hsapiens.NCBI.T2TCHM13v2.0, crisprNuclease = SpCas9)

```














