---
title: "Backup Code of multi_target_sgrna_analysis"
output: github_document
---
# A venn diagram function to compare the similarity between three libraries

```{r}

venn_diagram <- function(set1, set2, name_set1, name_set2, color1, color2, name_venn) {
  only_set1 <- length(setdiff(set1, set2))
  only_set2 <- length(setdiff(set2, set1))
  both_sets  <- length(intersect(set1, set2))

  totalset1 <- only_set1 + both_sets
  totalset2 <- only_set2 + both_sets

  pct_set1 <- round((only_set1 / totalset1) * 100, 2)
  pct_set2 <- round((only_set2 / totalset2) * 100, 2)
  pct_overlap_set1 <- round((both_sets / totalset1) * 100, 2)
  pct_overlap_set2 <- round((both_sets / totalset2) * 100, 2)

  sgRNA_similarity <- venn.diagram(
        x = list(set1, set2),
        category.names = c(name_set1, name_set2),
        filename = NULL,  
        fill = c(color1, color2),
        alpha = 0.5,
        cex = 0,  # Set text size to 0 to hide numbers
        cat.cex = 0.75,  # Set category label size
        cat.pos = c(-2, 2),  # Adjust category label positions
        cat.dist = c(0.05, 0.05),  # Adjust category label distances
        lwd = 2
  )

  png(filename = name_venn, width = 1250, height = 1250, res = 300)
  grid.newpage()
  grid.draw(sgRNA_similarity)

  grid.text(label = paste0(only_set2, '\n', pct_set1, '%'), x = 0.25, y = 0.50, gp = gpar(fontsize = 10))
  grid.text(label = paste0(only_set1, '\n', pct_set2, '%'), x = 0.75, y = 0.50, gp = gpar(fontsize = 10))
  grid.text(label = paste0(pct_overlap_set1, '%', '\nor\n', pct_overlap_set2, '%'), x = 2.5, y = 0.50, gp = gpar(fontsize = 7))

  dev.off()
}


```

# compare the similarity of all sequences between three libraries

```{r}
venn_diagram(brunello$spacer, toronto$spacer, "Brunello \n 77K sgRNAs", "Toronto V3 \n 71K sgRNAs", 'powderblue', 'mistyrose', 'VennDiagram_sgRNAs_brunello_toronto.png')
venn_diagram(brunello$spacer, cellecta$spacer, "Brunello \n 77K sgRNAs", "Cellecta \n 150K sgRNAs", 'powderblue', 'lightgreen', 'VennDiagram_sgRNAs_brunello_cellecta.png') 
venn_diagram(toronto$spacer, cellecta$spacer, "Toronto V3 \n 77K sgRNAs", "Cellecta \n 150K sgRNAs", 'mistyrose', 'lightgreen', 'VennDiagram_sgRNAs_toronto_cellecta.png') 
```

# off-target guides

```{r}
# get list off target guides
list_off_target_guides <- function(alignment_data) {
  # start with multi-target sgRNA
  multi_target_library <- multi_target_sgrna(alignment_data)
  
  # removal of multi-target sgRNA from the library
  alignment_data_2 <- alignment_data %>%
    filter(!sgRNA %in% multi_target_library$list_guides)
  
  # continue with single mismatch sgRNA
  single_mismatch_library <- single_mismatch_sgrna(alignment_data_2)
  
  # removal of single mismatch sgRNA from the library
  alignment_data_3 <- alignment_data_2 %>%
    filter(!sgRNA %in% single_mismatch_library$list_guides)
  
  # continue with pam double mismatch sgRNA
  pam_double_mismatch_library <- pam_distal_double_mismatch(alignment_data_3)
  
  list_off_target_guides <- unique(union(multi_target_library$list_guides, unique(union(single_mismatch_library$list_guides, pam_double_mismatch_library$list_guides))))
  
  return(list(
    multi_target_guides = unique(multi_target_library$list_guides),
    single_mismatch_guides = unique(single_mismatch_library$list_guides), 
    pam_double_mismatch_guides = unique(pam_double_mismatch_library$list_guides),
    all_off_target_guides = list_off_target_guides))
}
```

# off target annotation

```{r}
# off target annotation (can be multi-target, single-mismatch, or pam-distal double mismatches) used to annotate whether the sgRNA target random genomic loci, protein-coding genes or intended target gene

off_target_annotation <- function(alignment_data, num_mismatch, library, library_name){
  gene_df <- annotate_off_target(alignment_data)
  gene_df <- gene_df %>%
    dplyr::rename('actual_chr' = 'chr',
                  'actual_gene' = 'gene')

  alignment_data <- add_cut_pos_pam_pos(alignment_data)
  alignment_data <- alignment_data %>%
    select(unique_aln_id, sgRNA, spacer, protospacer, gene, chr, n_mismatches, cut_pos) %>%
    dplyr::rename('sgrna' = 'sgRNA')
  
  # one version of alignment data for paralog analysis (you will need the actual gene to detect which multi-target sgRNA targeting only two protein-coding genes - one is the actual gene, the other one is the paralog pairs)
  alignment_data_gene_df_1 <- alignment_data %>%
    left_join(gene_df, by = join_by(unique_aln_id, sgrna, spacer, protospacer, cut_pos)) %>%
    select(unique_aln_id, sgrna, spacer, protospacer, gene, chr, n_mismatches, cut_pos, actual_chr, actual_gene) %>%
    filter(n_mismatches == num_mismatch) %>%
    mutate(target = case_when(gene == actual_gene ~ 'actual gene',
                              gene != actual_gene ~ 'protein-coding genes',
                              is.na(actual_gene) ~ 'other genomic loci'))
  
  # 2nd version of alignment data for frequency table and histogram plot
  alignment_data_gene_df_2 <- alignment_data_gene_df_1 %>%
    filter(target != 'actual gene')
  
  summary_target_df <- alignment_data_gene_df_2 %>%
    select(sgrna, target) %>%
    group_by(sgrna) %>%
    mutate(target = ifelse(n() > 1, 'other genomic loci or protein-coding genes', target)) %>%
    ungroup() %>%
    distinct()
  
  calculate_freq_table <- summary_target_df %>%
    select(target) %>%
    count(target) %>%
    dplyr::rename('count_sgrna' = 'n') %>%
    arrange(desc(count_sgrna)) %>%
    mutate(percent = round(count_sgrna/length(library$sgRNA)*100 ,2),
           library = library_name)

  return(list(
    aln_data_df_1 = alignment_data_gene_df_1,
    aln_data_df_2 = alignment_data_gene_df_2,
    summary_df = summary_target_df, 
    freq_table = calculate_freq_table))
}
```


# Count on-target alignments

```{r}
count_on_target_alignments <- library_alignment %>%
  count(sgRNA, protospacer) %>%
  mutate(n = if_else(is.na(protospacer), 0, n)) %>%
  dplyr::rename("number_of_alignments" = 'n') %>%
  select(-protospacer) %>%
  group_by(sgRNA) %>%
  summarise(number_of_alignments = sum(number_of_alignments), .groups = "drop")

```




## Is the proportion of sgRNAs targeting paralog pairs higher among multi-target sgRNAs compared to non-multi-target ones? (actual question to the Fisher exact test we performed)


```{r}

fisher_multi_target_paralog_genes <- function(off_target_paralog_genes, off_target_not_paralog_genes, normal_paralog_genes, normal_not_paralog_genes, sgrna_library) {

  print(paste('Paralog genes targeted by multi-target sgRNAs:', length(off_target_paralog_genes)))
  print(paste('Non paralog genes targeted by multi-target sgRNAs:', length(off_target_not_paralog_genes)))
  print(paste('Paralog genes not targeted by multi-target sgRNAs:', length(normal_paralog_genes)))
  print(paste('Non paralog genes not targeted by multi-target sgRNAs:', length(normal_not_paralog_genes)))

  # Create the contingency table
  contingency_table <- matrix(
    c(
      length(off_target_paralog_genes),  # Multi-target sgRNAs targeting paralog genes
      length(off_target_not_paralog_genes),  # Multi-target sgRNAs not targeting paralog genes
      length(normal_paralog_genes),  # Non-multi-target sgRNAs targeting paralog genes
      length(normal_not_paralog_genes)  # Non-multi-target sgRNAs not targeting paralog genes
    ),
    nrow = 2
  )

  print(contingency_table)

  # Check if the total matches the unique sgRNAs in alignment_data
  total_contingency_table <- sum(contingency_table)
  total_unique_gene <- length(unique(sgrna_library$gene))

  print(paste('total sgRNA count in contingency table:', total_contingency_table))
  print(paste('total original number of gene:', total_unique_gene))

  if (total_contingency_table == total_unique_gene) {
    print("The sum of the contingency table matches the total unique gene in sgRNA library.")
  } else {
    print("Mismatch detected between the contingency table sum and the total unique gene in sgRNA library")
  }

  # Perform Fisher's test
  fischer_library_multi_target <- fisher.test(contingency_table)
  print(fischer_library_multi_target)

}

gene_paralog_target <- function(sgrna_library, gene_paralog_multi_target, gene_not_paralog_multi_target, barbara_paralog_dataset) {
  
  # Load paralog gene pairs
  barbara_paralog <- read_csv(barbara_paralog_dataset)
  
  # Extract the vector of paralog genes from column A2
  paralog_genes <- pull(barbara_paralog, A2)
  
  # Genes that are paralogs but not multi-target or not in the not-paralog list
  gene_paralog_normal_target <- sgrna_library %>%
    filter(
      gene %in% paralog_genes,
      !gene %in% gene_paralog_multi_target,
      !gene %in% gene_not_paralog_multi_target
    )
  
  # Genes that are NOT paralogs, and also not multi-target or in not-paralog list
  gene_not_paralog_normal_target <- sgrna_library %>%
    filter(
      !gene %in% paralog_genes,
      !gene %in% gene_paralog_multi_target,
      !gene %in% gene_not_paralog_multi_target
    )
  
  return(list(
    paralog_normal_target = unique(gene_paralog_normal_target$gene),
    not_paralog_normal_target = unique(gene_not_paralog_normal_target$gene)
  ))
}


# Function to identify which sgRNAs among all libraries that are not targeting paralog pairs

find_not_multi_target_paralog <- function(sgrna_library, alignment_gene_df, paralog_dataset) {
  
  # read paralog dataset
  paralog <- read_csv(paralog_dataset)
  
  # non multi-target sgRNA in the library
  sgrna_library_not_multi_target <- sgrna_library %>%
    filter(!sgRNA %in% alignment_gene_df$sgrna)
  
  # normal sgRNA targeting paralog
  sgrna_library_paralog <- sgrna_library_not_multi_target %>%
    filter(gene %in% unique(c(paralog$A2, paralog$A1)))
  
  # normal sgRNA not targeting paralog
  sgrna_library_not_paralog <- setdiff(unique(sgrna_library_not_multi_target$sgRNA), unique(sgrna_library_paralog$sgRNA))
  
  return (list(normal_sgrna_paralog = unique(sgrna_library_paralog$sgRNA),
               normal_sgrna_not_paralog = sgrna_library_not_paralog))
  
}


brunello_paralog_normal <- gene_paralog_target(brunello, brunello_gene_paralog_multi_target, brunello_gene_not_paralog_multi_target, '../barbara_36K_paralog_pairs.csv')

brunello_gene_fisher <- fisher_multi_target_paralog_genes(brunello_gene_paralog_multi_target, 
                                                          brunello_gene_not_paralog_multi_target,
                                                          brunello_paralog_normal$paralog_normal_target,
                                                          brunello_paralog_normal$not_paralog_normal_target,
                                                          brunello)


```


### What is percentage of sgRNAs that target multiple genomic sites with mismatches?
### What is percentage of sgRNAs that target multiple protein-coding genes with mismatches?

```{r}

avana_stratify_single_mismatch <- off_target_alignment_bin(avana_classification[[1]], c("single mismatch"))
visualize_off_target_alignment(avana_stratify_single_mismatch[[2]], "Number of single mismatch off-target alignments")

brunello_stratify_single_mismatch <- off_target_alignment_bin(brunello_classification[[1]], c("single mismatch"))
visualize_off_target_alignment(brunello_stratify_single_mismatch[[2]], "Number of single mismatch off-target alignments")

toronto_stratify_single_mismatch <- off_target_alignment_bin(toronto_classification[[1]], c("single mismatch"))
visualize_off_target_alignment(toronto_stratify_single_mismatch[[2]], "Number of single mismatch off-target alignments")

yusa_stratify_single_mismatch <- off_target_alignment_bin(yusa_classification[[1]], c("single mismatch"))
visualize_off_target_alignment(yusa_stratify_single_mismatch[[2]], "Number of single mismatch off-target alignments")
  

```


```{r}
list_guides <- avana_stratify_single_mismatch[[1]] %>% filter(alignment_bin %in% c(1,2,3,4,5,6,7,8,"> 8")) %>% pull(sgRNA)
avana_single_mismatch_coding_genes <- off_target_annotation(avana_alignment_meyers, 
                                                            avana_library_meyers, 
                                                            list_guides,
                                                            1,
                                                            "avana")

list_guides <- brunello_stratify_single_mismatch[[1]] %>% filter(alignment_bin %in% c(1,2,3,4,5,6,7,8,"> 8")) %>% pull(sgRNA)
brunello_single_mismatch_coding_genes <- off_target_annotation(brunello_alignment, 
                                                            brunello_library, 
                                                            list_guides,
                                                            1,
                                                            "brunello")

list_guides <- toronto_stratify_single_mismatch[[1]] %>% filter(alignment_bin %in% c(1,2,3,4,5,6,7,8,"> 8")) %>% pull(sgRNA)
toronto_single_mismatch_coding_genes <- off_target_annotation(toronto_alignment, 
                                                            toronto_library, 
                                                            list_guides,
                                                            1,
                                                            "toronto")

list_guides <- yusa_stratify_single_mismatch[[1]] %>% filter(alignment_bin %in% c(1,2,3,4,5,6,7,8,"> 8")) %>% pull(sgRNA)
yusa_single_mismatch_coding_genes <- off_target_annotation(yusa_alignment, 
                                                            yusa_library, 
                                                            list_guides,
                                                            1,
                                                            "yusa")

table_single_mismatch_target <- rbind(brunello_single_mismatch_coding_genes$freq_table,
                            toronto_single_mismatch_coding_genes$freq_table,
                            yusa_single_mismatch_coding_genes$freq_table,
                            avana_single_mismatch_coding_genes$freq_table)
table_single_mismatch_target <- table_single_mismatch_target %>%
  arrange(target)

# Grouped bar chart
grouped_barplot <- ggplot(table_single_mismatch_target, aes(x = factor(library, levels = c('brunello', 'toronto', 'yusa', 'avana')), y = percent_sgrna_among_library, fill = target)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.9)) +
  theme_minimal() +
  labs(title = "", x = "sgRNA libraries", y = "Single mismatch off-target sgRNAs (%)", fill = "Target") +
  scale_x_discrete(labels = c('brunello' = 'Brunello',
                'yusa' = 'Yusa \nProject Score',
                'toronto' = 'Toronto \nKnockout V3',
                'avana' = 'Avana')) +
  scale_fill_manual(labels = c("other genomic loci" = "Other genomic loci",
                               "other protein-coding genes" = "Protein-coding genes"),
                    values = c("other genomic loci" = "#B3589A",
                              "other protein-coding genes" = "#E7F7D5")) +
  theme(axis.text.x = element_text(size = 30, color = "black"),
      axis.text.y = element_text(size = 30, colour = "black"),
      axis.title.x = element_text(size = 0, vjust = -1, margin = margin(r = 10), colour = "black"),
      axis.title.y = element_text(size = 30, margin = margin(r = 10), colour = "black"),
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      axis.line.x = element_line(size = 0.5),
      axis.line.y = element_line(size = 0.5),
      legend.title = element_text(size = 30),
      legend.text = element_text(size = 30)) +
  geom_text(aes(label = paste0(percent_sgrna_among_library, "%")), position = position_dodge(width = 0.8), vjust = -0.5, size = 6.5)  +
  ylim(0, 7)

grouped_barplot

ggsave(
    path = './',
    width = 18,
    height = 15,
    dpi = 300,
    plot = grouped_barplot,
    filename = paste('./plot/percentage_single_mismatch.png', sep = ''))

```

## Double-target guides and their sgRNA log-fold-change
## Find "clean" double-target guides
Clean double-target guides should be
1. Multi-target guides that strictly target 2 protein-coding genes
2. Not having additional single mismatches
3. Not having additional double mismatches

Afterwards we try to find where guide that co-targeting a pair of paralog genes (clean double-target guides) and compare it with at least one single-target guide for each of the two paralogs


```{r}

clean_double_target_guides <- function(list_guides, alignment_data, library_data, library_name, refined_library_dir) {
  
  # fetch double-target guides (without any additional single or double mismatches)
  cleaned_double_target_guides <- alignment_data %>%
        filter(sgRNA %in% list_guides) %>%
        group_by(sgRNA) %>%
        filter(all(n_mismatches == 0)) %>%
        distinct(sgRNA) %>%
        pull(sgRNA)
    
  # annotate the cleaned double-target guides, fetch double-target guides targeting protein-coding genes only
  annotate_cleaned_double_target_guides <- off_target_annotation(
                                            alignment_data,                                       
                                            annotation,
                                            bsgenome,
                                            genome_type_version,
                                            library_data, 
                                            cleaned_double_target_guides,
                                            0,
                                            library_name)
  
  # fetch double-target guides targeting paralogs
  double_target_paralog <- find_multi_target_paralog(annotate_cleaned_double_target_guides[[1]], 
                                                     cleaned_double_target_guides, 
                                                     paralog_pairs, 
                                                     library_data, 
                                                     library_name)
  
  # list of cleaned double-target guides targeting paralogs
  cleaned_double_target_paralog_guides <- double_target_paralog[[4]]
  
  # list of paralog genes affected by double-target guides
  paralog_affected_double_target_guides <- double_target_paralog[[7]]
  
  # import refined library and filter for gene paralog that affected by double-target guides
  # this is to filter the paralog gene that still has "clean" sgRNA upon refinement
  refined_library <- import_library(refined_library_dir) %>%
    filter(gene %in% paralog_affected_double_target_guides)
  
  # dataframe containing the alignment of cleaned double-target guides targeting paralogs
  # this is to filter for cleaned double-target-guides and gene in the refined library
  aln_double_target_paralog_guides <- double_target_paralog[[1]] %>% 
    filter(sgrna %in% cleaned_double_target_paralog_guides & gene %in% refined_library$gene)
    # select(sgrna, spacer, gene, actual_gene, target)
  
  # add another filter to fetch genes that match the criteria from aln_double_target_paralog_guides
  refined_library <- refined_library %>%
    filter(gene %in% unique(aln_double_target_paralog_guides$gene))
  
  combined_df <- bind_rows(refined_library, aln_double_target_paralog_guides)
  combined_df <- combined_df %>% 
    arrange(sgrna, gene)
    
  return(list(
    double_target_paralog = double_target_paralog,
    dbl_target_paralog_guides = cleaned_double_target_paralog_guides, 
    paralog_affected = paralog_affected_double_target_guides,
    df_single_target_guides = refined_library,
    df_co_targeting_guides = aln_double_target_paralog_guides,
    combined_df = combined_df))

}

# fetch perfect alignment at two locations
list_query_guides_paralog <- avana_stratify_multi_target[[1]] %>% filter(alignment_bin == 2) %>% pull(sgRNA)

avana_dbl_target_guides <- clean_double_target_guides(list_query_guides_paralog, 
                               avana_alignment_meyers, 
                               avana_library_meyers, 
                               "avana",
                               "../CRISPR-KO-library-refinement-pipeline/output_cleaning/hg38/avana_library_56K_refined.tsv")


list_query_guides_paralog <- brunello_stratify_multi_target[[1]] %>% filter(alignment_bin == 2) %>% pull(sgRNA)

brunello_dbl_target_guides <- clean_double_target_guides(list_query_guides_paralog, 
                               brunello_alignment, 
                               brunello_library, 
                               "brunello",
                               "../CRISPR-KO-library-refinement-pipeline/output_cleaning/hg38/broadgpp-brunello-library-contents_65K_refined.tsv")


list_query_guides_paralog <- toronto_stratify_multi_target[[1]] %>% filter(alignment_bin == 2) %>% pull(sgRNA)

toronto_dbl_target_guides <- clean_double_target_guides(list_query_guides_paralog, 
                               toronto_alignment, 
                               toronto_library, 
                               "toronto",
                               "../CRISPR-KO-library-refinement-pipeline/output_cleaning/hg38/tkov3_guide_sequence_62K_refined.tsv")


list_query_guides_paralog <- yusa_stratify_multi_target[[1]] %>% filter(alignment_bin == 2) %>% pull(sgRNA)

yusa_dbl_target_guides <- clean_double_target_guides(list_query_guides_paralog, 
                               yusa_alignment, 
                               yusa_library, 
                               "yusa",
                               "../CRISPR-KO-library-refinement-pipeline/output_cleaning/hg38/yusa_hcrispr_ko_grnas_76K_refined.tsv")



```







# Backup code for 02_analysis_paralog_off-target-sgrna
## Percentage of multi-target sgRNAs in each libraries and their target locations

```{r}

table_multi_target <- rbind(brunello_multi_protein_coding_genes$freq_table,
                            toronto_multi_protein_coding_genes$freq_table,
                            yusa_multi_protein_coding_genes$freq_table,
                            avana_multi_protein_coding_genes$freq_table)
table_multi_target <- table_multi_target %>%
  arrange(target)


# Grouped bar chart
grouped_barplot <- ggplot(table_multi_target, aes(x = factor(library, levels = c('brunello', 'toronto', 'yusa', 'avana')), y = percent_sgrna_among_library, fill = target)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.9)) +
  theme_minimal() +
  labs(title = "", x = "sgRNA libraries", y = "Multi-target sgRNAs (%)", fill = "Target") +
  scale_x_discrete(labels = c('brunello' = 'Brunello',
                'yusa' = 'Yusa \nProject Score',
                'toronto' = 'Toronto \nKnockout V3',
                'avana' = 'Avana')) +
  scale_fill_manual(labels = c("other genomic loci" = "Other genomic loci",
                               "other protein-coding genes" = "Protein-coding genes"),
                    values = c("other genomic loci" = "#B3589A",
                              "other protein-coding genes" = "#E7F7D5")) +
  theme(axis.text.x = element_text(size = 30, color = "black"),
      axis.text.y = element_text(size = 30, colour = "black"),
      axis.title.x = element_text(size = 0, vjust = -1, margin = margin(r = 10), colour = "black"),
      axis.title.y = element_text(size = 30, margin = margin(r = 10), colour = "black"),
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      axis.line.x = element_line(size = 0.5),
      axis.line.y = element_line(size = 0.5),
      legend.title = element_text(size = 30),
      legend.text = element_text(size = 30)) +
  geom_text(aes(label = paste0(percent_sgrna_among_library, "%")), position = position_dodge(width = 0.8), vjust = -0.5, size = 6.5)  +
  ylim(0, 7)

grouped_barplot

ggsave(
    path = './',
    width = 18,
    height = 15,
    dpi = 300,
    plot = grouped_barplot,
    filename = paste('../figure/02_fig_analysis_paralog_off-target-sgrna/percentage_multi_target.png', sep = ''))

```

## Percentage of genes affected by multi-target sgRNAs across libraries

```{r}
# barchart for genes

grouped_barplot <- table_multi_target %>%
  filter(target == "other protein-coding genes") %>%
  ggplot(aes(x = factor(library, levels = c('brunello', 'toronto', 'yusa', 'avana')), y = percent_genes)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.8), fill = "#E7F7D5") +
  theme_minimal() +
  labs(title = "", x = "sgRNA libraries", y = "Genes affected by multi-target sgRNAs (%)") +
  scale_x_discrete(labels = c('brunello' = 'Brunello',
                'yusa' = 'Yusa \nProject Score',
                'toronto' = 'Toronto \nKnockout V3',
                'avana' = 'Avana')) +
  theme(axis.text.x = element_text(size = 15),
      axis.text.y = element_text(size = 15),
      axis.title.x = element_text(size = 18, vjust = -1, margin = margin(r = 10)),
      axis.title.y = element_text(size = 18, margin = margin(r = 10)),
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      axis.line.x = element_line(size = 0.5),
      axis.line.y = element_line(size = 0.5),
      legend.title = element_text(size = 15),
      legend.text = element_text(size = 13)) +
  geom_text(aes(label = paste0(percent_genes, "%")), position = position_dodge(width = 0.8), vjust = -0.5, size = 4.5)

grouped_barplot

ggsave(
    path = './',
    width = 7,
    height = 9,
    dpi = 300,
    plot = grouped_barplot,
    filename = paste('./plot/genes_multi_target.png', sep = ''))
```

## Backup code function 00_CRISPR-KO-library-survey-functions.R

```{r}

off_target_annotation <- function(alignment_data, annotation_data, bsgenome, bsgenome_text, library, list_sgrna, num_mismatches, library_name){
  
  alignment_data <- alignment_data %>% filter(sgRNA %in% list_sgrna)
  
  gene_df <- annotate_off_target(alignment_data, annotation_data, bsgenome, bsgenome_text)
  gene_df <- gene_df %>%
    dplyr::rename('actual_chr' = 'chr',
                  'actual_gene' = 'gene')
  
  alignment_data <- add_cut_pos_pam_pos(alignment_data)
  alignment_data <- alignment_data %>%
    select(unique_aln_id, sgRNA, spacer, protospacer, gene, chr, n_mismatches, cut_pos) %>%
    dplyr::rename('sgrna' = 'sgRNA')
  
  # one version of alignment data for paralog analysis (you will need the actual gene to detect which multi-target sgRNA targeting only two protein-coding genes - one is the actual gene, the other one is the paralog pairs)
  alignment_data_gene_df <- alignment_data %>%
    filter(n_mismatches == num_mismatches) %>%
    left_join(gene_df, by = join_by(unique_aln_id, sgrna, spacer, protospacer, cut_pos)) %>%
    select(unique_aln_id, sgrna, spacer, protospacer, gene, chr, n_mismatches, cut_pos, actual_chr, actual_gene) %>%
    mutate(target = case_when(gene == actual_gene ~ 'actual target gene',
                              gene != actual_gene ~ 'other protein-coding genes',
                              is.na(actual_gene) ~ 'other genomic loci')) %>%
    distinct()
  
  summary_target_df <- alignment_data_gene_df %>%
    filter(target != "actual target gene") %>%
    select(sgrna, target) %>%
    group_by(sgrna) %>%
    # first removal of duplicate data to remove duplicate alignment to multiple locations 
    distinct() %>%
    # there may be a case where a single sgRNA can perfectly target both protein-coding gene and genomic loci, this code is used assign those sgRNA those sgRNA as targeting protein-coding gene only (which may result in duplicated data)
    mutate(target = ifelse(n() > 1, 'other protein-coding genes', target)) %>%
    ungroup() %>%
    # second removal of duplicate data
    distinct()
  
  # there is a possibility a single guide target multiple locations of the same gene and other gene/random genomic locations, this code is used to separate the sgRNA that align multiple times to the same gene with those align to other gene/genomic loci
  multiple_aln_same_target <- alignment_data_gene_df %>%
    filter(target == "actual target gene" & !sgrna %in% summary_target_df$sgrna) %>%
    count(sgrna) %>%
    filter(n > 1) %>%
    dplyr::rename("number_of_alignments" = "n")
  
  # count affected genes
  
  sgrna_target_genes <- summary_target_df %>%
    filter(target == "other protein-coding genes") %>%
    distinct() %>%
    pull(sgrna)
  
  affected_genes <- alignment_data_gene_df %>%
    filter(sgrna %in% sgrna_target_genes & target == "other protein-coding genes") %>%
    select(actual_gene) %>%
    distinct() %>%
    pull(actual_gene)
  
  df_affected_genes_percentage <- data.frame(target = c("other genomic loci", "other protein-coding genes"),
                                             count_genes = c(0, length(affected_genes)),
                                             percent_genes = c(0, round(length(affected_genes)/length(unique(library$gene))*100, 2)))
  
  
  # create frequency table
  
  calculate_freq_table <- summary_target_df %>%
    select(target) %>%
    count(target) %>%
    dplyr::rename('count_sgrna' = 'n') %>%
    arrange(desc(count_sgrna)) %>%
    mutate(percent_sgrna_among_library = round(count_sgrna/length(unique(library$sgRNA))*100 ,2),
           percent_sgrna_among_off_target = round(count_sgrna/length(unique(alignment_data_gene_df$sgrna))*100, 2),
           library = library_name) %>%
    relocate(library, .before = target) %>%
    left_join(df_affected_genes_percentage, join_by(target))
  
  
  return(list(
    aln_data_df = alignment_data_gene_df,
    multiple_aln_same_target = multiple_aln_same_target,
    summary_df = summary_target_df, 
    freq_table = calculate_freq_table,
    list_affected_genes = affected_genes))
}


# ----------------------------------------------------------------------------------------------------------------------------------------------


# Function to identify multi-target alignment targeting paralog pairs, strictly only for 2 protein-coding genes (paralog A1 and paralog A2)

find_multi_target_paralog <- function(alignment_gene_df, list_guides, paralog_dataset, library, library_name) {
  dbl_target_classification <- alignment_gene_df %>%
    filter(sgrna %in% list_guides) %>%
    mutate(target = case_when(paste(gene, actual_gene, sep ='_') %in% paralog_dataset ~ 'targeting paralog gene',
                              gene == actual_gene ~ 'actual gene',
                              is.na(actual_gene) ~ 'other genomic loci',
                              TRUE ~ 'other protein-coding gene'))
  
  aln_multi_target_paralog <- dbl_target_classification %>%
    group_by(sgrna) %>%
    filter(all(c("targeting paralog gene", "actual gene") %in% target),
           n_distinct(target) > 1) %>%
    ungroup()
  
  list_paralog_guides <- dbl_target_classification %>% filter(target == "targeting paralog gene") %>% pull(sgrna) %>% unique()
  list_other_genes_guides <- dbl_target_classification %>% filter(target == "other protein-coding gene") %>% pull(sgrna) %>% unique()
  list_other_genomic_loci <- dbl_target_classification %>% filter(target == "other genomic loci") %>% pull(sgrna) %>% unique()
  
  list_other_genomic_loci_guides <- setdiff(
    list_other_genomic_loci,
    c(list_other_genes_guides, list_paralog_guides)
  )
  
  
  
  list_not_paralog_guides <- union(list_other_genes_guides, list_other_genomic_loci_guides)
  
  # obtain the affected genes by looking at the actual target from double-target guides
  list_paralog_genes <- unique(aln_multi_target_paralog$actual_gene)
  
  # count percentage of the double target paralog guides
  
  count_sgrna_paralog <- length(list_paralog_guides)
  count_sgrna_other_genes <- length(list_other_genes_guides)
  count_sgrna_other_genomic_loci <- length(list_other_genomic_loci_guides)
  
  percentage_sgrna_paralog <- round((count_sgrna_paralog/sum(count_sgrna_paralog, count_sgrna_other_genes, count_sgrna_other_genomic_loci) * 100), 2)
  percentage_sgrna_other_genes <- round((count_sgrna_other_genes/sum(count_sgrna_paralog, count_sgrna_other_genes, count_sgrna_other_genomic_loci) * 100), 2)
  percentage_sgrna_other_genomic_loci <- round((count_sgrna_other_genomic_loci/sum(count_sgrna_paralog, count_sgrna_other_genes, count_sgrna_other_genomic_loci) * 100), 2)
  
  summary_df <- data.frame(library_name,
                           count_sgrna_paralog, 
                           count_sgrna_other_genes, 
                           count_sgrna_other_genomic_loci, 
                           percentage_sgrna_paralog, 
                           percentage_sgrna_other_genes, 
                           percentage_sgrna_other_genomic_loci)
  
  # count percentage of genes affected
  
  paralog_genes_affected <- length(list_paralog_genes)
  total_genes_library <- length(unique(library$gene))
  percentage_genes_affected <- round((paralog_genes_affected/total_genes_library)*100, 2)
  
  count_gene <- data.frame(library_name, paralog_genes_affected, total_genes_library, percentage_genes_affected)
  
  return(list(df_target_classification = dbl_target_classification,
              aln_multi_target_paralog_df = aln_multi_target_paralog,
              summary_count_df = summary_df,
              multi_target_paralog_guides = list_paralog_guides,
              multi_target_other_genes = list_other_genes_guides,
              multi_target_other_genomic_loci = list_other_genomic_loci_guides,
              paralog_affected_list = list_paralog_genes,
              paralog_affected_df = count_gene))
  
}


```

# Code to separate paralog and singleton of double-target guides

```{r}
avana_dbl_target_genes_paralog <- summarize_paralog_targets(dbl_target_avana, paralog_individual, "avana")[[1]]
brunello_dbl_target_genes_paralog <- summarize_paralog_targets(dbl_target_brunello, paralog_individual, "brunello")[[1]]
toronto_dbl_target_genes_paralog <- summarize_paralog_targets(dbl_target_toronto, paralog_individual, "toronto_v3")[[1]]
yusa_dbl_target_genes_paralog <- summarize_paralog_targets(dbl_target_yusa, paralog_individual, "yusa")[[1]]

df_summary = rbind(avana_dbl_target_genes_paralog, 
                   brunello_dbl_target_genes_paralog, 
                   toronto_dbl_target_genes_paralog, 
                   yusa_dbl_target_genes_paralog)

df_summary <- df_summary %>% mutate(library = factor(library, levels = c("brunello", "toronto_v3", "yusa", "avana")),
                                    category = factor(category, levels = c("singleton", "paralog")))

df_paralog_dbl_target_genes <- ggplot(df_summary, aes(x = library, y = percentage, fill = category)) +
      geom_bar(stat = "identity", position = "stack") +
      scale_fill_discrete(
          name = "",
          labels = c("Singleton", "Paralog")) +
        theme_minimal() +
        labs(title = "", x = "", y = "Intended target gene of double-target guides (%)", fill = "Target") +
        scale_x_discrete(labels = c("brunello" = "Brunello",
                                     "toronto_v3" = "TKOv3",
                                     "yusa" = "Yusa \n(Project Score)",
                                     "avana" = "Avana")) +
        geom_text(aes(label = percentage), position = position_stack(), vjust = 2, size = 8) +
        theme(axis.text.x = element_text(size = 25, vjust = 0.7, colour = "black"),
              axis.text.y = element_text(size = 25, colour = "black"),
              axis.title.y = element_text(size = 25, vjust = 0.5, margin = margin(r = 20)),
              panel.grid.major = element_blank(),
              panel.grid.minor = element_blank(),
              axis.line.x = element_line(size = 0.5),
              axis.line.y = element_line(size = 0.5),
              legend.title = element_text(size = 25),
              legend.text = element_text(size = 25),
              plot.margin = margin(1, 1, 1, 1), 
              line = element_line((size = 4), colour = "black"),
              legend.spacing.x = unit(1, 'cm')) +
      ylim(0, 100)

df_paralog_dbl_target_genes

ggsave(
    path = './',
    width = 12,
    height = 10,
    dpi = 300,
    plot = df_paralog_dbl_target_genes,
    filename = paste('../figure/02_fig_analysis_paralog_off-target-sgrna/dbl_target_guides_original_genes_paralog_T2T.png', sep = ''))
```


# Relevant statistical test for sgRNA Log2FC

```{r}

calculate_spearman_corr <- function(input_data) {
  
  input_data$alignment_bin_num <- as.numeric(ifelse(input_data$alignment_bin == "> 5", 7, input_data$alignment_bin))
  spearman_test <- cor.test(input_data$alignment_bin_num, as.numeric(input_data$mean), method="spearman", alternative="less")
  
  paste("spearman corr rho: ", spearman_test$estimate)
  paste("spearman pvalue: ", format.pval(spearman_test$p.value, digits=5))
  
  return(list(spearman_test, input_data))
}


tkov3_spearman <- calculate_spearman_corr(tkov3_library_and_lfc_data)

# Basic Kruskal–Wallis test
kruskal.test(tkov3_library_and_lfc_data$mean ~ tkov3_library_and_lfc_data$alignment_bin, data = tkov3_library_and_lfc_data)

# Post-hoc pairwise comparisons (non-parametric)
pairwise.wilcox.test(tkov3_library_and_lfc_data$mean, tkov3_library_and_lfc_data$alignment_bin, p.adjust.method = "BH")

# Jonckheree test
tkov3_library_and_lfc_data$alignment_bin_num <- as.numeric(as.character(tkov3_library_and_lfc_data$alignment_bin))
tkov3_library_and_lfc_data$alignment_bin_num[tkov3_library_and_lfc_data$alignment_bin == ">5"] <- 6
jonckheere.test(x = tkov3_library_and_lfc_data$mean, g = tkov3_library_and_lfc_data$alignment_bin_num, alternative = "decreasing")

```


# Backup Code BIOGRID

# Prepare data from BIOGRID

```{r}

biogrid_index <- read_tsv("../data/biogrid_crispr_ko_data/important_data/BIOGRID-ORCS-SCREEN_INDEX-2.0.18.index.tab.txt")

# we select CRISPR KO screen based on cell proliferation, library type is CRISPR knockout (CRISPRn), and containing list of full genes and their hits (full size available - yes)

# available library total:  37
# selected library:
# 1. Avana
# 2. Brunello (human)
# 3. Toronto Knockout version 3 (TKOv3)
# 4. None for Project Score

# create blank dataframe
df_avana_screen <- data.frame()

avana_screen_studies <- biogrid_index %>%
  filter(PHENOTYPE == "cell proliferation" & LIBRARY_TYPE == "CRISPRn" & FULL_SIZE_AVAILABLE == "Yes") %>%
  # count(AUTHOR) %>%
  # dplyr::rename("number_of_screens" = "n")
  pull(`#SCREEN_ID`)

for (screen_id in avana_screen_studies) {
  data = read_tsv(paste("../data/biogrid_crispr_ko_data/BIOGRID-ORCS-SCREEN_", screen_id, "-2.0.18.screen.tab.txt", sep = ""), col_names = FALSE, skip = 1)
  
  # X13 is column indicating whether the genes is the hits or not
  data <- data %>% filter(X13 == "YES")
  df_avana_screen = rbind(df_avana_screen, data)
}

colnames(df_avana_screen) <- c("SCREEN_ID", "IDENTIFIER_ID", "IDENTIFIER_TYPE", "OFFICIAL_SYMBOL", "ALIASES", "ORGANISM_ID", "ORGANISM_OFFICIAL", "SCORE.1", "SCORE.2", "SCORE.3", "SCORE.4", "SCORE.5", "HIT", "SOURCE")

```

# Check if the removed genes is in the CRISPR screen data

```{r}

removed_genes_not_hits <- setdiff(unique(removed_genes$avana_removed_genes_all), unique(df_avana_screen$OFFICIAL_SYMBOL))

removed_genes_hits <- df_avana_screen %>% 
  filter(OFFICIAL_SYMBOL %in% removed_genes$avana_removed_genes_all) %>%
  count(OFFICIAL_SYMBOL) %>%
  dplyr::rename("count_appear_in_screen" = "n") %>%
  count(count_appear_in_screen) %>%
  dplyr::rename("num_genes" = "n") %>%
  add_row(count_appear_in_screen = 0, num_genes = length(removed_genes_not_hits)) %>%
  arrange(count_appear_in_screen)

breaks <- c(-Inf, 0, 50, 100, 150, 200, 250, 300, Inf)
labels <- c("0", "1–50", "50–100", "100–150", "150–200", "200–250", "250–300", ">300")
  
removed_genes_hits$appearance_bin <- cut(removed_genes_hits$count_appear_in_screen,
                                        breaks = breaks,
                                        labels = labels,
                                        right = TRUE)

removed_gene_hits_bin <- removed_genes_hits %>%
  select(num_genes, appearance_bin) %>%
  group_by(appearance_bin) %>%
  summarise(num_genes_total = sum(num_genes))


```



```{r}
removed_genes_list_hits <- df_avana_screen %>% 
  filter(OFFICIAL_SYMBOL %in% removed_genes$avana_removed_genes_all) %>%
  group_by(OFFICIAL_SYMBOL) %>%
  count(OFFICIAL_SYMBOL) %>%
  dplyr::rename("count_as_hits" = "n") %>%
  ungroup()



import_report_without_control <- function(file_dir, control_terms) {
  
  data <- read_excel(paste(file_dir))
  data <- data %>% filter(!str_detect(gene, terms_for_filtering))
  return(data)
  
}

avana <- 'avana_library'

terms <- c("CONTROL", "Control", "control", "INTRON", "Intron", "intron", "LacZ", "luciferase")
# lacZ and luciferase are the control in Toronto V3 library

terms_for_filtering <- paste(terms, collapse = "|")

avana_report <- import_report_without_control(paste('~/CRISPR-KO-GuideRefine/output_cleaning/T2T-CHM13/', avana, '_full_report.xlsx', sep = ''))

query_hits <- removed_genes_list_hits %>% filter(count_as_hits == 344) %>% pull(OFFICIAL_SYMBOL)

avana_report_query_hits <- avana_report %>% filter(gene %in% query_hits)

```


```{r}
# check duplicated sgRNA in library and lfc data


#find duplicated sgRNA
# check_duplicate <- brunello_library_and_lfc_data[[1]] %>%
#   filter(duplicated(sgRNA) | duplicated(sgRNA, fromLast = TRUE)) %>%
#   distinct()



check_pam_distal <- jacquere_library_and_lfc_data[[1]] %>% filter(alignment == "pam-distal single mismatch") %>% select(sgRNA) %>% distinct()

avana_library_and_lfc_data[[1]] <- remove_duplicate_sgrna(avana_library_and_lfc_data[[1]]) %>% mutate(library = "avana") 
tkov3_library_and_lfc_data[[1]] <- remove_duplicate_sgrna(tkov3_library_and_lfc_data[[1]]) %>% mutate(library = "toronto_v3") 
yusa_library_and_lfc_data[[1]] <- remove_duplicate_sgrna(yusa_library_and_lfc_data[[1]]) %>% mutate(library = "yusa") 
brunello_library_and_lfc_data[[1]] <- remove_duplicate_sgrna(brunello_library_and_lfc_data[[1]]) %>% mutate(library = "brunello") 
jacquere_library_and_lfc_data[[1]] <- remove_duplicate_sgrna(jacquere_library_and_lfc_data[[1]]) %>% mutate(library = "jacquere") 

combined_data <- rbind(avana_library_and_lfc_data[[1]], 
                       tkov3_library_and_lfc_data[[1]], 
                       yusa_library_and_lfc_data[[1]], 
                       brunello_library_and_lfc_data[[1]], 
                       jacquere_library_and_lfc_data[[1]])

combined_data <- combined_data %>% mutate(library = factor(library, levels = c("brunello", "toronto_v3", "yusa", "avana", "jacquere")))


test <- combined_data %>%
  select(sgRNA, alignment, library) %>%
  distinct() %>%
  group_by(library) %>%
  count(alignment)


facet_labels = c("brunello" = "Brunello",
                 "toronto_v3" = "TKOv3",
                 "yusa" = "Yusa (Project Score)",
                 "avana" = "Avana",
                 "jacquere" = "Jacquere")

boxplot_lfc <- ggboxplot(combined_data, 
                         x = "alignment", 
                         y = "mean",
                         fill = "alignment",
                         facet.by = "library",
                         outliers = FALSE) +
  stat_compare_means(comparisons = list(c("pam-distal single mismatch", "single mismatch"), c("pam-distal single mismatch", "perfect"), c("single mismatch", "perfect")),
                     method = "wilcox.test",
                     method.args = list(alternative = "less"),
                     size = 3.5,
                     step.increase = 0.05,
                     tip.length = 0.01,
                     label = "p.format",
                     label.y = 1) +
  scale_y_continuous(limits = c(-3, 3), breaks = seq(-2, 2, by = 1)) +
  scale_fill_manual(
    name = "sgRNA: ",
    values = c("perfect" = "#0072B2", 
               "single mismatch" = "mistyrose",
               "pam-distal single mismatch" = "#E69F00"),
    labels = c("On-target", "single mismatch", "pam-distal single mismatch")) +
  labs(title = "", 
       x = "", 
       y = "Averaged median sgRNA Log2FC",
       fill = "Alignment:") +
  theme(plot.margin = margin(10, 10, 25, 10),
        strip.text.x = element_text(size = 12),
        axis.text.x = element_blank(),
        axis.text.y = element_text(size = 12, colour = "black"),
        axis.ticks.y = element_line(size = 1.5),
        axis.ticks.x = element_blank(), 
        axis.ticks.length = unit(0.1, "cm"),
        axis.title.y = element_text(size = 14, vjust = 0.5, margin = margin(r = 20)),
        panel.grid.major = element_line(size = 0.5),
        panel.grid.minor = element_blank(),
        axis.line.x = element_line(size = 0.5),
        axis.line.y = element_line(size = 0.5), 
        line = element_line((size = 2), colour = 'black'),
        legend.text = element_text(size = 12)) +
  facet_wrap(~library, ncol = 5, labeller = labeller(library = facet_labels))


b <- ggplot_build(boxplot_lfc)
stats <- b$data[[1]]


test <- test %>% bind_cols(stats %>% select(x, ymin))


boxplot_lfc <-  boxplot_lfc +
  geom_text(
    data = test,
    aes(x = x, y = ymin - 0.5, label = paste0("n= ", n)),
    angle = 45,
    vjust = 0,
    hjust = 0.5,
    size = 3
  )

boxplot_lfc


colnames(test)


ggsave(
  path = './',
  width = 10,
  height = 5,
  dpi = 1000,
  plot = boxplot_lfc,
  filename = "../figure/03_fig_analysis_lfc_off-target-sgrna/combined_library_single_mismatch_pam.png")

```

# Backup function to remove duplicate sgRNA

```{r}
# remove duplicated sgRNA after obtain required data
remove_duplicate_sgrna <- function(library_and_lfc_data) {
  
  condition_priority <- c(
    "perfect" = 1,
    "single mismatch" = 2,
    "pam-distal single mismatch" = 3
  )
  
  # Alignment number priority
  alignment_levels <- c("0", "1", "2", "3", "4", "5", "> 5")
  alignment_priority <- setNames(seq_along(alignment_levels), alignment_levels)
  
  df_selected <- library_and_lfc_data %>%
    # Add numeric priority columns
    mutate(
      cond_priority = condition_priority[alignment],
      align_priority = alignment_priority[as.character(alignment_bin)]
    ) %>%
    
    # Group by sgRNA (or your key)
    group_by(sgRNA) %>%
    
    # Keep only rows with the highest condition priority
    filter(cond_priority == max(cond_priority)) %>%
    
    # Within those, keep only the row with highest alignment number
    filter(align_priority == max(align_priority)) %>%
    
    ungroup() %>%
    
    # Drop temporary columns
    select(-cond_priority, -align_priority) %>%
    
    distinct()
  
}
```

# Backup function for obtain required data

```{r}

obtain_required_data <- function(library, normalized_lfc_data, stratification_data) {
  
  library <- read_tsv(library, col_names = FALSE)
  colnames(library) <- c("sgRNA", "spacer", "gene")
  
  normalized_lfc_data <- read_csv(normalized_lfc_data)
  library_stratification <- read_csv(stratification_data, col_types = cols(alignment_bin = col_character()
  )) %>% select(sgRNA, alignment, alignment_bin)
  
  # left join
  library_lfc_data <- normalized_lfc_data %>%
    left_join(library, by = c("sgRNA", "spacer", "gene")) %>%
    relocate(sgRNA, .before = spacer) %>%
    relocate(gene, .after = spacer) %>%
    arrange(sgRNA)
  
  terms <- c("CONTROL", "Control", "control", "INTRON", "Intron", "intron", "LacZ", "luciferase")
  # lacZ and luciferase are the control in Toronto V3 library
  terms_for_filtering <- paste(terms, collapse = "|")
  
  # separate control from the library
  
  control_lfc_data <- library_lfc_data %>% filter(str_detect(gene, terms_for_filtering))
  
  library_lfc <- library_lfc_data %>%
    filter(!str_detect(gene, terms_for_filtering)) %>%
    left_join(library_stratification, join_by(sgRNA))
  
  # TODO: fix this if condition logic, the problem is in selecting single mismatch and single mismatch at pam distal
  
  if (str_detect(as.character(stratification_data), "multi_target")) {
    selected_data <- "multi-target guides"
  # } else if (str_detect(as.character(stratification_data), "single_mismatch")) {
  #   selected_data <- "single mismatch"
  } else if (str_detect(as.character(stratification_data), "single_mismatch")) {
    selected_data <- "pam-distal single mismatch"
  }
  
  # for the first boxplot: stratify the number of multi-target guides and single mismatch guides based on the number of alignment
  library_lfc_alignment <- library_lfc %>%
    # multi-target guides or single mismatch
    filter(alignment %in% c("perfect", "single mismatch", paste0(selected_data)))
  
  library_lfc_alignment$alignment <- factor(library_lfc_alignment$alignment, levels = c("perfect", "single mismatch", paste0(selected_data)))
  
  # for the second boxplot: select guides only targeting multi-target guides or single mismatch
  library_lfc_alignment_bin <- library_lfc %>%
    filter(alignment_bin %in% c(0, 1, 2, 3, 4, 5,"> 5"))
  
  library_lfc_alignment_bin$alignment_bin <- factor(library_lfc_alignment_bin$alignment_bin, levels = c(0, 1, 2, 3, 4, 5, "> 5"))
  # library_lfc_alignment is for comparison of sgRNA Log2FC between two group
  # library_lfc_alignment_bin is for stratification of multi-target guides (the reason why we put this because we included guides not targeting anywhere)
  return(list(library_lfc_alignment, library_lfc_alignment_bin))
}

```

# Backup code for analysis of removed genes (negative result)

1. Our first hypothesis is maybe the same sgRNA design were used across libraries for some genes. Which led to the failure of on-target activity 

```{r}

duplicated_sequence <- combined_alignment %>%
  # filter(duplicated(spacer) | duplicated(spacer, fromLast = TRUE)) %>%
  dplyr::select(spacer, library) %>%
  group_by(library) %>%
  summarise(seqs = list(unique(spacer))) %>%
  deframe()
  
upset(fromList(duplicated_sequence),
      order.by = "degree",
      main.bar.color = "#4B0092",
      sets.bar.color = "#1AFF1A",
      text.scale = 1.5)

```

However, only 30 sgRNA sequences, that failed and shared among the libraries (except Yusa Project Score). Therefore, sgRNA sequences is not a contributing factor for multi-targeting-/off-targeting sgRNAs.

We asked whether paralogs are the major contributing factor in causing multi-target-/off-targeting activity in sgRNAs. 

```{r}

# count how many 520 genes has paralogs
df <- hgnc %>%
  left_join(ensembl_paralog[[4]], join_by(gene, locus_group, ensembl_gene_id)) %>%
  mutate(has_paralogs = replace_na(has_paralogs, FALSE))
  
# count percentage of paralogs
df_percent_paralogs <- df %>%
  dplyr::select(gene, has_paralogs) %>%
  distinct() %>%
  count(has_paralogs) %>%
  dplyr::rename("count" = "n") %>%
  mutate(percent = count / sum(count) * 100)

# 377 genes out of 520 that were removed across libraries are paralogs (72.5%), indicating significant number of enrichment


# we focused our attention to the 377 paralog genes that were removed across libraries. Furthermore, we discussed the 177 singleton genes separately

removed_paralog <- df %>% 
  dplyr::select(gene, has_paralogs) %>%
  group_by(gene) %>%
  summarise(count_paralog = if_else(has_paralogs == TRUE, n(), 0)) %>%
  arrange(count_paralog) %>%
  ungroup() %>%
  distinct()

removed_paralog$paralogs_bin <- cut(removed_paralog$count_paralog,
                                    breaks = c(-0.5, 0.5:5.5, Inf),  # -0.5 to 5.5 for 0–5, then Inf for >5
                                    labels = c(as.character(0:5), "> 5"),
                                    right = TRUE)

removed_paralog %>% count(paralogs_bin)

# 0	143			
# 1	67			
# 2	38			
# 3	25			
# 4	11			
# 5	11			
# > 5	225	

removed_paralog_bin <- removed_paralog %>% 
  count(paralogs_bin) %>% 
  dplyr::rename("count_paralog" = "n")

boxplot_paralog <- ggplot(removed_paralog_bin, aes(x = paralogs_bin, y = count_paralog)) +
    geom_col(fill = "#1E88E5") +
    theme_minimal() +
  labs(x = "Number of paralogs", y = "Removed genes across 5 libraries") +
  geom_text(aes(label = paste("n=", count_paralog, sep="")),
            nudge_y = 10,
            nudge_x = 0,
            size = 5) +
  theme(axis.text.x = element_text(size = 15, vjust = 0.7, colour = 'black'),
        axis.text.y = element_text(size = 15, colour = "black"),
        axis.title.x = element_text(size = 15, vjust = 0.5, margin = margin(t = 10), colour = "black"),
        axis.title.y = element_text(size = 15, vjust = 0.5, margin = margin(r = 10), colour = "black"),
        axis.ticks.y = element_line(size = 1.5),
        axis.ticks.length = unit(0.15, "cm"),
        axis.line.y.right = element_blank(),
        axis.ticks.y.right = element_blank(), 
        axis.text.y.right = element_blank(), 
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.line.x = element_line(size = 0.5, colour = "black"),
        axis.line.y = element_line(size = 0.5, colour = "black"),
        legend.position = "none",
        plot.margin = margin(1, 1, 1, 1),
        legend.spacing.x = unit(1, 'cm'))

ggsave(
  path = '../figure/01_fig_analysis_off-target-sgrna/',
  width = 6.5,
  height = 5,
  dpi = 1000,
  plot = boxplot_paralog,
  filename = "removed_genes_paralogs.png")


# we found most removed genes are containing paralogs family, with more than >1 paralogs

# since paralogs came from gene duplication event, we asked whether the genes with more paralogs have their sequences more identical with the original genes

removed_paralog_max <- df %>% 
  dplyr::select(gene, has_paralogs, percent_identity_human_gene_identical_to_query) %>%
  group_by(gene) %>%
  # count paralog NA because they dont have paralog gene
  summarise(count_paralog = if_else(has_paralogs == TRUE, n(), NA),
            max_percent_identity = max(percent_identity_human_gene_identical_to_query, na.rm = TRUE)) %>%
  arrange(count_paralog) %>%
  ungroup() %>%
  distinct() %>%
  # drop NA/removed genes without paralogs
  drop_na()

removed_paralog_max$paralogs_bin <- cut(removed_paralog_max$count_paralog,
                                    breaks = c(-0.5, 0.5:5.5, Inf),  # -0.5 to 5.5 for 0–5, then Inf for >5
                                    labels = c(as.character(0:5), "> 5"),
                                    right = TRUE)


annotation_df <- removed_paralog_max %>%
    count(paralogs_bin)

boxplot_percent_identity_paralog <- removed_paralog_max %>%
    ggplot(aes(paralogs_bin, max_percent_identity, group = paralogs_bin)) +
    geom_boxplot(outlier.shape = NA, size = 0.5) +
    theme_minimal() +
    labs(title = "", 
         x = "Number of paralogs for a single gene", 
         y = "% of protein sequences \nmatched in closest paralog") +
    theme(
      axis.text.x = element_text(size = 15, vjust = 0.7, colour = "black"),
      axis.text.y = element_text(size = 15, colour = "black"),
      axis.title.x = element_text(size = 15, margin = margin(t = 10)),
      axis.title.y = element_text(size = 15, margin = margin(r = 10)),
      axis.ticks = element_line(size = 1.5),
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      axis.line.x = element_line(size = 0.5),
      axis.line.y = element_line(size = 0.5))

  b <- ggplot_build(boxplot_percent_identity_paralog)
  boxplot_whisker <- b$data[[1]]$ymax
  
  boxplot_percent_identity_paralog <- boxplot_percent_identity_paralog +
    geom_text(
      data = annotation_df,
      aes(x = paralogs_bin, y = boxplot_whisker + 0.1, label = paste0("n= ", n)),
      size = 3.5,
      angle = 45,
      vjust = -0.2,
      hjust = -0.2
    ) +
    coord_cartesian(clip = "off")


ggsave(
  path = '../figure/01_fig_analysis_off-target-sgrna/',
  width = 5,
  height = 6,
  dpi = 1000,
  plot = boxplot_percent_identity_paralog,
  filename = "removed_genes_paralogs_percent_identity.png")



```

## Backup code for percentage off-target sgRNA T2T-CHM13

```{r}

# the import_and_process_report() already update the gene symbol of each guides, hence the multi-target and off-target calculation is already correct

brunello_multi_target <- round(calculate_percentage(brunello_report, "multi-target sgRNA", "sgRNA number"), 2)
toronto_multi_target <- round(calculate_percentage(toronto_report, "multi-target sgRNA", "sgRNA number"), 2)
avana_multi_target <- round(calculate_percentage(avana_report, "multi-target sgRNA", "sgRNA number"), 2)
yusa_multi_target <- round(calculate_percentage(yusa_report, "multi-target sgRNA", "sgRNA number"), 2)
jacquere_multi_target <- round(calculate_percentage(jacquere_report, "multi-target sgRNA", "sgRNA number"), 2)

brunello_single_mismatch <- round(calculate_percentage(brunello_report, "single mismatch sgRNA", "sgRNA number"), 2)
toronto_single_mismatch <- round(calculate_percentage(toronto_report, "single mismatch sgRNA", "sgRNA number"), 2)
avana_single_mismatch <- round(calculate_percentage(avana_report, "single mismatch sgRNA", "sgRNA number"), 2)
yusa_single_mismatch <- round(calculate_percentage(yusa_report, "single mismatch sgRNA", "sgRNA number"), 2)
jacquere_single_mismatch <- round(calculate_percentage(jacquere_report, "single mismatch sgRNA", "sgRNA number"), 2)

brunello_pam_distal_double <- round(calculate_percentage(brunello_report, "PAM distal double mismatch", "sgRNA number"), 2)
toronto_pam_distal_double <- round(calculate_percentage(toronto_report, "PAM distal double mismatch", "sgRNA number"), 2)
avana_pam_distal_double <- round(calculate_percentage(avana_report, "PAM distal double mismatch", "sgRNA number"), 2)
yusa_pam_distal_double <- round(calculate_percentage(yusa_report, "PAM distal double mismatch", "sgRNA number"), 2)
jacquere_pam_distal_double <- round(calculate_percentage(jacquere_report, "PAM distal double mismatch", "sgRNA number"), 2)

# Create dataframe for multi-target, single-mismatch, and pam-distal from all library

multi_target <- c(brunello_multi_target, toronto_multi_target, yusa_multi_target, avana_multi_target, jacquere_multi_target)
single_mismatch <- c(brunello_single_mismatch, toronto_single_mismatch, yusa_single_mismatch, avana_single_mismatch, jacquere_single_mismatch)
pam_distal_double <- c(brunello_pam_distal_double, toronto_pam_distal_double, yusa_pam_distal_double, avana_pam_distal_double, jacquere_pam_distal_double)

library <- c("brunello", "toronto_v3", "yusa", "avana", "jacquere")
factor <- c(1,2,3,4,5)

sum_dataframe <- data.frame(library, multi_target, single_mismatch, pam_distal_double)

# Reorder the library factor to ensure the bars are ordered as Brunello, Toronto, Yusa, Avana
sum_dataframe_long <- sum_dataframe %>%
    pivot_longer(cols = c("multi_target", "single_mismatch", "pam_distal_double"), 
                 names_to = "category", 
                 values_to = "percentage") %>%
    mutate(category = factor(category, levels = c("multi_target", "single_mismatch", "pam_distal_double")),
           library = factor(library, levels = c("brunello", "toronto_v3", "yusa", "avana", "jacquere")))


# Define custom colors for the libraries
my_palette <- brewer.pal(n = 5, name = "Set2")


# Create the grouped multibar chart for each sgRNA libraries
multi_bar_chart <- ggplot(sum_dataframe_long, aes(x = category, y = percentage, fill = library)) +
    geom_bar(stat = "identity", position = position_dodge2(padding = 0.2), width = 0.7) +
    theme_minimal() +
    labs(title = "", x = "", y = "sgRNA (%)") +
    scale_x_discrete(labels = c('multi_target' = 'Multi-target sgRNA',
                                'single_mismatch' = 'Single mismatch sgRNA',
                                'pam_distal_double' = 'PAM distal \n double mismatch sgRNA')) +
    geom_text(aes(label = percentage), position = position_dodge2(width = 0.7), vjust = -0.5, size = 6.5) +
    theme(axis.text.x = element_text(size = 25, vjust = 0.7, colour = "black"),
          axis.text.y = element_text(size = 25, colour = "black"),
          axis.title.y = element_text(size = 25, vjust = 0.5, margin = margin(r = 20)),
          panel.grid.major = element_blank(),
          panel.grid.minor = element_blank(),
          axis.line.x = element_line(size = 0.5),
          axis.line.y = element_line(size = 0.5),
          axis.ticks.y = element_line(size = 1.5),
          axis.ticks.length = unit(0.3, "cm"),
          legend.title = element_text(size = 25),
          legend.text = element_text(size = 25),
          plot.margin = margin(1, 1, 1, 1), 
          line = element_line((size = 4), colour = 'black'),
          legend.spacing.x = unit(1, 'cm')) +
    scale_fill_manual(values = my_palette,
                      name = 'Library',
                      breaks = c("brunello", "toronto_v3", "yusa", "avana", "jacquere"),
                      labels = c("brunello" = "Brunello",
                                 "toronto_v3" = "TKOv3",
                                 "yusa" = "Yusa (Project Score)",
                                 "avana" = "Avana",
                                 "jacquere" = "Jacquere")) + 
  scale_y_continuous(breaks=seq(0, 10, 2), limits=c(0, 10))

multi_bar_chart

ggsave(
    path = './',
    width = 20,
    height = 10,
    dpi = 300,
    plot = multi_bar_chart,
    filename = paste('../figure/01_fig_analysis_off-target-sgrna/percentage_off_target_T2T.png', sep = ''))
```


# Backup code for Gene Ontology Olivieri Screens

Perform gene ontology enrichment analysis

```{r}
hits_before_refining <- c(AZD6738_original_only_hits, AZD6738_overlap_hits)

GOBP_hits_before_refining <- enrichGO(gene = hits_before_refining,
                                     OrgDb = org.Hs.eg.db,
                                     ont = "BP",
                                     pAdjustMethod = "BH",
                                     qvalueCutoff = 0.05,
                                     keyType = "SYMBOL",
                                     minGSSize = 10,
                                     universe = original_library)
GOBP_hits_before_refining_simplified <- simplify(
  GOBP_hits_before_refining,
  cutoff = 0.7,          # similarity threshold
  by = "p.adjust",       # which column defines "best"
  select_fun = min,
  measure = "Wang"       # GO semantic similarity
)

dotplot(GOBP_hits_before_refining_simplified, showCategory = 10) +
  ggtitle("AZD6738 original hits")

```


```{r}
hits_after_refining <- c(AZD6738_refined_only_hits, AZD6738_overlap_hits)

GOBP_hits_after_refining <- enrichGO(gene = hits_after_refining,
                                     OrgDb = org.Hs.eg.db,
                                     ont = "BP",
                                     pAdjustMethod = "BH",
                                     qvalueCutoff = 0.05,
                                     keyType = "SYMBOL",
                                     minGSSize = 10,
                                     universe = refined_library)
GOBP_hits_after_refining_simplified <- simplify(
  GOBP_hits_after_refining,
  cutoff = 0.7,          # similarity threshold
  by = "p.adjust",       # which column defines "best"
  select_fun = min,
  measure = "Wang"       # GO semantic similarity
)

dotplot(GOBP_hits_after_refining_simplified, showCategory = 10) +
  ggtitle("AZD6738 hits after refining")

```

```{r}
GOBP_original_only_hits <- enrichGO(gene = AZD6738_original_only_hits,
                                     OrgDb = org.Hs.eg.db,
                                     ont = "BP",
                                     pAdjustMethod = "BH",
                                     qvalueCutoff = 0.05,
                                     keyType = "SYMBOL",
                                     minGSSize = 10,
                                     universe = original_library)

GOBP_original_only_hits_simplified <- simplify(
  GOBP_original_only_hits,
  cutoff = 0.7,          # similarity threshold
  by = "p.adjust",       # which column defines "best"
  select_fun = min,
  measure = "Wang"       # GO semantic similarity
)

dotplot(GOBP_original_only_hits_simplified, showCategory = 10) +
  ggtitle("AZD6738 original only gene hits")

```




















