---
title: "Percentage off-target sgRNAs and removed genes across CRISPR-KO libraries"
author: "Author: Stefanus Bernard"
date: "Date: `r format(Sys.time(), '%d %B %Y')`"
output: 
    html_document:
      theme: cerulean
      toc: true
      toc_float: true
      number_sections: true
      self_contained: true
---

```{r}
library(tidyverse)
library(readxl)
library(grid)
library(paletteer)
library(RColorBrewer)
library(UpSetR)
library(gprofiler2)
library(HGNChelper)
source('../../GuideRefine/GuideRefine_functions.R')
source("00_CRISPR-KO-library-survey-functions.R")
```

# Import the data of sgRNA library report

```{r}

terms <- c("CONTROL", "Control", "control", "INTRON", "Intron", "intron", "LacZ", "luciferase", "NO_SITE", "ONE_INTERGENIC_SITE")
# lacZ and luciferase are the control in Toronto V3 library

terms_for_filtering <- paste(terms, collapse = "|")

brunello <- 'broadgpp-brunello-library-contents'
toronto <- 'tkov3_guide_sequence'
avana <- 'avana_library'
yusa <- 'yusa_hcrispr_ko_grnas'
jacquere <- 'Jacquere_PerGuideAnnotations_Quota4'

# import sgRNA refining report

avana_report <- import_and_process_report(paste0('../../GuideRefine/output_cleaning/T2T-CHM13/', avana, '_full_report.xlsx'))
brunello_report <- import_and_process_report(paste0('../../GuideRefine/output_cleaning/T2T-CHM13/', brunello, '_full_report.xlsx'))
toronto_report <- import_and_process_report(paste0('../../GuideRefine/output_cleaning/T2T-CHM13/', toronto, '_full_report.xlsx'))
yusa_report <- import_and_process_report(paste0('../../GuideRefine/output_cleaning/T2T-CHM13/', yusa, '_full_report.xlsx'))
jacquere_report <- import_and_process_report(paste0('../../GuideRefine/output_cleaning/T2T-CHM13/', jacquere, '_full_report.xlsx'))

```

# Figure 2A - Percentage of multi-target, single-mismatch, and PAM-distal double mismatch sgRNA in each libraries

```{r}
pick_alignment_type_and_process_data <- function(classification_data, alignment_type, stratification_data, library_name) {
  data <- read_csv(classification_data) %>%
    filter(alignment %in% alignment_type)

  data <- data %>%
    mutate(library = library_name) %>%
    group_by(library, alignment) %>%
    select(library, alignment) %>%
    summarise(num_sgrna = n()) %>%
    mutate(percentage = round(num_sgrna/sum(num_sgrna) * 100, 2))
  
  
  return(data)
}

avana_classification <- pick_alignment_type_and_process_data("../data/sgrna_lfc_data/avana_data/avana_sgrna_alignment_classification_T2T.csv", c("perfect", "single mismatch", "pam-distal single mismatch", "pam-distal double mismatch", "multi-target guides"), library_name = "avana")
  
tkov3_classification <- pick_alignment_type_and_process_data("../data/sgrna_lfc_data/tkov3_data/tkov3_sgrna_alignment_classification_T2T.csv", c("perfect", "single mismatch", "pam-distal single mismatch", "pam-distal double mismatch", "multi-target guides"), library_name = "tkov3")

yusa_classification <- pick_alignment_type_and_process_data("../data/sgrna_lfc_data/yusa_data/yusa_sgrna_alignment_classification_T2T.csv", c("perfect", "single mismatch", "pam-distal single mismatch", "pam-distal double mismatch", "multi-target guides"), library_name = "yusa")

brunello_classification <- pick_alignment_type_and_process_data("../data/sgrna_lfc_data/brunello_data/brunello_sgrna_alignment_classification_T2T.csv", c("perfect", "single mismatch", "pam-distal single mismatch", "pam-distal double mismatch", "multi-target guides"), library_name = "brunello")

jacquere_classification <- pick_alignment_type_and_process_data("../data/sgrna_lfc_data/jacquere_data/jacquere_sgrna_alignment_classification_T2T.csv", c("perfect", "single mismatch", "pam-distal single mismatch", "pam-distal double mismatch", "multi-target guides"), library_name = "jacquere")

sum_dataframe_long <- rbind(avana_classification, tkov3_classification, yusa_classification, brunello_classification, jacquere_classification) %>%
  filter(!alignment == "perfect") %>%
  mutate(alignment = factor(alignment, levels = c("multi-target guides", "single mismatch", "pam-distal single mismatch", "pam-distal double mismatch")),
         library = factor(library, levels = c("brunello", "tkov3", "yusa", "avana", "jacquere")))

# Define custom colors for the libraries
my_palette <- brewer.pal(n = 5, name = "Set2")


# Create the grouped multibar chart for each sgRNA libraries
multi_bar_chart <- ggplot(sum_dataframe_long, aes(x = alignment, y = percentage, fill = library)) +
    geom_bar(stat = "identity", position = position_dodge2(padding = 0.2), width = 0.7) +
    theme_minimal() +
    labs(title = "", x = "", y = "sgRNA (%)") +
    scale_x_discrete(labels = c('multi-target guides' = 'Multi-target sgRNA',
                                'single mismatch' = 'Single mismatch sgRNA',
                                'pam-distal single mismatch' = 'PAM distal \n single mismatch sgRNA',
                                'pam-distal double mismatch' = 'PAM distal \n double mismatch sgRNA')) +
    geom_text(aes(label = percentage), position = position_dodge2(width = 0.7), vjust = -0.5, size = 8) +
    theme(axis.text.x = element_text(size = 35, vjust = 0.7, colour = "black"),
          axis.text.y = element_text(size = 35, colour = "black"),
          axis.title.y = element_text(size = 35, vjust = 0.5, margin = margin(r = 20)),
          panel.grid.major = element_blank(),
          panel.grid.minor = element_blank(),
          axis.line.x = element_line(size = 0.5),
          axis.line.y = element_line(size = 0.5),
          axis.ticks.y = element_line(size = 1.5),
          axis.ticks.length = unit(0.3, "cm"),
          legend.title = element_text(size = 25),
          legend.text = element_text(size = 25),
          plot.margin = margin(1, 1, 1, 1), 
          line = element_line((size = 4), colour = 'black'),
          legend.spacing.x = unit(1, 'cm')) +
    scale_fill_manual(values = my_palette,
                      name = 'Library',
                      breaks = c("brunello", "tkov3", "yusa", "avana", "jacquere"),
                      labels = c("brunello" = "Brunello",
                                 "tkov3" = "TKOv3",
                                 "yusa" = "Yusa (Project Score)",
                                 "avana" = "Avana",
                                 "jacquere" = "Jacquere")) + 
  scale_y_continuous(breaks=seq(0, 10, 2), limits=c(0, 10))

multi_bar_chart

ggsave(
    path = './',
    width = 22,
    height = 10,
    dpi = 300,
    plot = multi_bar_chart,
    filename = paste('../figure/Fig02_analysis_off-target-sgrna/Fig2A_percentage_off_target_T2T.png', sep = ''))

```

### Percentage of genes not represented after library refinement

```{r}

# in here we pull the failed/passed genes based on protein-coding genes defined by HGNC (gene_hgnc), it was first standardized using import_and_process_report()

avana_result_gene <- percentages_and_list_genes(avana_report, NULL)
brunello_result_gene <- percentages_and_list_genes(brunello_report, NULL)
toronto_result_gene <- percentages_and_list_genes(toronto_report, NULL)
yusa_result_gene <- percentages_and_list_genes(yusa_report, NULL)
jacquere_result_gene <- percentages_and_list_genes(jacquere_report, NULL)

```

### Export the removed genes from each library (input for script 04_analysis_removed_genes_biogrid.Rmd)

```{r}

max_len <- max(length(avana_result_gene[["failed_genes"]]),
               length(brunello_result_gene[["failed_genes"]]),
               length(toronto_result_gene[["failed_genes"]]),
               length(yusa_result_gene[["failed_genes"]]),
               length(jacquere_result_gene[["failed_genes"]]))

pad <- function(x, n) {
  length(x) <- n
  return(x)
}

df_removed_genes <- data.frame(
  avana_removed_genes = pad(avana_result_gene[["failed_genes"]], max_len),
  brunello_removed_genes = pad(brunello_result_gene[["failed_genes"]], max_len),
  toronto_removed_genes = pad(toronto_result_gene[["failed_genes"]], max_len),
  yusa_removed_genes = pad(yusa_result_gene[["failed_genes"]], max_len),
  jacquere_removed_genes = pad(jacquere_result_gene[["failed_genes"]], max_len),
  stringsAsFactors = FALSE
)

write_csv(df_removed_genes, "../data/biogrid_crispr_ko_data/important_data/removed_genes_all_library.csv")
```

### Total percentages removed sgRNAs and unrepresented genes across libraries

```{r}

main_df_percentage <- data.frame(
  library = c("brunello", "toronto_v3", "yusa", "avana", "jacquere"),
  total_percentage_failed_genes = c(brunello_result_gene[["total_percentage_failed_genes"]],
                                    toronto_result_gene[["total_percentage_failed_genes"]],
                                    yusa_result_gene[["total_percentage_failed_genes"]],
                                    avana_result_gene[["total_percentage_failed_genes"]],
                                    jacquere_result_gene[["total_percentage_failed_genes"]]),
  total_percentage_failed_sgrna = c(brunello_result_gene[["total_percentage_failed_sgrna"]],
                                    toronto_result_gene[["total_percentage_failed_sgrna"]],
                                    yusa_result_gene[["total_percentage_failed_sgrna"]],
                                    avana_result_gene[["total_percentage_failed_sgrna"]],
                                    jacquere_result_gene[["total_percentage_failed_sgrna"]])
)

```

# Figure 2B - plot of sgRNA that is not represented in the refined library

```{r}

# in here we pull the failed/passed genes based on protein-coding genes defined by HGNC (gene_hgnc). then the failed_sgrna_percentage came from main_df_percentage, which it was from the percentage_and_list_gene(), a function to standardize gene symbol based on HGNC and select protein-coding genes only

failed_sgrna_percentage <- main_df_percentage %>%
  dplyr::select(library, total_percentage_failed_sgrna) %>%
  mutate(library = factor(library, levels = c("brunello", "toronto_v3", "yusa", "avana", "jacquere")))

percentage_bar_chart <- ggplot(failed_sgrna_percentage, aes(x = library, y = total_percentage_failed_sgrna)) +
    geom_bar(stat = "identity", position = "stack", fill = "#1E88E5") +
        theme_minimal() +
        labs(title = "", x = "", y = "Removed sgRNAs (%)") +
        scale_x_discrete(labels = c("brunello" = "Brunello",
                                     "toronto_v3" = "TKOv3",
                                     "yusa" = "Yusa \n(Project Score)",
                                     "avana" = "Avana",
                                     "jacquere" = "Jacquere")) +
        geom_text(aes(label = total_percentage_failed_sgrna), position = position_stack(), vjust = -0.5, size = 10.5) +
        theme(axis.text.x = element_text(size = 32.5, vjust = 0.7, colour = 'black'),
              axis.text.y = element_text(size = 35, colour = 'black'),
              axis.title.y = element_text(size = 35, vjust = 0.5, margin = margin(r = 20)),
              panel.grid.major = element_blank(),
              panel.grid.minor = element_blank(),
              axis.line.x = element_line(size = 0.5),
              axis.line.y = element_line(size = 0.5),
              axis.ticks.y = element_line(size = 1.5),
              axis.ticks.length = unit(0.3, "cm"),
              plot.margin = margin(1, 1, 1, 1), 
              line = element_line((size = 4), colour = 'black'),
              legend.spacing.x = unit(1, 'cm')) +
    scale_fill_manual(values = "",
                      name = 'Library',
                      breaks = c("brunello", "toronto_v3", "yusa", "avana", "jacquere"),
                      labels = c("brunello" = "Brunello",
                                 "toronto_v3" = "TKOv3",
                                 "yusa" = "Yusa (Project Score)",
                                 "avana" = "Avana",
                                 "jacquere" = "Jacquere")) +
      ylim(0, 100)

percentage_bar_chart

ggsave(
    path = './',
    width = 13.5,
    height = 10,
    dpi = 300,
    plot = percentage_bar_chart,
    filename = paste('../figure/Fig02_analysis_off-target-sgrna/Fig2B_percentage_removed_sgrna_T2T.png', sep = ''))


```

# Figure 3A - Plot of genes that are not represented in the refined library

```{r}

failed_genes_percentage <- main_df_percentage %>%
  mutate(library = factor(library, levels = c("brunello", "toronto_v3", "yusa", "avana", "jacquere")))

percentage_bar_chart <- ggplot(failed_genes_percentage, aes(x = library, y = total_percentage_failed_genes)) +
    geom_bar(stat = "identity", position = "stack", fill = "#1E88E5") +
        theme_minimal() +
        labs(title = "", x = "", y = "Protein-coding genes targeted by < 3 sgRNAs (%)") +
        scale_x_discrete(labels = c("brunello" = "Brunello",
                                     "toronto_v3" = "TKOv3",
                                     "yusa" = "Yusa \n(Project Score)",
                                     "avana" = "Avana",
                                     "jacquere" = "Jacquere")) +
        geom_text(aes(label = total_percentage_failed_genes), position = position_stack(), vjust = -0.5, size = 10.5) +
        theme(axis.text.x = element_text(size = 32.5, vjust = 0.7, colour = 'black'),
              axis.text.y = element_text(size = 35, colour = 'black'),
              axis.title.y = element_text(size = 35, vjust = 0.5, margin = margin(r = 20)),
              panel.grid.major = element_blank(),
              panel.grid.minor = element_blank(),
              axis.line.x = element_line(size = 0.5),
              axis.line.y = element_line(size = 0.5),
              axis.ticks.y = element_line(size = 1.5),
              axis.ticks.length = unit(0.3, "cm"),
              plot.margin = margin(1, 1, 1, 1), 
              line = element_line((size = 4), colour = 'black'),
              legend.spacing.x = unit(1, 'cm')) +
    scale_fill_manual(values = "",
                      name = 'Library',
                      breaks = c("brunello", "toronto_v3", "yusa", "avana", "jacquere"),
                      labels = c("brunello" = "Brunello",
                                 "toronto_v3" = "TKOv3",
                                 "yusa" = "Yusa (Project Score)",
                                 "avana" = "Avana",
                                 "jacquere" = "Jacquere")) +
      ylim(0, 100)

percentage_bar_chart

ggsave(
    path = './',
    width = 13,
    height = 10,
    dpi = 300,
    plot = percentage_bar_chart,
    filename = paste('../figure/Fig03_analysis_removed_genes/Fig3A_library_genes_bar_chart_T2T.png', sep = ''))

```





