---
title: "Analysis of multi-target and off-target sgRNAs across libraries"
author: "Author: Stefanus Bernard"
date: "Date: `r format(Sys.time(), '%d %B %Y')`"
output:
  html_document :
    theme : cerulean
    toc : true
    toc_float : true
    number_section : true
---

```{r setup, include=FALSE}
library(here)
library(tidyverse)
library(ggbreak)
library(clinfun)
library(ggsignif)
library(ggpubr)
library(rstatix)
library(effsize)
library(dplyr)

source(file.path(dirname(here::here()), "GuideRefine", "GuideRefine_functions.R"))
source(here::here("scripts", "00_CRISPR-KO-library-survey-functions.R"))
```

# Load data (once)

```{r load-classification}

avana_raw    <- read_csv(here::here("data", "sgrna_lfc_data", "avana_data", "avana_sgrna_alignment_classification_T2T.csv"))
tkov3_raw    <- read_csv(here::here("data", "sgrna_lfc_data", "tkov3_data", "tkov3_sgrna_alignment_classification_T2T.csv"))
yusa_raw     <- read_csv(here::here("data", "sgrna_lfc_data", "yusa_data", "yusa_sgrna_alignment_classification_T2T.csv"))
brunello_raw <- read_csv(here::here("data", "sgrna_lfc_data", "brunello_data", "brunello_sgrna_alignment_classification_T2T.csv"))
jacquere_raw <- read_csv(here::here("data", "sgrna_lfc_data", "jacquere_data", "jacquere_sgrna_alignment_classification_T2T.csv"))

```

```{r load-library-lfc}

avana_lfc_raw    <- load_library_lfc_raw(here::here("data", "library_data", "original_library", "avana_library.tsv"),
                                          here::here("data", "sgrna_lfc_data", "output_normalized", "avana_normalized_lfc.csv"))

tkov3_lfc_raw    <- load_library_lfc_raw(here::here("data", "library_data", "original_library", "tkov3_guide_sequence.tsv"),
                                          here::here("data", "sgrna_lfc_data", "output_normalized", "tkov3_rpe1_normalized_lfc.csv"))

yusa_lfc_raw     <- load_library_lfc_raw(here::here("data", "library_data", "original_library", "yusa_hcrispr_ko_grnas.tsv"),
                                          here::here("data", "sgrna_lfc_data", "output_normalized", "yusa_normalized_lfc.csv"))

brunello_lfc_raw <- load_library_lfc_raw(here::here("data", "library_data", "original_library", "broadgpp-brunello-library-contents.tsv"),
                                          here::here("data", "sgrna_lfc_data", "output_normalized", "brunello_a375_normalized_lfc.csv"))

# Jacquere sgRNA Log2FC data only available for ~60,000 sgRNAs
jacquere_lfc_raw <- load_library_lfc_raw(here::here("data", "library_data", "original_library", "Jacquere_PerGuideAnnotations_Quota4.tsv"),
                                          here::here("data", "sgrna_lfc_data", "output_normalized", "jacquere_normalized_lfc.csv"))

```

# Multi-target sgRNA analysis — Fig2C, SuppFig2A–E

```{r multi-target-stratify}
source(here::here("scripts", "00_CRISPR-KO-library-survey-functions.R"))

multi_guides <- c("perfect", "multi-target guides")

avana_multi    <- count_alignment_bin(filter(avana_raw,    alignment %in% multi_guides), multi_guides)
tkov3_multi    <- count_alignment_bin(filter(tkov3_raw,    alignment %in% multi_guides), multi_guides)
yusa_multi     <- count_alignment_bin(filter(yusa_raw,     alignment %in% multi_guides), multi_guides)
brunello_multi <- count_alignment_bin(filter(brunello_raw, alignment %in% multi_guides), multi_guides)
jacquere_multi <- count_alignment_bin(filter(jacquere_raw, alignment %in% multi_guides), multi_guides)

avana_lfc_multi    <- apply_stratification(avana_lfc_raw,    avana_multi[[1]])    %>% mutate(library = "avana")      %>% drop_na()
tkov3_lfc_multi    <- apply_stratification(tkov3_lfc_raw,    tkov3_multi[[1]])    %>% mutate(library = "toronto_v3") %>% drop_na()
yusa_lfc_multi     <- apply_stratification(yusa_lfc_raw,     yusa_multi[[1]])     %>% mutate(library = "yusa")       %>% drop_na()
brunello_lfc_multi <- apply_stratification(brunello_lfc_raw, brunello_multi[[1]]) %>% mutate(library = "brunello")   %>% drop_na()
jacquere_lfc_multi <- apply_stratification(jacquere_lfc_raw, jacquere_multi[[1]]) %>% mutate(library = "jacquere")  %>% drop_na()

combined_multi <- rbind(avana_lfc_multi, tkov3_lfc_multi, yusa_lfc_multi, brunello_lfc_multi, jacquere_lfc_multi) %>%
  mutate(
    library   = factor(library,   levels = c("brunello", "toronto_v3", "yusa", "avana", "jacquere")),
    alignment = factor(alignment, levels = multi_guides)
  )

```

```{r multi-target-export}

visualize_two_group(combined_multi,
                    here::here("figure", "Fig2C_combined_library_multi_target.png"))

visualize_stratify_alignment(brunello_lfc_multi, "number of perfect on-target alignment", here::here("figure", "SuppFig2A_brunello_multi_target.png"))
visualize_stratify_alignment(tkov3_lfc_multi,    "number of perfect on-target alignment", here::here("figure", "SuppFig2B_tkov3_multi_target.png"))
visualize_stratify_alignment(yusa_lfc_multi,     "number of perfect on-target alignment", here::here("figure", "SuppFig2C_yusa_multi_target.png"))
visualize_stratify_alignment(avana_lfc_multi,    "number of perfect on-target alignment", here::here("figure", "SuppFig2D_avana_multi_target.png"))
visualize_stratify_alignment(jacquere_lfc_multi, "number of perfect on-target alignment", here::here("figure", "SuppFig2E_jacquere_multi_target.png"))

```

# Single mismatch sgRNA analysis — Fig2D, SuppFig2F–J

```{r single-mismatch-stratify}
source(here::here("scripts", "00_CRISPR-KO-library-survey-functions.R"))

single_guides <- c("perfect", "single mismatch")

avana_single    <- count_alignment_bin(filter(avana_raw,    alignment %in% single_guides), single_guides)
tkov3_single    <- count_alignment_bin(filter(tkov3_raw,    alignment %in% single_guides), single_guides)
yusa_single     <- count_alignment_bin(filter(yusa_raw,     alignment %in% single_guides), single_guides)
brunello_single <- count_alignment_bin(filter(brunello_raw, alignment %in% single_guides), single_guides)
jacquere_single <- count_alignment_bin(filter(jacquere_raw, alignment %in% single_guides), single_guides)

avana_lfc_single    <- apply_stratification(avana_lfc_raw,    avana_single[[1]])    %>% mutate(library = "avana")      %>% drop_na()
tkov3_lfc_single    <- apply_stratification(tkov3_lfc_raw,    tkov3_single[[1]])    %>% mutate(library = "toronto_v3") %>% drop_na()
yusa_lfc_single     <- apply_stratification(yusa_lfc_raw,     yusa_single[[1]])     %>% mutate(library = "yusa")       %>% drop_na()
brunello_lfc_single <- apply_stratification(brunello_lfc_raw, brunello_single[[1]]) %>% mutate(library = "brunello")   %>% drop_na()
jacquere_lfc_single <- apply_stratification(jacquere_lfc_raw, jacquere_single[[1]]) %>% mutate(library = "jacquere")  %>% drop_na()

combined_single <- rbind(avana_lfc_single, tkov3_lfc_single, yusa_lfc_single, brunello_lfc_single, jacquere_lfc_single) %>%
  mutate(
    library   = factor(library,   levels = c("brunello", "toronto_v3", "yusa", "avana", "jacquere")),
    alignment = factor(alignment, levels = single_guides)
  )

```

```{r single-mismatch-export}

visualize_two_group(combined_single,
                    here::here("figure", "Fig2D_combined_library_single_mismatch.png"),
                    comparison_group = "single mismatch",
                    comparison_label = "Single mismatch")

visualize_stratify_alignment(brunello_lfc_single, "number of single mismatch \noff-target alignment", here::here("figure", "SuppFig2F_brunello_single_mismatch.png"))
visualize_stratify_alignment(tkov3_lfc_single,    "number of single mismatch \noff-target alignment", here::here("figure", "SuppFig2G_tkov3_single_mismatch.png"))
visualize_stratify_alignment(yusa_lfc_single,     "number of single mismatch \noff-target alignment", here::here("figure", "SuppFig2H_yusa_single_mismatch.png"))
visualize_stratify_alignment(avana_lfc_single,    "number of single mismatch \noff-target alignment", here::here("figure", "SuppFig2I_avana_single_mismatch.png"))
visualize_stratify_alignment(jacquere_lfc_single, "number of single mismatch \noff-target alignment", here::here("figure", "SuppFig2J_jacquere_single_mismatch.png"))

```

# Wilcoxon rank-sum test values — Fig 2C and Fig 2D

```{r wilcoxon-fig2c-fig2d}

# Fig 2C: perfect vs multi-target guides (Wilcoxon unpaired, two-sided), per library
# library             W   p_value n_perfect n_multi
#   <fct>           <dbl>     <dbl>     <int>   <int>
# 1 brunello   129775196. 5.35e-242     64718    2946
# 2 toronto_v3  77225532  3.51e- 49     64690    2001
# 3 yusa       204894804  0             78803    3366
# 4 avana      237346891  0             59860    5027
# 5 jacquere    44073838  4.25e-108     51842    1245

wilcoxon_fig2c <- combined_multi %>%
  group_by(library) %>%
  group_modify(~ {
    perf <- filter(.x, alignment == "perfect")$mean
    tgt  <- filter(.x, alignment == "multi-target guides")$mean
    w    <- wilcox.test(perf, tgt, alternative = "two.sided")
    tibble(W = w$statistic, p_value = w$p.value,
           n_perfect = length(perf), n_multi = length(tgt))
  })
print(wilcoxon_fig2c)

# Fig 2D: perfect vs single mismatch (Wilcoxon unpaired, two-sided), per library
# ---------------------------------------------------------------------------------
#   library             W   p_value n_perfect n_single
#   <fct>           <dbl>     <dbl>     <int>    <int>
# 1 brunello   219728676. 5.61e- 60     64718     6025
# 2 toronto_v3 106678302  2.49e- 19     64690     3007
# 3 yusa       321381273  4.36e-190     78803     6708
# 4 avana      223715847  1.52e- 80     59860     6540
# 5 jacquere   184407364. 6.79e-166     51842     5837

wilcoxon_fig2d <- combined_single %>%
  group_by(library) %>%
  group_modify(~ {
    perf   <- filter(.x, alignment == "perfect")$mean
    single <- filter(.x, alignment == "single mismatch")$mean
    w      <- wilcox.test(perf, single, alternative = "two.sided")
    tibble(W = w$statistic, p_value = w$p.value,
           n_perfect = length(perf), n_single = length(single))
  })
print(wilcoxon_fig2d)

```

# Jonckheere-Terpstra trend test — Supp Fig 2A–2J

Tests the hypothesis that median sgRNA LFC decreases monotonically as the
number of off-target alignments increases across ordered bins.
Asymptotic normal approximation; accurate for n > 100 per group.

```{r jt-trend-supp2a-2j}

# JT statistic computed as sum of Mann-Whitney U over all ordered group pairs (i < j),
# matching clinfun's convention of counting pairs where x_i < x_j.
# Asymptotic normal approximation used for z-score and p-value.
run_jt <- function(df, fig_label, lib_name, analysis) {
  g      <- as.numeric(df$alignment_bin)
  x      <- df$mean
  groups <- sort(unique(g))
  k      <- length(groups)
  N      <- length(x)
  n      <- as.integer(table(g))

  jt_stat <- 0
  for (i in seq_len(k - 1)) {
    for (j in (i + 1):k) {
      xi <- x[g == groups[i]]
      xj <- x[g == groups[j]]
      jt_stat <- jt_stat + suppressWarnings(wilcox.test(xj, xi)$statistic)
    }
  }

  mu     <- (N^2 - sum(n^2)) / 4
  sigma2 <- (N^2 * (2*N + 3) - sum(n^2 * (2*n + 3))) / 72
  z      <- (jt_stat - mu) / sqrt(sigma2)

  tibble(figure       = fig_label,
         library      = lib_name,
         analysis     = analysis,
         n_total      = N,
         JT_statistic = round(jt_stat),
         z_score      = round(z, 1),
         p_value      = pnorm(z))
}

jt_results <- bind_rows(
  run_jt(brunello_lfc_multi,  "SuppFig2A", "Brunello",  "Multi-target guides"),
  run_jt(tkov3_lfc_multi,     "SuppFig2B", "TKOv3",     "Multi-target guides"),
  run_jt(yusa_lfc_multi,      "SuppFig2C", "Yusa",      "Multi-target guides"),
  run_jt(avana_lfc_multi,     "SuppFig2D", "Avana",     "Multi-target guides"),
  run_jt(jacquere_lfc_multi,  "SuppFig2E", "Jacquere",  "Multi-target guides"),
  run_jt(brunello_lfc_single, "SuppFig2F", "Brunello",  "Single mismatch off-target"),
  run_jt(yusa_lfc_single,     "SuppFig2G", "Yusa",      "Single mismatch off-target"),
  run_jt(tkov3_lfc_single,    "SuppFig2H", "TKOv3",     "Single mismatch off-target"),
  run_jt(avana_lfc_single,    "SuppFig2I", "Avana",     "Single mismatch off-target"),
  run_jt(jacquere_lfc_single, "SuppFig2J", "Jacquere",  "Single mismatch off-target")
)
print(jt_results)

# figure    library  analysis            n_total JT_statistic z_score   p_value
#    <chr>     <chr>    <chr>                 <int>        <dbl>   <dbl>     <dbl>
#  1 SuppFig2A Brunello Multi-target guides   67664     61920967   -33.6 3.95e-247
#  2 SuppFig2B TKOv3    Multi-target guides   66691     52433949   -14.8 8.93e- 50
#  3 SuppFig2C Yusa     Multi-target guides   82169     61005189   -53.8 0        
#  4 SuppFig2D Avana    Multi-target guides   64887     65877308   -69.4 0        
#  5 SuppFig2E Jacquere Multi-target guides   53087     20629019   -22.2 2.58e-109
#  6 SuppFig2F Brunello Single mismatch of…   70743    173872602   -16.7 1.17e- 62
#  7 SuppFig2G Yusa     Single mismatch of…   85511    210035071   -29.8 4.57e-196
#  8 SuppFig2H TKOv3    Single mismatch of…   67697     88313450    -9   1.24e- 19
#  9 SuppFig2I Avana    Single mismatch of…   66400    171715438   -19.5 7.67e- 85
# 10 SuppFig2J Jacquere Single mismatch of…   57679    121786702   -27.8 4.84e-170

```
