---
title: "Analysis of multi-target and off-target sgRNAs across libraries"
author: "Author: Stefanus Bernard"
date: "Date: `r format(Sys.time(), '%d %B %Y')`"
output:
  html_document :
    theme : cerulean
    toc : true
    toc_float : true
    number_section : true
---

```{r setup, include=FALSE}
library(here)
library(tidyverse)
library(ggsignif)
library(ggpubr)
library(clinfun)

source(here::here("scripts", "GuideRefine-functions.R"))
source(here::here("scripts", "CRISPR-KO-library-survey-functions.R"))
```

# Load data (once)

```{r load-classification}

avana_raw    <- read_csv(here::here("data", "sgrna_lfc_data", "avana_data", "avana_sgrna_alignment_classification_T2T.csv"))
tkov3_raw    <- read_csv(here::here("data", "sgrna_lfc_data", "tkov3_data", "tkov3_sgrna_alignment_classification_T2T.csv"))
yusa_raw     <- read_csv(here::here("data", "sgrna_lfc_data", "yusa_data", "yusa_sgrna_alignment_classification_T2T.csv"))
brunello_raw <- read_csv(here::here("data", "sgrna_lfc_data", "brunello_data", "brunello_sgrna_alignment_classification_T2T.csv"))
jacquere_raw <- read_csv(here::here("data", "sgrna_lfc_data", "jacquere_data", "jacquere_sgrna_alignment_classification_T2T.csv"))

```

```{r control-terms}
generic_terms <- c("CONTROL", "Control", "control", "INTRON", "Intron", "intron",
                    "LacZ", "luciferase", "EGFP", "NO_SITE", "ONE_SITE_INTERGENIC")
terms_for_filtering <- paste(generic_terms, collapse = "|")
```

```{r load-library-lfc}

avana_lfc_raw    <- load_library_lfc_raw(here::here("data", "library_data", "restricted_library", "avana_library.tsv"),
                                          here::here("data", "sgrna_lfc_data", "output_normalized", "avana_normalized_lfc.csv"),
                                          terms_for_filtering)

tkov3_lfc_raw    <- load_library_lfc_raw(here::here("data", "library_data", "restricted_library", "tkov3_guide_sequence.tsv"),
                                          here::here("data", "sgrna_lfc_data", "output_normalized", "tkov3_rpe1_normalized_lfc.csv"),
                                          terms_for_filtering)

yusa_lfc_raw     <- load_library_lfc_raw(here::here("data", "library_data", "restricted_library", "yusa_hcrispr_ko_grnas.tsv"),
                                          here::here("data", "sgrna_lfc_data", "output_normalized", "yusa_normalized_lfc.csv"),
                                          terms_for_filtering)

brunello_lfc_raw <- load_library_lfc_raw(here::here("data", "library_data", "restricted_library", "broadgpp-brunello-library-contents.tsv"),
                                          here::here("data", "sgrna_lfc_data", "output_normalized", "brunello_a375_normalized_lfc.csv"),
                                          terms_for_filtering)

# Jacquere sgRNA Log2FC data only available for ~60,000 sgRNAs
jacquere_lfc_raw <- load_library_lfc_raw(here::here("data", "library_data", "restricted_library", "Jacquere_PerGuideAnnotations_Quota4.tsv"),
                                          here::here("data", "sgrna_lfc_data", "output_normalized", "jacquere_normalized_lfc.csv"),
                                          terms_for_filtering)

```

# Multi-target sgRNA analysis — Fig2C, SuppFig1B–F

```{r multi-target-stratify}
source(here::here("scripts", "CRISPR-KO-library-survey-functions.R"))

multi_guides <- c("perfect", "multi-target guides")

avana_multi    <- count_alignment_bin(filter(avana_raw,    alignment %in% multi_guides), multi_guides)
tkov3_multi    <- count_alignment_bin(filter(tkov3_raw,    alignment %in% multi_guides), multi_guides)
yusa_multi     <- count_alignment_bin(filter(yusa_raw,     alignment %in% multi_guides), multi_guides)
brunello_multi <- count_alignment_bin(filter(brunello_raw, alignment %in% multi_guides), multi_guides)
jacquere_multi <- count_alignment_bin(filter(jacquere_raw, alignment %in% multi_guides), multi_guides)

avana_lfc_multi    <- apply_stratification(avana_lfc_raw,    avana_multi[[1]])    %>% mutate(library = "avana")      %>% drop_na()
tkov3_lfc_multi    <- apply_stratification(tkov3_lfc_raw,    tkov3_multi[[1]])    %>% mutate(library = "toronto_v3") %>% drop_na()
yusa_lfc_multi     <- apply_stratification(yusa_lfc_raw,     yusa_multi[[1]])     %>% mutate(library = "yusa")       %>% drop_na()
brunello_lfc_multi <- apply_stratification(brunello_lfc_raw, brunello_multi[[1]]) %>% mutate(library = "brunello")   %>% drop_na()
jacquere_lfc_multi <- apply_stratification(jacquere_lfc_raw, jacquere_multi[[1]]) %>% mutate(library = "jacquere")  %>% drop_na()

combined_multi <- rbind(avana_lfc_multi, tkov3_lfc_multi, yusa_lfc_multi, brunello_lfc_multi, jacquere_lfc_multi) %>%
  mutate(
    library   = factor(library,   levels = c("brunello", "toronto_v3", "yusa", "avana", "jacquere")),
    alignment = factor(alignment, levels = multi_guides)
  )

```

```{r multi-target-export}

visualize_two_group(combined_multi,
                    here::here("figure", "Fig2C_combined_library_multi_target.png"))

visualize_stratify_alignment(brunello_lfc_multi, "number of perfect on-target alignment", here::here("figure", "SuppFig1B_brunello_multi_target.png"))
visualize_stratify_alignment(tkov3_lfc_multi,    "number of perfect on-target alignment", here::here("figure", "SuppFig1C_tkov3_multi_target.png"))
visualize_stratify_alignment(yusa_lfc_multi,     "number of perfect on-target alignment", here::here("figure", "SuppFig1D_yusa_multi_target.png"))
visualize_stratify_alignment(avana_lfc_multi,    "number of perfect on-target alignment", here::here("figure", "SuppFig1E_avana_multi_target.png"))
visualize_stratify_alignment(jacquere_lfc_multi, "number of perfect on-target alignment", here::here("figure", "SuppFig1F_jacquere_multi_target.png"))

```

# Single mismatch sgRNA analysis — Fig2D, SuppFig1G–K

```{r single-mismatch-stratify}
source(here::here("scripts", "CRISPR-KO-library-survey-functions.R"))

single_guides <- c("perfect", "single mismatch")

avana_single    <- count_alignment_bin(filter(avana_raw,    alignment %in% single_guides), single_guides)
tkov3_single    <- count_alignment_bin(filter(tkov3_raw,    alignment %in% single_guides), single_guides)
yusa_single     <- count_alignment_bin(filter(yusa_raw,     alignment %in% single_guides), single_guides)
brunello_single <- count_alignment_bin(filter(brunello_raw, alignment %in% single_guides), single_guides)
jacquere_single <- count_alignment_bin(filter(jacquere_raw, alignment %in% single_guides), single_guides)

avana_lfc_single    <- apply_stratification(avana_lfc_raw,    avana_single[[1]])    %>% mutate(library = "avana")      %>% drop_na()
tkov3_lfc_single    <- apply_stratification(tkov3_lfc_raw,    tkov3_single[[1]])    %>% mutate(library = "toronto_v3") %>% drop_na()
yusa_lfc_single     <- apply_stratification(yusa_lfc_raw,     yusa_single[[1]])     %>% mutate(library = "yusa")       %>% drop_na()
brunello_lfc_single <- apply_stratification(brunello_lfc_raw, brunello_single[[1]]) %>% mutate(library = "brunello")   %>% drop_na()
jacquere_lfc_single <- apply_stratification(jacquere_lfc_raw, jacquere_single[[1]]) %>% mutate(library = "jacquere")  %>% drop_na()

combined_single <- rbind(avana_lfc_single, tkov3_lfc_single, yusa_lfc_single, brunello_lfc_single, jacquere_lfc_single) %>%
  mutate(library   = factor(library,   levels = c("brunello", "toronto_v3", "yusa", "avana", "jacquere")),
         alignment = factor(alignment, levels = single_guides))

```

```{r single-mismatch-export}

visualize_two_group(combined_single,
                    here::here("figure", "Fig2D_combined_library_single_mismatch.png"),
                    comparison_group = "single mismatch",
                    comparison_label = "Single mismatch")

visualize_stratify_alignment(brunello_lfc_single, "number of single mismatch \noff-target alignment", here::here("figure", "SuppFig1G_brunello_single_mismatch.png"))
visualize_stratify_alignment(tkov3_lfc_single,    "number of single mismatch \noff-target alignment", here::here("figure", "SuppFig1H_tkov3_single_mismatch.png"))
visualize_stratify_alignment(yusa_lfc_single,     "number of single mismatch \noff-target alignment", here::here("figure", "SuppFig1I_yusa_single_mismatch.png"))
visualize_stratify_alignment(avana_lfc_single,    "number of single mismatch \noff-target alignment", here::here("figure", "SuppFig1J_avana_single_mismatch.png"))
visualize_stratify_alignment(jacquere_lfc_single, "number of single mismatch \noff-target alignment", here::here("figure", "SuppFig1K_jacquere_single_mismatch.png"))

```

# Wilcoxon rank-sum test values — Fig 2C and Fig 2D

```{r wilcoxon-fig2c-fig2d}
wilcoxon_fig2c <- combined_multi %>%
  group_by(library) %>%
  group_modify(~ {
    perf <- filter(.x, alignment == "perfect")$mean
    tgt  <- filter(.x, alignment == "multi-target guides")$mean
    w    <- wilcox.test(perf, tgt, alternative = "two.sided")
    tibble(W = w$statistic, p_value = w$p.value,
           n_perfect = length(perf), n_multi = length(tgt))
  })
print(wilcoxon_fig2c)

# Fig 2C: perfect vs multi-target guides (Wilcoxon unpaired, two-sided), per library
# library             W   p_value n_perfect n_multi
#   <fct>           <dbl>     <dbl>     <int>   <int>
# 1 brunello   110354292. 8.96e-203     62851    2599
# 2 toronto_v3  73519464. 4.74e- 48     63009    1956
# 3 yusa       194149227  0             77119    3254
# 4 avana      215704119  0             58386    4707
# 5 jacquere    34527918  1.77e- 85     49105    1038

wilcoxon_fig2d <- combined_single %>%
  group_by(library) %>%
  group_modify(~ {
    perf   <- filter(.x, alignment == "perfect")$mean
    single <- filter(.x, alignment == "single mismatch")$mean
    w      <- wilcox.test(perf, single, alternative = "two.sided")
    tibble(W = w$statistic, p_value = w$p.value,
           n_perfect = length(perf), n_single = length(single))
  })
print(wilcoxon_fig2d)

# Fig 2D: perfect vs single mismatch (Wilcoxon unpaired, two-sided), per library
# ---------------------------------------------------------------------------------
#   library             W   p_value n_perfect n_single
#   <fct>           <dbl>     <dbl>     <int>    <int>
# 1 brunello   201371736  6.41e- 59     62851     5673
# 2 toronto_v3 101151776  5.19e- 18     63009     2934
# 3 yusa       304714705  3.23e-183     77119     6503
# 4 avana      209698076  4.40e- 81     58386     6266
# 5 jacquere   158493056. 2.08e-165     49105     5249

```

# Jonckheere-Terpstra using clinfun — SuppFig1B–1K

Tests the hypothesis that median sgRNA LFC decreases monotonically as the
number of off-target alignments increases across ordered bins.
Asymptotic normal approximation; accurate for n > 100 per group.

```{r jt-clinfun}
run_jt_clinfun <- function(df, fig_label, lib_name, analysis) {
  g   <- as.numeric(df$alignment_bin)
  x   <- df$mean
  N   <- length(x)
  n   <- as.integer(table(g))

  # clinfun for the JT statistic; p.value overflows for large N so compute manually
  res     <- jonckheere.test(x, g, alternative = "decreasing", nperm = 0)
  jt_stat <- as.numeric(res$statistic)

  mu     <- (N^2 - sum(n^2)) / 4
  sigma2 <- (N^2 * (2*N + 3) - sum(n^2 * (2*n + 3))) / 72
  z      <- (jt_stat - mu) / sqrt(sigma2)

  tibble(
    figure       = fig_label,
    library      = lib_name,
    analysis     = analysis,
    n_total      = N,
    JT_statistic = round(jt_stat),
    z_score      = round(z, 1),
    p_value      = pnorm(z)
  )
}

jt_clinfun_results <- bind_rows(
  run_jt_clinfun(brunello_lfc_multi,  "SuppFig1B", "Brunello",  "Multi-target guides"),
  run_jt_clinfun(tkov3_lfc_multi,     "SuppFig1C", "TKOv3",     "Multi-target guides"),
  run_jt_clinfun(yusa_lfc_multi,      "SuppFig1D", "Yusa",      "Multi-target guides"),
  run_jt_clinfun(avana_lfc_multi,     "SuppFig1E", "Avana",     "Multi-target guides"),
  run_jt_clinfun(jacquere_lfc_multi,  "SuppFig1F", "Jacquere",  "Multi-target guides"),
  run_jt_clinfun(brunello_lfc_single, "SuppFig1G", "Brunello",  "Single mismatch off-target"),
  run_jt_clinfun(tkov3_lfc_single,    "SuppFig1H", "TKOv3",     "Single mismatch off-target"),
  run_jt_clinfun(yusa_lfc_single,     "SuppFig1I", "Yusa",      "Single mismatch off-target"),
  run_jt_clinfun(avana_lfc_single,    "SuppFig1J", "Avana",     "Single mismatch off-target"),
  run_jt_clinfun(jacquere_lfc_single, "SuppFig1K", "Jacquere",  "Single mismatch off-target")
)
print(jt_clinfun_results)

# figure    library  analysis            n_total JT_statistic z_score   p_value
#    <chr>     <chr>    <chr>                 <int>        <dbl>   <dbl>     <dbl>
#  1 SuppFig1B Brunello Multi-target guides   65450     53793708   -30.6 1.42e-206
#  2 SuppFig1C TKOv3    Multi-target guides   64965     49933331   -14.6 1.25e- 48
#  3 SuppFig1D Yusa     Multi-target guides   80373     57403158   -53.1 0        
#  4 SuppFig1E Avana    Multi-target guides   63093     61141296   -66.3 0        
#  5 SuppFig1F Jacquere Multi-target guides   50143     16553810   -19.7 2.19e- 86
#  6 SuppFig1G Brunello Single mismatch of…   68524    158338452   -16.5 1.79e- 61
#  7 SuppFig1H TKOv3    Single mismatch of…   65943     84159896    -8.6 2.63e- 18
#  8 SuppFig1I Yusa     Single mismatch of…   83622    199399257   -29.3 6.14e-189
#  9 SuppFig1J Avana    Single mismatch of…   64652    159699358   -19.5 2.70e- 85
# 10 SuppFig1K Jacquere Single mismatch of…   54354    102032516   -27.7 2.00e-169

```
