---
title: "Analysis of commonly removed genes across libraries"
author: "Author: Stefanus Bernard"
date: "Date: `r format(Sys.time(), '%d %B %Y')`"
output:
  html_document:
    theme: cerulean
    toc: true
    toc_float: true
    number_section: true
---

```{r setup, include=FALSE}
library(here)
library(tidyverse)
library(UpSetR)
library(GenomicFeatures)
library(GenomeInfoDb)
library(BSgenome.Hsapiens.NCBI.T2TCHM13v2.0)
library(GenomeInfoDbData)
library(txdbmaker)
library(crisprVerse)
library(ggpubr)

source(here::here("scripts", "CRISPR-KO-library-survey-functions.R"))
```

## Constants and helper functions

```{r helpers}

# RefSeq accession → chromosome name for T2T-CHM13v2.0 (defined once, reused throughout)
refseq_to_chr <- c(
  "NC_060925.1" = "1",  "NC_060926.1" = "2",  "NC_060927.1" = "3",
  "NC_060928.1" = "4",  "NC_060929.1" = "5",  "NC_060930.1" = "6",
  "NC_060931.1" = "7",  "NC_060932.1" = "8",  "NC_060933.1" = "9",
  "NC_060934.1" = "10", "NC_060935.1" = "11", "NC_060936.1" = "12",
  "NC_060937.1" = "13", "NC_060938.1" = "14", "NC_060939.1" = "15",
  "NC_060940.1" = "16", "NC_060941.1" = "17", "NC_060942.1" = "18",
  "NC_060943.1" = "19", "NC_060944.1" = "20", "NC_060945.1" = "21",
  "NC_060946.1" = "22", "NC_060947.1" = "X",  "NC_060948.1" = "Y"
)

# ---------------------------------------------------------------------------
# Compute total CDS or exon length (non-redundant) + genomic gene span per gene.
#   txdb       : TxDb object
#   gene_list  : character vector of gene IDs to subset
#   feature    : "cds" or "exon"
# Returns a data.frame with columns: gene, total_<feature>_length_bp, total_gene_length_bp
# ---------------------------------------------------------------------------
calculate_feature_length <- function(txdb, gene_list, feature = c("cds", "exon")) {
  feature <- match.arg(feature)

  # Gene span
  genes_gr     <- genes(txdb)
  genes_sub    <- genes_gr[mcols(genes_gr)$gene_id %in% gene_list]
  gene_span_df <- data.frame(
    gene                 = mcols(genes_sub)$gene_id,
    total_gene_length_bp = width(genes_sub)
  )

  # Feature (CDS or exon) length — reduce to remove transcript-level duplicates
  feat_by_gene <- if (feature == "cds") cdsBy(txdb, by = "gene") else exonsBy(txdb, by = "gene")
  feat_gr      <- feat_by_gene[names(feat_by_gene) %in% gene_list]
  total_len    <- sum(width(reduce(feat_gr)))

  len_col  <- if (feature == "cds") "total_cds_length_bp" else "total_exon_length_bp"
  feat_df  <- data.frame(gene = names(total_len), len = as.numeric(total_len))
  names(feat_df)[2] <- len_col
  rownames(feat_df) <- NULL

  feat_df %>% left_join(gene_span_df, join_by(gene))
}

# ---------------------------------------------------------------------------
# Count unique PAM sites per gene from a spacer data frame.
# Expects columns: gene_id, pam_site
# ---------------------------------------------------------------------------
count_pam_per_gene <- function(spacer_df) {
  spacer_df %>%
    distinct(gene_id, pam_site) %>%
    count(gene_id, name = "num_pam_site")
}

# ---------------------------------------------------------------------------
# Compute PAM-site density (PAM sites per kb of CDS) for a set of genes.
#   spacer_df : output of read_csv for spacer data (needs gene_id, pam_site)
#   cds_df    : output of calculate_feature_length(..., feature = "cds")
# ---------------------------------------------------------------------------
compute_pam_density <- function(spacer_df, cds_df) {
  count_pam_per_gene(spacer_df) %>%
    left_join(cds_df, join_by(gene_id == gene)) %>%
    mutate(pam_density = num_pam_site / total_cds_length_bp * 1000)
}

# ---------------------------------------------------------------------------
# Binned barplot: percentage of genes per bin.
#   data      : data.frame containing the bin column
#   bin_col   : column name (string) that holds the bin factor
#   xlabel    : x-axis label
#   ylim_max  : upper y-axis limit (default 50)
# ---------------------------------------------------------------------------
make_bin_barplot <- function(data, bin_col, xlabel, ylim_max = 50) {
  df <- data %>%
    count(.data[[bin_col]], name = "num_genes") %>%
    mutate(percentage = round(num_genes / sum(num_genes) * 100, 2))

  ggplot(df, aes(x = .data[[bin_col]], y = percentage)) +
    geom_col(fill = "#1E88E5") +
    geom_text(aes(label = percentage), nudge_y = 2, size = 7) +
    theme_minimal() +
    labs(x = xlabel, y = "Percentage of genes (%)") +
    theme(
      axis.text.x        = element_text(size = 20, colour = "black"),
      axis.text.y        = element_text(size = 20, colour = "black"),
      axis.title.x       = element_text(size = 20, margin = margin(t = 20), colour = "black"),
      axis.title.y       = element_text(size = 20, margin = margin(r = 20), colour = "black"),
      axis.ticks         = element_line(size = 1),
      axis.ticks.length  = unit(0.15, "cm"),
      axis.line.y.right  = element_blank(),
      axis.ticks.y.right = element_blank(),
      axis.text.y.right  = element_blank(),
      axis.line.x        = element_line(size = 0.5, colour = "black"),
      axis.line.y        = element_line(size = 0.5, colour = "black"),
      panel.grid.major   = element_blank(),
      panel.grid.minor   = element_blank(),
      legend.position    = "none",
      plot.margin        = margin(1, 1, 1, 1)
    ) +
    ylim(0, ylim_max)
}

# ---------------------------------------------------------------------------
# Log10-transformed boxplot comparing removed vs control genes with
# Wilcoxon significance annotation and per-group n labels.
#   combined_df  : data.frame with columns gene_status (factor), <y_col>
#   y_col        : column name string for the log10-transformed variable
#   ylabel       : y-axis label
#   sig_label_y  : y position of significance bracket (default 4)
#   ylim_range   : c(ymin, ymax) for y-axis limits (default c(0, 5))
# Returns a ggplot object (does NOT save — call ggsave() separately).
# ---------------------------------------------------------------------------
make_log10_boxplot <- function(combined_df, y_col, ylabel, sig_label_y = 4, ylim_range = c(0, 5)) {

  p <- ggboxplot(combined_df,
                 x        = "gene_status",
                 y        = y_col,
                 fill     = "gene_status",
                 outliers = FALSE) +
    stat_compare_means(
      comparisons  = list(c("removed genes", "control genes")),
      method       = "wilcox.test",
      method.args  = list(alternative = "two.sided"),
      size         = 5, step.increase = 0.05, tip.length = 0.01,
      label        = "p.signif", label.y = sig_label_y
    ) +
    scale_fill_manual(
      name   = "",
      values = c("control genes" = "#1E88E5", "removed genes" = "#E69F00"),
      labels = c("Other protein-coding genes", "Common genes with < 3 sgRNAs")
    ) +
    labs(x = "", y = ylabel) +
    theme(
      plot.margin       = margin(10, 10, 25, 10),
      strip.text.x      = element_text(size = 17.5),
      axis.text.x       = element_blank(),
      axis.text.y       = element_text(size = 17.5, colour = "black"),
      axis.ticks.y      = element_line(size = 1.5),
      axis.ticks.x      = element_blank(),
      axis.ticks.length = unit(0.1, "cm"),
      axis.title.y      = element_text(size = 17.5, vjust = 0.5, margin = margin(r = 20)),
      panel.grid.major  = element_line(size = 0.5),
      panel.grid.minor  = element_blank(),
      axis.line.x       = element_line(size = 0.5),
      axis.line.y       = element_line(size = 0.5),
      line              = element_line(size = 2, colour = "black"),
      legend.position   = "top",
      legend.direction  = "vertical",
      legend.text       = element_text(size = 13)
    ) +
    ylim(ylim_range[1], ylim_range[2])

  # Add n labels just below the lower whisker of each group
  stats    <- ggplot_build(p)$data[[1]]
  count_df <- combined_df %>%
    distinct() %>%
    count(gene_status) %>%
    bind_cols(stats %>% dplyr::select(x, ymin))

  p + geom_text(
    data  = count_df,
    aes(x = x, y = ymin - 0.5, label = paste0("n= ", n)),
    angle = 45, vjust = 0, hjust = 0.5, size = 6
  )
}
```

## Import data

```{r import-data}

removed_genes <- read_csv(here::here("data", "removed_genes_survey", "removed_genes_all_library.csv"))

removed_gene_list_combined <- list(
  Avana    = na.omit(removed_genes[["avana_removed_genes_all"]]),
  Brunello = na.omit(removed_genes[["brunello_removed_genes_all"]]),
  TKOv3    = na.omit(removed_genes[["toronto_removed_genes_all"]]),
  Yusa     = na.omit(removed_genes[["yusa_removed_genes_all"]]),
  Jacquere = na.omit(removed_genes[["jacquere_removed_genes_all"]])
)

# Genes removed (< 3 surviving guides) in every one of the 5 libraries.
# removed_genes_all_library.csv comes from Fig2A-B_Fig3A_library_overview.Rmd,
# which reads GuideRefine's fresh full_report.xlsx (Step 1-2) -- already
# restricted to the HGNC+T2T-CHM13 universe, so no separate gene-existence
# filter is needed here.
removed_genes_all_libraries <- unique(Reduce(intersect, removed_gene_list_combined))

# Control genes = the full restricted universe (HGNC protein-coding AND
# T2T-CHM13 RefSeq Select -- same definition build_restricted_library.Rmd
# uses) minus whatever's removed above. Recomputed here from the same two raw
# inputs (dated HGNC file, T2T-CHM13 annotation) rather than from a separate
# cached file -- both are cheap to re-filter/intersect, no caching needed.
hgnc_raw_path <- list.files(here::here("data"), pattern = "^hgnc_complete_set_.*\\.txt$", full.names = TRUE)[1]
hgnc_protein_coding <- read_tsv(hgnc_raw_path, show_col_types = FALSE) %>%
  filter(locus_group == "protein-coding gene" & !is.na(ensembl_gene_id)) %>%
  pull(symbol)
t2t_genes <- unique(readRDS(here::here("data", "annotation", "T2T-CHM13v2.0_gene_annot_granges.rds"))$gene)

eligible_genes <- intersect(hgnc_protein_coding, t2t_genes)
control_genes  <- setdiff(eligible_genes, removed_genes_all_libraries)
cat("Removed genes:", length(removed_genes_all_libraries), " | Control genes:", length(control_genes), "\n")

# Reslice the existing spacer/CDS/PAM data by the two gene lists above.
# Combine first, then reslice, so a gene's removed-vs-control label can't
# silently lose its data if it changes category between runs.
#
# - old_removed: the removed-genes spacer/CDS/PAM data. Regenerated fresh each
#   run by the `find-protospacers` chunk below (~5 min, cheap), dynamically
#   named to match the current removed-gene count -- so read it back the same
#   way, by pattern, rather than a hardcoded count that will go stale the
#   moment the removed-gene count changes.
# - old_control: the ~18,777-gene file, a static, unchanging ~1-week
#   findSpacers() computation -- always this exact name, never renamed or
#   regenerated.
old_removed_path <- list.files(here::here("data", "removed_genes_survey"),
                                pattern = "^all_spacers_cds_data_\\d+_removed_genes\\.csv$", full.names = TRUE)[1]
old_removed <- read_csv(old_removed_path, show_col_types = FALSE) %>%
  mutate(chromosome = as.character(chromosome))
old_control <- read_csv(here::here("data", "removed_genes_survey", "all_spacers_cds_data_18777_genes.csv"), show_col_types = FALSE) %>%
  mutate(chromosome = as.character(chromosome))
master_cds_pam <- bind_rows(old_removed, old_control) %>% distinct()
master_genes    <- unique(master_cds_pam$gene_id)

all_spacers_cds_removed_genes <- master_cds_pam %>% filter(gene_id %in% removed_genes_all_libraries)
all_spacers_cds_control_genes <- master_cds_pam %>% filter(gene_id %in% control_genes)

# Safety net: the only expected gap is the documented 7 control genes
# (OR2G6, PLN, SPRR2A/B/D/E/F) with zero spacer rows from strand-specific
# PAM counting on very short CDSs. Any other gap here means a new,
# undocumented problem.
missing_removed <- setdiff(removed_genes_all_libraries, master_genes)
missing_control <- setdiff(control_genes, master_genes)
cat("Removed genes missing from master table:", length(missing_removed), "\n")
cat("Control genes missing from master table:", length(missing_control), "-", paste(sort(missing_control), collapse = ", "), "\n")
stopifnot(length(missing_removed) == 0)
stopifnot(setequal(missing_control, c("OR2G6", "PLN", "SPRR2A", "SPRR2B", "SPRR2D", "SPRR2E", "SPRR2F")))

# Pre-compute PAM counts (reused in barplots, boxplots, and density sections)
removed_pam_count <- count_pam_per_gene(all_spacers_cds_removed_genes)
control_pam_count <- count_pam_per_gene(all_spacers_cds_control_genes)
```

## Figure 3B : UpSet plot commonality of removed genes between 5 libraries

```{r upset-plot}

png(here::here("figure", "Fig3B_removed_genes_all_libraries.png"),
    width = 14000, height = 7500, res = 1000)

par(mar = c(5, 30, 4, 4),
    col.axis = "black", col.lab = "black", col = "black",
    fg = "black", mgp = c(1.5, 0.8, 0))

upset(fromList(removed_gene_list_combined),
      order.by        = "degree",
      sets            = c("Jacquere", "Avana", "Yusa", "TKOv3", "Brunello"),
      keep.order      = TRUE,
      main.bar.color  = "#000000",
      sets.bar.color  = "#1E88E5",
      mainbar.y.label = "Number of genes with < 3 sgRNAs",
      sets.x.label    = "\nTotal genes",
      # text.scale: c(intersection size title, intersection size tick labels,
      #               set size title, set size tick labels, set names, numbers above bars)
      text.scale      = c(2.5, 2.2, 2.5, 2.2, 2.5, 2.0),
      point.size      = 4,
      line.size       = 1.5)

dev.off()
```

## Hypothesis 1: removed genes may be hard to target

First, build the T2T-CHM13 TxDb file (one-time setup — skip if already done).

```{r create-txdb, eval=FALSE}

T2T_txdb <- txdbmaker::makeTxDbFromGFF(
  file       = here::here("data", "annotation", "GCF_009914755.1_T2T-CHM13v2.0_genomic.gff.gz"),
  dataSource = "NCBI T2T Assembly GCF_009914755.1_T2T-CHM13v2.0",
  organism   = "Homo sapiens"
)
saveDb(T2T_txdb, here::here("data", "T2T_data", "TxDb_T2T-CHM13v2.sqlite"))

```

## Compute CDS lengths and gene spans for removed and control genes

```{r compute-cds-lengths}

T2T_txdb <- loadDb(here::here("data", "T2T_data", "TxDb_T2T-CHM13v2.sqlite"))

removed_genes_cds <- calculate_feature_length(T2T_txdb, unique(all_spacers_cds_removed_genes$gene_id), feature = "cds")
control_genes_cds <- calculate_feature_length(T2T_txdb, unique(all_spacers_cds_control_genes$gene_id), feature = "cds")

```

## Obtain possible protospacer sites from CDS of removed genes

```{r find-protospacers}

# Subset canonical CDS ranges for removed genes
cds_by_gene <- cdsBy(T2T_txdb, by = "gene")
cds_gr      <- cds_by_gene[names(cds_by_gene) %in% removed_genes_all_libraries]

# Keep only canonical NP transcripts
cds_gr <- endoapply(cds_gr, function(gr) gr[str_detect(gr$cds_name, "NP")])

# Remap RefSeq seqlevels to chromosome numbers
seqlevels(cds_gr) <- refseq_to_chr[seqlevels(cds_gr)]
cds_gr <- keepSeqlevels(cds_gr, seqlevels(BSgenome.Hsapiens.NCBI.T2TCHM13v2.0),
                        pruning.mode = "coarse")

# Flatten for overlap joining
cds_flat         <- unlist(cds_gr, use.names = FALSE)
cds_flat$gene_id <- rep(names(cds_gr), elementNROWS(cds_gr))
df_cds_flat      <- as.data.frame(cds_flat)

# Reduce (de-duplicate isoform-shared exons) before spacer search
cds_reduced <- unlist(reduce(split(cds_flat, cds_flat$gene_id)), use.names = FALSE)

# Find all SpCas9 protospacers within CDS
data(SpCas9, package = "crisprBase")
cds_find_spacer <- findSpacers(cds_reduced,
                               bsgenome      = BSgenome.Hsapiens.NCBI.T2TCHM13v2.0,
                               crisprNuclease = SpCas9)

hits <- findOverlaps(cds_find_spacer, cds_flat, type = "within")

all_spacers_cds_data <- data.frame(
  gene_id      = cds_flat$gene_id[subjectHits(hits)],
  cds_id       = cds_flat$cds_id[subjectHits(hits)],
  spacer_chr   = as.character(seqnames(cds_find_spacer)[queryHits(hits)]),
  spacer_strand = as.character(strand(cds_find_spacer)[queryHits(hits)]),
  pam          = cds_find_spacer$pam[queryHits(hits)],
  pam_site     = cds_find_spacer$pam_site[queryHits(hits)],
  cut_site     = cds_find_spacer$cut_site[queryHits(hits)],
  spacer_seq   = cds_find_spacer$protospacer[queryHits(hits)],
  stringsAsFactors = FALSE
) %>%
  left_join(df_cds_flat, join_by(gene_id, cds_id)) %>%
  dplyr::select(-c(seqnames, strand)) %>%
  dplyr::rename(start_cds  = start,
                end_cds    = end,
                chromosome = spacer_chr,
                strand     = spacer_strand) %>%
  relocate(cds_name, .before = start_cds)

# Dynamically named -- this recompute only takes ~5 min (unlike the 18777-gene
# file, ~1 week), so it's regenerated fresh under its current, correct gene
# count each run rather than kept under a stale "499" name.
write_csv(all_spacers_cds_data, here::here("data", "removed_genes_survey",
          paste0("all_spacers_cds_data_", length(unique(all_spacers_cds_data$gene_id)), "_removed_genes.csv")))

# Control-genes reslice (from the Import data step, cheap -- no findSpacers()
# recompute needed), also written to its own dynamically-named file.
write_csv(all_spacers_cds_control_genes, here::here("data", "removed_genes_survey",
          paste0("all_spacers_cds_data_", length(unique(all_spacers_cds_control_genes$gene_id)), "_control_genes.csv")))
```

## CDS length analysis: removed vs control genes

Every removed and control gene here already passed the HGNC + T2T-CHM13 restriction upstream (`build_restricted_library.Rmd`), so — unlike the pre-Plan-B pipeline — there is no separate category of genes "removed" for lacking a T2T-CHM13 mapping; that's excluded from the universe before this script ever runs.

### Supp Figure 3A and 3B: Barplot distribution of total CDS length

```{r cds-length-barplot}

cds_bins   <- c(0, 500, 1000, 1500, 2000, 2500, Inf)
cds_labels <- c("0–500", "500–1000", "1000–1500", "1500–2000", "2000–2500", ">2500")

removed_genes_cds$cds_length_bin <- cut(removed_genes_cds$total_cds_length_bp,
                                        breaks = cds_bins, labels = cds_labels,
                                        right = TRUE, include.lowest = TRUE)

control_genes_cds$cds_length_bin <- cut(control_genes_cds$total_cds_length_bp,
                                        breaks = cds_bins, labels = cds_labels,
                                        right = TRUE, include.lowest = TRUE)

barplot_cds_removed <- make_bin_barplot(
  removed_genes_cds, "cds_length_bin", "Total CDS length (bp)"
)
barplot_cds_control <- make_bin_barplot(
  control_genes_cds, "cds_length_bin", "Total CDS length (bp)"
)

ggsave(here::here("figure", paste0("SuppFig3A_cds_length_", length(unique(all_spacers_cds_removed_genes$gene_id)), "_removed_genes.png")),
       plot = barplot_cds_removed, width = 11, height = 5, dpi = 1000)

ggsave(here::here("figure", paste0("SuppFig3B_cds_length_", length(unique(all_spacers_cds_control_genes$gene_id)), "_control_genes.png")),
       plot = barplot_cds_control, width = 11, height = 5, dpi = 1000)
```

### Figure 3C: Boxplot log10 CDS length — removed vs control

```{r cds-length-boxplot}

combined_cds <- bind_rows(
  removed_genes_cds %>% mutate(gene_status = "removed genes",
                                log_transform_cds_length = log10(total_cds_length_bp)),
  control_genes_cds %>% mutate(gene_status = "control genes",
                                log_transform_cds_length = log10(total_cds_length_bp))
) %>%
  mutate(gene_status = factor(gene_status, levels = c("control genes", "removed genes")))

cds_length_boxplot <- make_log10_boxplot(
  combined_cds,
  y_col  = "log_transform_cds_length",
  ylabel = "log10-transformed total CDS length (bp)"
)
cds_length_boxplot

ggsave(here::here("figure", "Fig3C_log10_cds_length_removed_genes.png"),
       plot = cds_length_boxplot, width = 4, height = 6, dpi = 1000)
```

```{r wilcoxon-fig3c}

# Fig 3C: Wilcoxon rank-sum test (unpaired, two-sided)
# Comparison: log10(CDS length) of removed genes vs control genes
# ------------------------------------------------------------------
# Figures below are printed fresh each run (see output) -- comment numbers
# are only the last formally verified pass's reference point, not current.

wt_fig3c <- wilcox.test(log_transform_cds_length ~ gene_status,
                         data        = combined_cds,
                         alternative = "two.sided")
cat("Fig 3C  W =", wt_fig3c$statistic, " p =", wt_fig3c$p.value, "\n")
combined_cds %>%
  group_by(gene_status) %>%
  summarise(median_log10_cds = median(log_transform_cds_length, na.rm = TRUE),
            n = n())

```

## Protospacer / NGG PAM-site analysis: removed vs control genes

### Supp Figure 3C and 3D : Barplot distribution of protospacer count per gene

```{r pam-count-barplot}

pam_bins   <- c(0, 50, 100, 150, 200, 250, Inf)
pam_labels <- c("0–50", "50–100", "100–150", "150–200", "200–250", ">250")

removed_pam_count$count_bin <- cut(removed_pam_count$num_pam_site,
                                   breaks = pam_bins, labels = pam_labels,
                                   right = TRUE, include.lowest = TRUE)

control_pam_count$count_bin <- cut(control_pam_count$num_pam_site,
                                   breaks = pam_bins, labels = pam_labels,
                                   right = TRUE, include.lowest = TRUE)

barplot_pam_removed <- make_bin_barplot(
  removed_pam_count, "count_bin",
  "Number of spacer sequences with NGG PAM-site at CDS"
)
barplot_pam_control <- make_bin_barplot(
  control_pam_count, "count_bin",
  "Number of spacer sequences with NGG PAM-site at CDS"
)

ggsave(here::here("figure", paste0("SuppFig3C_spacer_pam_", length(unique(all_spacers_cds_removed_genes$gene_id)), "_removed_genes.png")),
       plot = barplot_pam_removed, width = 9, height = 5, dpi = 1000)

ggsave(here::here("figure", paste0("SuppFig3D_spacer_pam_", length(unique(all_spacers_cds_control_genes$gene_id)), "_control_genes.png")),
       plot = barplot_pam_control, width = 9, height = 5, dpi = 1000)
```

### Figure 3D: Boxplot log10 PAM count — removed vs control

```{r pam-count-boxplot}

combined_pam <- bind_rows(
  removed_pam_count %>% mutate(gene_status = "removed genes",
                                log_transform_num_pam_site = log10(num_pam_site)),
  control_pam_count %>% mutate(gene_status = "control genes",
                                log_transform_num_pam_site = log10(num_pam_site))
) %>%
  mutate(gene_status = factor(gene_status, levels = c("control genes", "removed genes")))

pam_boxplot <- make_log10_boxplot(
  combined_pam,
  y_col  = "log_transform_num_pam_site",
  ylabel = "log10-transformed protospacer count in CDS",
  sig_label_y = 3
)
pam_boxplot

ggsave(here::here("figure", "Fig3D_log10_pam_removed_genes.png"),
       plot = pam_boxplot, width = 4, height = 6, dpi = 1000)
```

```{r wilcoxon-fig3d}

# Fig 3D: Wilcoxon rank-sum test (unpaired, two-sided)
# Comparison: log10(PAM count) of removed genes vs control genes
# ------------------------------------------------------------------
# Figures below are printed fresh each run (see output) -- comment numbers
# are only the last formally verified pass's reference point, not current.

wt_fig3d <- wilcox.test(log_transform_num_pam_site ~ gene_status,
                         data        = combined_pam,
                         alternative = "two.sided")
cat("Fig 3D  W =", wt_fig3d$statistic, " p =", wt_fig3d$p.value, "\n")
combined_pam %>%
  group_by(gene_status) %>%
  summarise(median_log10_pam = median(log_transform_num_pam_site, na.rm = TRUE),
            n = n())

```

## PAM-site density relative to CDS length

Does a larger CDS simply provide more PAM sites?

```{r pam-density}

removed_genes_density <- compute_pam_density(all_spacers_cds_removed_genes, removed_genes_cds)
control_genes_density <- compute_pam_density(all_spacers_cds_control_genes, control_genes_cds)

# Pearson correlation: PAM count vs CDS length (expected ~0.9 — longer CDS → more PAM sites)
cat("Pearson r (removed genes):",
    cor(removed_genes_density$num_pam_site, removed_genes_density$total_cds_length_bp), "\n")
cat("Pearson r (control genes):",
    cor(control_genes_density$num_pam_site, control_genes_density$total_cds_length_bp), "\n")

# Both groups show r ≈ 0.90–0.92: longer CDS → more PAM sites, so raw count is not
# informative on its own. We therefore compare density (PAM sites per kb of CDS).

pam_density_df <- bind_rows(
  removed_genes_density %>% dplyr::select(gene_id, pam_density) %>% mutate(gene_status = "removed genes"),
  control_genes_density %>% dplyr::select(gene_id, pam_density) %>% mutate(gene_status = "control genes")
) %>%
  mutate(gene_status = factor(gene_status, levels = c("control genes", "removed genes")))
```

### Figure 3E: Density distribution of PAM sites per kb of CDS

```{r pam-density-plot}

# Group medians — used for vertical reference lines
pam_density_medians <- pam_density_df %>%
  group_by(gene_status) %>%
  summarise(median_density = median(pam_density, na.rm = TRUE), .groups = "drop")
# Median: control genes = 70.5,  removed genes = 64.6  (sites / kb CDS)

pam_density_df %>%
  group_by(gene_status) %>%
  summarise(median_pam_density = median(pam_density, na.rm = TRUE), n = n())

wt_fig3e <- wilcox.test(pam_density ~ gene_status,
                         data        = pam_density_df,
                         alternative = "two.sided")
cat("Fig 3E  W =", wt_fig3e$statistic, " p =", wt_fig3e$p.value, "\n")

pam_density_distribution <- ggplot(pam_density_df,
                                   aes(x = pam_density, colour = gene_status)) +
  geom_density(alpha = 0.8, size = 1) +
  # Dashed vertical lines mark each group's median
  geom_vline(data = pam_density_medians,
             aes(xintercept = median_density, colour = gene_status),
             linetype = "dashed", size = 0.7) +
  # Wilcoxon rank-sum p-value annotation
  annotate("text", x = 130, y = 0.028,
           label = paste0("Wilcoxon rank-sum\np = ", formatC(wt_fig3e$p.value, format = "e", digits = 2)),
           size = 4, hjust = 0, vjust = 1, colour = "black") +
  theme_minimal() +
  scale_color_manual(
    name   = "",
    values = c("control genes" = "#1E88E5", "removed genes" = "#E69F00"),
    labels = c("control genes" = "Other protein-coding genes",
               "removed genes" = "Common genes with < 3 sgRNAs")
  ) +
  labs(x = "Number of protospacers with PAM-site per 1 kb", y = "Density") +
  guides(linetype = "none") +
  theme(
    panel.grid.major  = element_blank(),
    panel.grid.minor  = element_blank(),
    axis.text.x       = element_text(size = 15, vjust = 0.7, colour = "black"),
    axis.text.y       = element_text(size = 15, colour = "black"),
    axis.title.x      = element_text(margin = margin(t = 10), size = 15),
    axis.title.y      = element_text(margin = margin(r = 10), size = 15),
    axis.line.x       = element_line(size = 0.5),
    axis.line.y       = element_line(size = 0.5),
    axis.ticks        = element_line(size = 1.5),
    axis.ticks.length = unit(0.1, "cm"),
    legend.position   = "top",
    legend.text       = element_text(size = 12)
  ) +
  xlim(0, 210)

ggsave(here::here("figure", "Fig3E_pam_density_distribution.png"),
       plot = pam_density_distribution, width = 6.5, height = 4, dpi = 1000)
```

Supplementary Table 1C (comprehensive removed-gene T2T-CHM13 CDS/PAM data) is
generated by `scripts/SuppTable1B-C_removed_genes_T2T_annotation.Rmd`, not
here. The old Supplementary Table 1B (GRCh38-vs-T2T status of unmapped genes)
is retired under Plan B; that letter has since been reused for the renamed
rescue-library table (formerly 1C).
