---
title: "Comparison of sgRNA off-target classification between T2T-CHM13 and GRCh38"
author: "Author: Stefanus Bernard"
date: "Date: `r format(Sys.time(), '%d %B %Y')`"
output:
  html_document :
    theme : cerulean
    toc : true
    toc_float : true
    number_section : true
---

```{r setup, include=FALSE}
library(here)
library(tidyverse)

source(here::here("scripts", "CRISPR-KO-library-survey-functions.R"))
```

# Load classification files for both genome builds

```{r}
libraries <- list(
  avana    = here::here("data", "sgrna_lfc_data", "avana_data"),
  brunello = here::here("data", "sgrna_lfc_data", "brunello_data"),
  tkov3    = here::here("data", "sgrna_lfc_data", "tkov3_data"),
  yusa     = here::here("data", "sgrna_lfc_data", "yusa_data"),
  jacquere = here::here("data", "sgrna_lfc_data", "jacquere_data")
)

load_classification <- function(dir_path, library_name, genome) {
  file_path <- file.path(dir_path, paste0(library_name, "_sgrna_alignment_classification_", genome, ".csv"))
  read_csv(file_path, show_col_types = FALSE) %>%
    dplyr::select(sgRNA, spacer, gene, n_mismatches, num_alignments, alignment)
}
```

# Join T2T and GRCh38 classification per sgRNA, per library

For each library, join on `sgRNA` and compare the `alignment` category assigned under each genome build. Any sgRNA whose category differs between builds is written out to its own diff CSV.

```{r}
output_dir <- here::here("data", "sgrna_lfc_data", "T2T_vs_hg38_comparison")
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = TRUE)

compare_library <- function(dir_path, library_name) {
  t2t  <- load_classification(dir_path, library_name, "T2T")
  hg38 <- load_classification(dir_path, library_name, "hg38")

  joined <- full_join(
    t2t, hg38,
    by = c("sgRNA", "spacer", "gene"),
    suffix = c("_T2T", "_hg38")
  )

  diff <- joined %>%
    filter(alignment_T2T != alignment_hg38 | is.na(alignment_T2T) != is.na(alignment_hg38)) %>%
    mutate(library = library_name) %>%
    relocate(library) %>%
    arrange(gene, sgRNA)

  write_csv(diff, here::here(output_dir, paste0(library_name, "_T2T_vs_hg38_diff.csv")))

  transition <- joined %>%
    count(alignment_T2T, alignment_hg38, name = "n_sgRNA") %>%
    mutate(library = library_name) %>%
    relocate(library)

  list(diff = diff, transition = transition, n_total = nrow(joined), n_diff = nrow(diff))
}

results <- imap(libraries, compare_library)
```

# Per-library summary

```{r}
summary_table <- map_dfr(names(results), function(lib) {
  tibble(
    library = lib,
    n_sgRNA_total = results[[lib]]$n_total,
    n_sgRNA_discordant = results[[lib]]$n_diff,
    pct_discordant = round(100 * results[[lib]]$n_diff / results[[lib]]$n_total, 3)
  )
})

write_csv(summary_table, here::here(output_dir, "summary_discordance_per_library.csv"))
summary_table
```

# Transition matrix: which categories shift between builds

For each library, this shows how many sgRNAs moved from a given T2T category to a given GRCh38 category (only rows where the two differ).

```{r}
all_transitions <- map_dfr(results, ~ .x$transition) %>%
  filter(alignment_T2T != alignment_hg38 | is.na(alignment_T2T) != is.na(alignment_hg38)) %>%
  arrange(library, desc(n_sgRNA))

write_csv(all_transitions, here::here(output_dir, "transition_matrix_all_libraries.csv"))
all_transitions
```

# Net change in multi-target and off-target counts (T2T minus GRCh38)

```{r}
category_totals <- function(df, genome_col) {
  df %>%
    count(alignment = .data[[genome_col]], name = "n") %>%
    filter(!is.na(alignment))
}

net_change <- map_dfr(names(results), function(lib) {
  t2t  <- load_classification(libraries[[lib]], lib, "T2T")
  hg38 <- load_classification(libraries[[lib]], lib, "hg38")

  full_join(
    count(t2t, alignment, name = "n_T2T"),
    count(hg38, alignment, name = "n_hg38"),
    by = "alignment"
  ) %>%
    mutate(across(c(n_T2T, n_hg38), ~ replace_na(.x, 0)),
           net_T2T_minus_hg38 = n_T2T - n_hg38,
           library = lib) %>%
    relocate(library)
})

write_csv(net_change, here::here(output_dir, "category_count_net_change_per_library.csv"))
net_change
```

# Figure: GRCh38 vs T2T-CHM13 percentage per category, faceted by library

```{r}
library_totals <- net_change %>%
  group_by(library) %>%
  summarise(total = sum(n_T2T), .groups = "drop")

plot_data <- net_change %>%
  filter(alignment %in% c("multi-target guides", "single mismatch", "pam-distal double mismatch")) %>%
  left_join(library_totals, by = "library") %>%
  pivot_longer(cols = c(n_T2T, n_hg38), names_to = "genome", values_to = "n") %>%
  mutate(genome = recode(genome, n_T2T = "T2T-CHM13", n_hg38 = "GRCh38"),
         percentage = round(100 * n / total, 2),
         alignment = factor(alignment, levels = c("multi-target guides", "single mismatch", "pam-distal double mismatch")),
         library = factor(library, levels = c("brunello", "tkov3", "yusa", "avana", "jacquere")),
         genome = factor(genome, levels = c("GRCh38", "T2T-CHM13")))

genome_palette <- c("GRCh38" = "#56B4E9", "T2T-CHM13" = "#E69F00")

comparison_bar_chart <- ggplot(plot_data, aes(x = alignment, y = percentage, fill = genome)) +
  geom_bar(stat = "identity", position = position_dodge2(padding = 0.2), width = 0.7) +
  facet_wrap(~ library, nrow = 1,
             labeller = labeller(library = c(brunello = "Brunello", tkov3 = "TKOv3",
                                              yusa = "Yusa (Project Score)", avana = "Avana",
                                              jacquere = "Jacquere"))) +
  theme_minimal() +
  labs(title = "", x = "", y = "sgRNA (%)") +
  scale_x_discrete(labels = c('multi-target guides' = 'Multi-target',
                               'single mismatch' = 'Single\nmismatch',
                               'pam-distal double mismatch' = 'PAM-distal\ndouble mismatch')) +
  geom_text(aes(label = percentage), position = position_dodge2(width = 0.7), vjust = -0.5, size = 3) +
  theme(axis.text.x = element_text(size = 10, colour = "black"),
        axis.text.y = element_text(size = 10, colour = "black"),
        strip.text = element_text(size = 11, face = "bold"),
        panel.grid.minor = element_blank(),
        legend.title = element_text(size = 11),
        legend.text = element_text(size = 10)) +
  scale_fill_manual(values = genome_palette, name = "Genome build")

comparison_bar_chart

ggsave(
  width = 16,
  height = 5,
  dpi = 300,
  plot = comparison_bar_chart,
  filename = here::here("figure", "SuppFig2A_T2T_vs_hg38_category_comparison.png"))
```

# Table version of SuppFig2A: GRCh38 vs T2T-CHM13 counts and percentages per category, per library

```{r}
comparison_table <- net_change %>%
  filter(alignment %in% c("multi-target guides", "single mismatch", "pam-distal double mismatch")) %>%
  left_join(library_totals, by = "library") %>%
  mutate(
    pct_hg38 = round(100 * n_hg38 / total, 2),
    pct_T2T = round(100 * n_T2T / total, 2),
    pct_point_change_T2T_minus_hg38 = round(pct_T2T - pct_hg38, 2)
  ) %>%
  dplyr::select(library, alignment, n_hg38, pct_hg38, n_T2T, pct_T2T,
                net_T2T_minus_hg38, pct_point_change_T2T_minus_hg38) %>%
  mutate(library = factor(library, levels = c("brunello", "tkov3", "yusa", "avana", "jacquere")),
         alignment = factor(alignment, levels = c("multi-target guides", "single mismatch", "pam-distal double mismatch"))) %>%
  arrange(library, alignment)

write_csv(comparison_table, here::here(output_dir, "SuppTable2A_T2T_vs_hg38_category_comparison.csv"))
comparison_table
```
