---
title: "Comparison of sgRNA off-target classification between T2T-CHM13 and GRCh38"
author: "Author: Stefanus Bernard"
date: "Date: `r format(Sys.time(), '%d %B %Y')`"
output:
  html_document :
    theme : cerulean
    toc : true
    toc_float : true
    number_section : true
---

```{r setup, include=FALSE}
library(here)
library(tidyverse)

source(here::here("scripts", "CRISPR-KO-library-survey-functions.R"))
```

# Load classification files for both genome builds

```{r}
libraries <- list(
  avana    = here::here("data", "sgrna_lfc_data", "avana_data"),
  brunello = here::here("data", "sgrna_lfc_data", "brunello_data"),
  tkov3    = here::here("data", "sgrna_lfc_data", "tkov3_data"),
  yusa     = here::here("data", "sgrna_lfc_data", "yusa_data"),
  jacquere = here::here("data", "sgrna_lfc_data", "jacquere_data")
)

library_levels <- c("brunello", "tkov3", "yusa", "avana", "jacquere")
alignment_levels <- c("multi-target guides", "single mismatch", "pam-distal double mismatch")

restricted_library_file <- c(
  avana = "avana_library", brunello = "broadgpp-brunello-library-contents",
  tkov3 = "tkov3_guide_sequence", yusa = "yusa_hcrispr_ko_grnas",
  jacquere = "Jacquere_PerGuideAnnotations_Quota4"
)
restricted_sgrna_ids <- function(library_name) {
  path <- here::here("public_crispr_library", "restricted_library", paste0(restricted_library_file[[library_name]], ".tsv"))
  read_tsv(path, col_names = c("sgRNA", "spacer", "gene"), show_col_types = FALSE)$sgRNA
}

load_classification <- function(dir_path, library_name, genome) {
  file_path <- file.path(dir_path, paste0(library_name, "_sgrna_alignment_classification_", genome, ".csv"))
  read_csv(file_path, show_col_types = FALSE) %>%
    dplyr::select(sgRNA, spacer, gene, n_mismatches, num_alignments, alignment) %>%
    filter(sgRNA %in% restricted_sgrna_ids(library_name))
}
```

# Join T2T and GRCh38 classification per sgRNA, per library

For each library, join on `sgRNA` and compare the `alignment` category assigned under each genome build. Any sgRNA whose category differs between builds is written out to its own diff CSV.

```{r}
output_dir <- here::here("data", "sgrna_lfc_data", "T2T_vs_hg38_comparison")
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = TRUE)

compare_library <- function(dir_path, library_name) {
  t2t  <- load_classification(dir_path, library_name, "T2T")
  hg38 <- load_classification(dir_path, library_name, "hg38")

  joined <- full_join(
    t2t, hg38,
    by = c("sgRNA", "spacer", "gene"),
    suffix = c("_T2T", "_hg38")
  )

  diff <- joined %>%
    filter(alignment_T2T != alignment_hg38 | is.na(alignment_T2T) != is.na(alignment_hg38)) %>%
    mutate(library = library_name) %>%
    relocate(library) %>%
    arrange(gene, sgRNA)

  write_csv(diff, here::here(output_dir, paste0(library_name, "_T2T_vs_hg38_diff.csv")))

  transition <- joined %>%
    count(alignment_T2T, alignment_hg38, name = "n_sgRNA") %>%
    mutate(library = library_name) %>%
    relocate(library)

  counts <- full_join(
    count(t2t, alignment, name = "n_T2T"),
    count(hg38, alignment, name = "n_hg38"),
    by = "alignment"
  )

  list(diff = diff, transition = transition, counts = counts,
       n_total = nrow(joined), n_diff = nrow(diff))
}

results <- imap(libraries, compare_library)
```

# Per-library summary

```{r}
summary_table <- map_dfr(names(results), function(lib) {
  tibble(
    library = lib,
    n_sgRNA_total = results[[lib]]$n_total,
    n_sgRNA_discordant = results[[lib]]$n_diff,
    pct_discordant = round(100 * results[[lib]]$n_diff / results[[lib]]$n_total, 3)
  )
})

write_csv(summary_table, here::here(output_dir, "summary_discordance_per_library.csv"))
summary_table
```

# Transition matrix: which categories shift between builds

For each library, this shows how many sgRNAs moved from a given T2T category to a given GRCh38 category (only rows where the two differ).

```{r}
all_transitions <- map_dfr(results, ~ .x$transition) %>%
  filter(alignment_T2T != alignment_hg38 | is.na(alignment_T2T) != is.na(alignment_hg38)) %>%
  arrange(library, desc(n_sgRNA))

write_csv(all_transitions, here::here(output_dir, "transition_matrix_all_libraries.csv"))
all_transitions
```

# Net change in multi-target and off-target counts (T2T minus GRCh38)

```{r}
net_change <- map_dfr(names(results), function(lib) {
  results[[lib]]$counts %>%
    mutate(across(c(n_T2T, n_hg38), ~ replace_na(.x, 0)),
           net_T2T_minus_hg38 = n_T2T - n_hg38,
           library = lib) %>%
    relocate(library)
})

write_csv(net_change, here::here(output_dir, "category_count_net_change_per_library.csv"))
net_change
```

# Figure: GRCh38 vs T2T-CHM13 percentage per category and genome build

```{r}
library_totals <- net_change %>%
  group_by(library) %>%
  summarise(total = sum(n_T2T), .groups = "drop")

library_palette <- c(
  "brunello" = "#E69F00",
  "tkov3"    = "#56B4E9",
  "yusa"     = "#009E73",
  "avana"    = "#F0E442",
  "jacquere" = "#0072B2"
)

x_group_levels <- c(
  "multi-target guides · GRCh38",
  "multi-target guides · T2T-CHM13",
  "single mismatch · GRCh38",
  "single mismatch · T2T-CHM13",
  "pam-distal double mismatch · GRCh38",
  "pam-distal double mismatch · T2T-CHM13"
)

x_group_labels <- c(
  "multi-target guides · GRCh38"            = "Multi-target sgRNA\n(GRCh38)",
  "multi-target guides · T2T-CHM13"         = "Multi-target sgRNA\n(T2T-CHM13)",
  "single mismatch · GRCh38"               = "Single mismatch sgRNA\n(GRCh38)",
  "single mismatch · T2T-CHM13"            = "Single mismatch sgRNA\n(T2T-CHM13)",
  "pam-distal double mismatch · GRCh38"    = "PAM-distal double\nmismatch sgRNA\n(GRCh38)",
  "pam-distal double mismatch · T2T-CHM13" = "PAM-distal double\nmismatch sgRNA\n(T2T-CHM13)"
)

plot_data <- net_change %>%
  filter(alignment %in% alignment_levels) %>%
  left_join(library_totals, by = "library") %>%
  pivot_longer(cols = c(n_T2T, n_hg38), names_to = "genome", values_to = "n") %>%
  mutate(
    genome           = recode(genome, n_T2T = "T2T-CHM13", n_hg38 = "GRCh38"),
    percentage       = round(100 * n / total, 2),
    percentage_label = sprintf("%.2f", percentage),
    library          = factor(library, levels = library_levels),
    x_group          = factor(paste(alignment, genome, sep = " · "), levels = x_group_levels)
  )

comparison_bar_chart <- ggplot(plot_data, aes(x = x_group, y = percentage, fill = library)) +
  geom_bar(stat = "identity", position = position_dodge2(padding = 0.2), width = 0.7) +
  geom_vline(xintercept = c(2.5, 4.5), linetype = "dashed", colour = "#555555", linewidth = 0.7) +
  theme_minimal() +
  labs(title = "", x = "", y = "sgRNA (%)") +
  scale_x_discrete(labels = x_group_labels) +
  scale_y_continuous(breaks = seq(0, 10, 2), limits = c(0, 10)) +
  geom_text(aes(label = percentage_label), position = position_dodge2(width = 0.7), vjust = -0.5, size = 5.25) +
  theme(
    axis.text.x       = element_text(size = 20, vjust = 0.7, colour = "black"),
    axis.text.y       = element_text(size = 20, colour = "black"),
    axis.title.y      = element_text(size = 20, vjust = 0.5, margin = margin(r = 10)),
    panel.grid.major  = element_blank(),
    panel.grid.minor  = element_blank(),
    axis.line.x       = element_line(size = 0.5),
    axis.line.y       = element_line(size = 0.5),
    axis.ticks.y      = element_line(size = 1),
    axis.ticks.length = unit(0.15, "cm"),
    legend.position      = c(0.91, 0.95),
    legend.justification = c(0.5, 1),
    legend.background    = element_rect(fill = "white", colour = NA),
    legend.title         = element_text(size = 20),
    legend.text          = element_text(size = 18),
    plot.margin          = margin(1, 1, 1, 1)
  ) +
  scale_fill_manual(
    values = library_palette,
    name   = "Library",
    breaks = library_levels,
    labels = c(
      "brunello" = "Brunello",
      "tkov3"    = "TKOv3",
      "yusa"     = "Yusa (Project Score)",
      "avana"    = "Avana",
      "jacquere" = "Jacquere"
    )
  )

comparison_bar_chart

ggsave(
  width    = 20,
  height   = 7,
  dpi      = 1000,
  plot     = comparison_bar_chart,
  filename = here::here("figure", "SuppFig1A_T2T_vs_hg38_category_comparison.png"))
```

# Table version of SuppFig1A: GRCh38 vs T2T-CHM13 counts and percentages per category, per library

```{r}
comparison_table <- net_change %>%
  filter(alignment %in% alignment_levels) %>%
  left_join(library_totals, by = "library") %>%
  mutate(
    pct_hg38 = round(100 * n_hg38 / total, 2),
    pct_T2T = round(100 * n_T2T / total, 2),
    pct_point_change_T2T_minus_hg38 = round(pct_T2T - pct_hg38, 2)
  ) %>%
  dplyr::select(library, alignment, n_hg38, pct_hg38, n_T2T, pct_T2T,
                net_T2T_minus_hg38, pct_point_change_T2T_minus_hg38) %>%
  mutate(library = factor(library, levels = library_levels),
         alignment = factor(alignment, levels = alignment_levels)) %>%
  arrange(library, alignment)

write_csv(comparison_table, here::here(output_dir, "SuppTable2A_T2T_vs_hg38_category_comparison.csv"))
comparison_table
```
