---
title: "Supplementary Tables 1D and 1B: 520-gene T2T-CHM13 annotation data and the 21-gene mapping-loss classification"
author: "Author: Stefanus Bernard"
date: "Date: `r format(Sys.time(), '%d %B %Y')`"
output:
  html_document:
    theme: cerulean
    toc: true
    toc_float: true
    number_section: true
---

```{r setup, include=FALSE}
library(here)
library(tidyverse)
library(data.table)
library(GenomicFeatures)
library(openxlsx)
```

## Objective

This is the single, combined source for two related supplementary tables
covering the 520 genes consistently removed across all 5 CRISPR-KO libraries:

- **Supplementary Table 1D** — comprehensive T2T-CHM13 CDS/PAM-site data for
  all 520 genes (`num_pam_site`, `total_cds_length_bp`, `total_gene_length_bp`,
  `cds_length_bin`), with `NA` marking the genes that don't map.
- **Supplementary Table 1B** — for the 21 genes that show up as `NA` in Table
  1D, the *reason* they don't map to T2T-CHM13, classified directly from the
  T2T-CHM13 GFF (no manual lookup): not annotated / reclassified as
  pseudogene / no curated RefSeq transcript / duplicate gene record.

## Step 1: the 520 removed genes (intersection across all 5 libraries)

```{r removed-genes}
removed_genes <- read_csv(here::here("data", "removed_genes_survey", "removed_genes_all_library.csv"),
                           show_col_types = FALSE)

removed_gene_list_combined <- list(
  Avana    = na.omit(removed_genes[["avana_removed_genes_all"]]),
  Brunello = na.omit(removed_genes[["brunello_removed_genes_all"]]),
  TKOv3    = na.omit(removed_genes[["toronto_removed_genes_all"]]),
  Yusa     = na.omit(removed_genes[["yusa_removed_genes_all"]]),
  Jacquere = na.omit(removed_genes[["jacquere_removed_genes_all"]])
)

removed_genes_all_libraries <- unique(Reduce(intersect, removed_gene_list_combined))
length(removed_genes_all_libraries)  # expect 520
```

## Step 2: PAM-site counts (from the pre-computed, NP_-filtered spacer search)

`all_spacers_cds_data_499_removed_genes.csv` was built (in `Fig3B-E_SuppFig2A-D_cds_pam_analysis.Rmd`)
by searching for protospacers only within CDS records whose `cds_name` contains
"NP" (curated RefSeq), while Gnomon-only (`XP_`) genes are excluded there. Its gene
list is therefore the authoritative 499 mapped genes.

```{r pam-counts}
all_spacers_cds_removed_genes <- read_csv(
  here::here("data", "removed_genes_survey", "all_spacers_cds_data_499_removed_genes.csv"),
  show_col_types = FALSE
)

removed_pam_count <- all_spacers_cds_removed_genes %>%
  distinct(gene_id, pam_site) %>%
  count(gene_id, name = "num_pam_site")

mapped_genes_499 <- unique(all_spacers_cds_removed_genes$gene_id)
length(mapped_genes_499)  # expect 499
```

## Step 3: load the T2T-CHM13 TxDb and compute per-gene CDS/gene length

```{r txdb-cds-length}
T2T_txdb <- loadDb(here::here("data", "T2T_data", "TxDb_T2T-CHM13v2.sqlite"))

calculate_cds_length <- function(txdb, gene_list) {
  genes_sub    <- genes(txdb, filter = list(gene_id = gene_list))
  gene_span_df <- data.frame(
    gene                 = mcols(genes_sub)$gene_id,
    total_gene_length_bp = width(genes_sub)
  )

  cds_by_gene <- cdsBy(txdb, by = "gene")
  cds_gr      <- cds_by_gene[names(cds_by_gene) %in% gene_list]
  total_len   <- sum(width(reduce(cds_gr)))

  feat_df <- data.frame(gene = names(total_len), total_cds_length_bp = as.numeric(total_len))
  feat_df %>% left_join(gene_span_df, join_by(gene))
}

# Pass in the pre-filtered 499-gene list (mapped_genes_499), NOT the raw 520 —
# cdsBy() alone would include the 6 Gnomon-only (XP_) genes too.
removed_genes_cds <- calculate_cds_length(T2T_txdb, mapped_genes_499)
nrow(removed_genes_cds)  # expect 499
```

## Step 4: build Supplementary Table 1D (all 520 genes; `NA` marks the 21 unmapped)

```{r build-table-1c}
cds_bins   <- c(0, 500, 1000, 1500, 2000, 2500, Inf)
cds_labels <- c("0–500", "500–1000", "1000–1500", "1500–2000", "2000–2500", ">2500")

hgnc <- read_tsv("https://storage.googleapis.com/public-download-files/hgnc/tsv/tsv/hgnc_complete_set.txt",
                 show_col_types = FALSE)

# HGNC is a live, continuously-updated feed — a gene can be renamed there
# (e.g. ZBTB8OS -> TSLIG1) while the T2T-CHM13 GFF release (and this
# project's gene lists) still use the older symbol. Match on current
# `symbol` first, then fall back to `prev_symbol` so renamed genes aren't
# silently dropped, but keep our own gene_id (not HGNC's current symbol) so
# this table stays internally consistent with the rest of the pipeline.
hgnc_current <- hgnc %>%
  filter(symbol %in% removed_genes_all_libraries) %>%
  transmute(gene_id = symbol, name)

still_missing <- setdiff(removed_genes_all_libraries, hgnc_current$gene_id)

hgnc_via_prev <- if (length(still_missing) > 0) {
  hgnc %>%
    separate_rows(prev_symbol, sep = "\\|") %>%
    filter(prev_symbol %in% still_missing) %>%
    transmute(gene_id = prev_symbol, name)
} else {
  tibble(gene_id = character(), name = character())
}

if (nrow(hgnc_via_prev) > 0) {
  cat("Resolved via HGNC prev_symbol (renamed since T2T-CHM13 annotation):\n")
  print(hgnc_via_prev)
}

hgnc_lookup <- bind_rows(hgnc_current, hgnc_via_prev)

supp_table_1d <- tibble(gene_id = removed_genes_all_libraries) %>%
  left_join(hgnc_lookup, by = "gene_id") %>%
  left_join(removed_pam_count, by = "gene_id") %>%
  left_join(removed_genes_cds, join_by(gene_id == gene)) %>%
  mutate(cds_length_bin = cut(total_cds_length_bp, breaks = cds_bins, labels = cds_labels,
                               right = TRUE, include.lowest = TRUE))

cat("Table 1D rows:", nrow(supp_table_1d), "(expect 520)\n")
cat("Rows with NA num_pam_site (unmapped):", sum(is.na(supp_table_1d$num_pam_site)), "(expect 21)\n")

write.xlsx(supp_table_1d, here::here("results", "supp_table_1d_520_genes.xlsx"))
```

## Step 5: classify the 21 unmapped genes directly from the T2T-CHM13 GFF

Pre-extraction step (run once, from a shell):

```sh
export LC_ALL=C
GFF=data/annotation/GCF_009914755.1_T2T-CHM13v2.0_genomic.gff.gz
OUT=data/T2T_data
genes="KRTAP5-1|KRTAP5-10|PLA2G10|TAS2R43|UPK3BL1|CD200R1L|CYP4B1|EFCAB13|KRT24|KRT81|LEO1|MS4A14|PRSS48|ZNF681|GSTM2|KCTD16|KRTAP9-7|SIGLEC5|SPRR3|TBCD|CLN3"

# Single decompress+scan pass, split by feature type via awk instead of
# re-reading the (multi-GB uncompressed) GFF once per output file.
zgrep -E "gene=(${genes})(;|\$)" "$GFF" | awk -F'\t' \
  -v gene_out="$OUT/_supp1c_gene_records_21genes.tsv" \
  -v cds_out="$OUT/_supp1c_cds_records_21genes.tsv" '
    $3 == "gene" || $3 == "pseudogene" { print > gene_out }
    $3 == "CDS"                        { print > cds_out }
  '
```

```{r target-genes}
unmapped_genes <- setdiff(removed_genes_all_libraries, removed_genes_cds$gene)
length(unmapped_genes)  # expect 21

cat(paste(sort(unmapped_genes), collapse = "|"))
```

```{r refseq-chr-map}
# T2T-CHM13v2.0 RefSeq accessions run sequentially: NC_060925.1-NC_060946.1 = chr1-22,
# then NC_060947.1 = X, NC_060948.1 = Y (same mapping used in Fig3B-E_SuppFig2A-D_cds_pam_analysis.Rmd)
refseq_to_chr <- setNames(c(as.character(1:22), "X", "Y"), sprintf("NC_%06d.1", 60925:60948))
```

```{r gff-read-helper}
read_gff_slice <- function(path) {
  if (!file.exists(path)) {
    stop("Missing ", path, " — run the shell pre-extraction step above first.")
  }
  fread(path, sep = "\t", header = FALSE, quote = "",
        col.names = c("seqid","source","type","start","end","score","strand","phase","attr"))
}
```

```{r gff-gene-records}
gene_lines <- read_gff_slice(here::here("data", "T2T_data", "_supp1c_gene_records_21genes.tsv"))

gene_records <- gene_lines[type %in% c("gene", "pseudogene")]
gene_records[, `:=`(
  gene_id      = sub(".*[;]?gene=([^;]+).*", "\\1", attr),
  gene_biotype = sub(".*gene_biotype=([^;]+).*", "\\1", attr),
  record_id    = sub(".*ID=([^;]+);.*", "\\1", attr),
  chr          = refseq_to_chr[seqid]
)]

gene_records[, .(gene_id, type, gene_biotype, record_id, chr)]
```

```{r gff-cds-check}
cds_lines <- read_gff_slice(here::here("data", "T2T_data", "_supp1c_cds_records_21genes.tsv"))

cds_lines[, `:=`(
  gene_id    = sub(".*gene=([^;]+).*", "\\1", attr),
  protein_id = sub(".*protein_id=([^;]+).*", "\\1", attr)
)]

cds_summary <- cds_lines[, .(
  has_NP_cds = any(startsWith(protein_id, "NP_")),
  has_XP_cds = any(startsWith(protein_id, "XP_"))
), by = gene_id]
```

### Classification rule (applied in order)

1. **Not annotated** — zero gene-level GFF records at all.
2. **Multiple gene-level records** — symbol resolves to >1 GFF record. If
   exactly one of those is `gene_biotype == "protein_coding"`, classification
   proceeds using that record (so a genuinely protein-coding locus is still
   reported by its real status — e.g. "no curated NP_ transcript" — rather
   than being reported only as ambiguous); otherwise (no protein-coding
   record, or more than one) it's reported as **Duplicate gene record**,
   since there's no single unambiguous locus to classify.
3. **Reclassified as pseudogene** — `type == "pseudogene"` /
   `gene_biotype` contains "pseudogene".
4. **No curated RefSeq (NP_) transcript** — `gene_biotype == "protein_coding"`
   but every associated CDS `protein_id` is `XP_` (Gnomon), none `NP_`.

```{r classify}
make_row <- function(g, category, t2t_status, reason)
  data.table(gene_id = g, category = category, t2t_chm13_status = t2t_status, reason_for_loss = reason)

classify_gene <- function(g) {
  recs <- gene_records[gene_id == g]

  if (nrow(recs) == 0) {
    return(make_row(g, "Not annotated in T2T-CHM13v2.0",
               "Absent - no gene model",
               "Locus not annotated in T2T-CHM13v2.0 RefSeq (GCF_009914755.1)"))
  }

  extra_record_note <- ""
  if (uniqueN(recs$record_id) > 1) {
    protein_coding_recs <- recs[gene_biotype == "protein_coding"]
    if (uniqueN(protein_coding_recs$record_id) == 1) {
      # Symbol also resolves to non-coding/partial record(s), but there's
      # exactly one protein-coding locus — classify using that one (below)
      # instead of reporting a merely ambiguous "duplicate", while keeping
      # a note of the other record(s) for transparency.
      other_recs <- recs[record_id != protein_coding_recs$record_id[1]]
      extra_record_note <- paste0(" (symbol also resolves to a separate non-coding/partial GFF record: ",
                                   paste(unique(other_recs$record_id), collapse = ", "),
                                   ", biotype ", paste(unique(other_recs$gene_biotype), collapse = ", "), ")")
      recs <- protein_coding_recs
    } else {
      return(make_row(g, "Duplicate gene record in T2T-CHM13v2.0",
                 paste0("split record: ", paste(recs$record_id, collapse = " + ")),
                 paste0("Symbol resolves to ", uniqueN(recs$record_id), " GFF records (",
                        paste(recs$record_id, collapse = ", "), "), biotypes: ",
                        paste(recs$gene_biotype, collapse = ", "))))
    }
  }
  if (any(recs$gene_biotype %in% c("pseudogene", "transcribed_pseudogene"), recs$type == "pseudogene")) {
    return(make_row(g, "Reclassified as pseudogene in T2T-CHM13v2.0",
               recs$gene_biotype[1],
               paste0("Re-annotated as ", recs$gene_biotype[1], " in T2T-CHM13", extra_record_note)))
  }
  cds <- cds_summary[gene_id == g]
  if (recs$gene_biotype == "protein_coding" && nrow(cds) > 0 && cds$has_XP_cds && !cds$has_NP_cds) {
    return(make_row(g, "No curated RefSeq (NP_) transcript in T2T-CHM13v2.0",
               "protein-coding gene model (Gnomon XP_ only)",
               paste0("Gnomon-predicted (XP_) only, no curated NP_ transcript", extra_record_note)))
  }
  make_row(g, paste0("Other: ", recs$gene_biotype[1]), recs$gene_biotype[1], paste0("unclassified", extra_record_note))
}

t2t_chr_lookup <- gene_records[, .(t2t_chromosome = paste(unique(chr), collapse = ", ")), by = gene_id]

classification <- rbindlist(lapply(unmapped_genes, classify_gene))

classification
```

## Step 6: build Supplementary Table 1B (GRCh38 status from CCDS, T2T-CHM13 status from above)

```{r grch38-lookup}
ccds <- fread(here::here("data", "annotation", "CCDS.20221027.txt"), sep = "\t", quote = "")
setnames(ccds, "#chromosome", "chromosome")

grch38_lookup <- ccds[ccds_status == "Public" & gene %in% unmapped_genes,
                       .(grch38_chromosome = chromosome[1]), by = .(gene_id = gene)]
```

```{r build-table-1d}
supp_table_1b <- classification %>%
  left_join(t2t_chr_lookup, by = "gene_id") %>%
  left_join(grch38_lookup, by = "gene_id") %>%
  mutate(grch38_status = "protein-coding") %>%
  dplyr::select(gene_id, category, grch38_status, grch38_chromosome,
                t2t_chm13_status, t2t_chromosome, reason_for_loss) %>%
  arrange(gene_id)

supp_table_1b

write.xlsx(supp_table_1b, here::here("library_survey_supplementary_materials", "supp_table_1b_GRCh38_vs_T2T_21_unmapped_genes.xlsx"))
```
