---
title: "Supplementary Table 1C: T2T-CHM13 annotation data for removed genes"
author: "Author: Stefanus Bernard"
date: "Date: `r format(Sys.time(), '%d %B %Y')`"
output:
  html_document:
    theme: cerulean
    toc: true
    toc_float: true
    number_section: true
---

```{r setup, include=FALSE}
library(here)
library(tidyverse)
library(data.table)
library(GenomicFeatures)
library(openxlsx)
```

## Objective

**Supplementary Table 1C** — comprehensive T2T-CHM13 CDS/PAM-site data for
every gene removed (< 3 surviving guides in all 5 libraries), under the
restricted HGNC+T2T-CHM13 universe. Supplementary Table 1B (GRCh38-vs-T2T
status for genes that don't map to T2T-CHM13) is retired — every gene here
already passed the T2T-CHM13 restriction upstream, so there's no "unmapped
gene" category left to explain.

## Step 1: removed genes (intersection across all 5 libraries)

```{r removed-genes}
removed_genes <- read_csv(here::here("data", "removed_genes_survey", "removed_genes_all_library.csv"),
                           show_col_types = FALSE)

removed_gene_list_combined <- list(
  Avana    = na.omit(removed_genes[["avana_removed_genes_all"]]),
  Brunello = na.omit(removed_genes[["brunello_removed_genes_all"]]),
  TKOv3    = na.omit(removed_genes[["toronto_removed_genes_all"]]),
  Yusa     = na.omit(removed_genes[["yusa_removed_genes_all"]]),
  Jacquere = na.omit(removed_genes[["jacquere_removed_genes_all"]])
)

removed_genes_all_libraries <- unique(Reduce(intersect, removed_gene_list_combined))
length(removed_genes_all_libraries)
```

## Step 2: PAM-site counts (reslice the existing spacer/CDS data, don't recompute)

Reslice the existing spacer/CDS CSVs by the current removed-gene list, same
pattern as `Fig3B-E_SuppFig2A-D_cds_pam_analysis.Rmd` — concatenate both
existing files first (a gene's removed-vs-control label can flip between runs),
then filter. `old_removed` is read by pattern, not a hardcoded gene count --
that file is regenerated fresh under its current count every run (~5 min,
cheap), so a fixed name would go stale the moment the removed-gene count
changes (this is what the "499" name used to be, before that mismatch was
found and fixed).

```{r pam-counts}
old_removed_path <- list.files(here::here("data", "removed_genes_survey"),
                                pattern = "^all_spacers_cds_data_\\d+_removed_genes\\.csv$", full.names = TRUE)[1]
old_removed <- read_csv(old_removed_path, show_col_types = FALSE) %>%
  mutate(chromosome = as.character(chromosome))
old_control <- read_csv(here::here("data", "removed_genes_survey", "all_spacers_cds_data_18777_genes.csv"), show_col_types = FALSE) %>%
  mutate(chromosome = as.character(chromosome))
master_cds_pam <- bind_rows(old_removed, old_control) %>% distinct()

all_spacers_cds_removed_genes <- master_cds_pam %>% filter(gene_id %in% removed_genes_all_libraries)

removed_pam_count <- all_spacers_cds_removed_genes %>%
  distinct(gene_id, pam_site) %>%
  count(gene_id, name = "num_pam_site")

mapped_removed_genes <- unique(all_spacers_cds_removed_genes$gene_id)
length(mapped_removed_genes)
```

## Step 3: load the T2T-CHM13 TxDb and compute per-gene CDS/gene length

```{r txdb-cds-length}
T2T_txdb <- loadDb(here::here("data", "T2T_data", "TxDb_T2T-CHM13v2.sqlite"))

calculate_cds_length <- function(txdb, gene_list) {
  genes_sub    <- genes(txdb, filter = list(gene_id = gene_list))
  gene_span_df <- data.frame(
    gene                 = mcols(genes_sub)$gene_id,
    total_gene_length_bp = width(genes_sub)
  )

  cds_by_gene <- cdsBy(txdb, by = "gene")
  cds_gr      <- cds_by_gene[names(cds_by_gene) %in% gene_list]
  total_len   <- sum(width(reduce(cds_gr)))

  feat_df <- data.frame(gene = names(total_len), total_cds_length_bp = as.numeric(total_len))
  feat_df %>% left_join(gene_span_df, join_by(gene))
}

removed_genes_cds <- calculate_cds_length(T2T_txdb, mapped_removed_genes)
nrow(removed_genes_cds)
```

## Step 4: build Supplementary Table 1C

```{r build-table-1c}
cds_bins   <- c(0, 500, 1000, 1500, 2000, 2500, Inf)
cds_labels <- c("0–500", "500–1000", "1000–1500", "1500–2000", "2000–2500", ">2500")

hgnc_raw_path <- list.files(here::here("data"), pattern = "^hgnc_complete_set_.*\\.txt$", full.names = TRUE)[1]
hgnc <- read_tsv(hgnc_raw_path, show_col_types = FALSE)

# HGNC is a live, continuously-updated feed — a gene can be renamed there
# (e.g. ZBTB8OS -> TSLIG1) while the T2T-CHM13 GFF release (and this
# project's gene lists) still use the older symbol. Match on current
# `symbol` first, then fall back to `prev_symbol` so renamed genes aren't
# silently dropped, but keep our own gene_id (not HGNC's current symbol) so
# this table stays internally consistent with the rest of the pipeline.
hgnc_current <- hgnc %>%
  filter(symbol %in% removed_genes_all_libraries) %>%
  transmute(gene_id = symbol, name)

still_missing <- setdiff(removed_genes_all_libraries, hgnc_current$gene_id)

hgnc_via_prev <- if (length(still_missing) > 0) {
  hgnc %>%
    separate_rows(prev_symbol, sep = "\\|") %>%
    filter(prev_symbol %in% still_missing) %>%
    transmute(gene_id = prev_symbol, name)
} else {
  tibble(gene_id = character(), name = character())
}

if (nrow(hgnc_via_prev) > 0) {
  cat("Resolved via HGNC prev_symbol (renamed since T2T-CHM13 annotation):\n")
  print(hgnc_via_prev)
}

hgnc_lookup <- bind_rows(hgnc_current, hgnc_via_prev)

supp_table_1c <- tibble(gene_id = removed_genes_all_libraries) %>%
  left_join(hgnc_lookup, by = "gene_id") %>%
  left_join(removed_pam_count, by = "gene_id") %>%
  left_join(removed_genes_cds, join_by(gene_id == gene)) %>%
  mutate(cds_length_bin = cut(total_cds_length_bp, breaks = cds_bins, labels = cds_labels,
                               right = TRUE, include.lowest = TRUE))

cat("Table 1C rows:", nrow(supp_table_1c), "\n")
cat("Rows with NA num_pam_site:", sum(is.na(supp_table_1c$num_pam_site)),
    "(expect only the documented 7-gene strand-specific PAM-counting gap, not a T2T-mapping gap -- see PLAN_B notes)\n")

write.xlsx(supp_table_1c, here::here("library_survey_supplementary_materials", paste0("supp_table_1c_", nrow(supp_table_1c), "_genes.xlsx")))
```
