---
title: "Comparison of Drepanos et al. CFD100 off-target calls vs. our Jacquere alignment classification"
author: "Author: Stefanus Bernard"
date: "Date: `r format(Sys.time(), '%d %B %Y')`"
output:
  html_document :
    theme : cerulean
    toc : true
    toc_float : true
    number_section : true
---

```{r}
library(here)
library(tidyverse)
library(UpSetR)
library(crisprScore)
library(clinfun)
library(openxlsx)

source(here::here("scripts", "CRISPR-KO-library-survey-functions.R"))
output_dir <- here::here("data", "drepanos_bernard_difference")

```


```{r setup, include=FALSE}
# Max CFD score per guide across its off-target site(s) in `sites`, grouped by
# sgRNA plus any extra columns the caller needs downstream (e.g. "upset_group"
# for Section 1, or c("gene", "alignment_bin") for Section 3's per-bin summary).
score_by_site <- function(guides, sites, extra_group_cols = character()) {
  guides %>%
    inner_join(sites, by = c("sgRNA", "spacer")) %>%
    mutate(cfd_score = crisprScore::getCFDScores(spacer, protospacer, pam)$score) %>%
    group_by(across(all_of(c("sgRNA", extra_group_cols)))) %>%
    summarise(cfd_score = max(cfd_score), .groups = "drop")
}

# Jonckheere-Terpstra trend test (same as SuppFig1B-K in
# Fig2C-D_SuppFig1B-K_lfc_stratification.Rmd): tests whether median sgRNA LFC
# decreases monotonically as the number of off-target alignments increases
# across ordered bins. Asymptotic normal approximation; accurate for n > 100
# per group. `df` needs `alignment_bin` (ordered factor) and `mean` (LFC).
run_jt_clinfun <- function(df, fig_label, lib_name, analysis) {
  g   <- as.numeric(df$alignment_bin)
  x   <- df$mean
  N   <- length(x)
  n   <- as.integer(table(g))

  # clinfun for the JT statistic; p.value overflows for large N so compute manually
  res     <- jonckheere.test(x, g, alternative = "decreasing", nperm = 0)
  jt_stat <- as.numeric(res$statistic)

  mu     <- (N^2 - sum(n^2)) / 4
  sigma2 <- (N^2 * (2*N + 3) - sum(n^2 * (2*n + 3))) / 72
  z      <- (jt_stat - mu) / sqrt(sigma2)

  tibble(
    figure       = fig_label,
    library      = lib_name,
    analysis     = analysis,
    n_total      = N,
    JT_statistic = round(jt_stat),
    z_score      = round(z, 1),
    p_value      = pnorm(z)
  )
}
```

> **Scope**: GRCh38/hg38 throughout, to match Drepanos et al.'s RefSeq-GRCh38
> `Jacquere_PerGuideAnnotations_Quota4.csv`. Kept separate from the T2T-CHM13
> figures/scripts; LFC data is keyed by spacer identity, not coordinates, so
> it overlays directly. Runs last (SuppFig1L-N) as an external validation
> case study on top of SuppFig1B-K.
>
> **Restricted to the Fig1B universe** (HGNC + T2T-CHM13), reslicing the
> existing GRCh38 classification down to the restricted+control subset's
> sgRNA IDs rather than re-running GuideRefine for GRCh38. Reframes this from
> "does everything match Drepanos" to "for guides we can compare, is one
> build worse" — the Drepanos-side data itself isn't restricted; anything it
> doesn't cover is simply absent from the inner join, not a gap to explain.
>
> **Note on source file**: uses `Jacquere_PerGuideAnnotations_Quota4.csv`
> (80,124 rows, full coverage of our 4-guides/gene Jacquere library), not
> the earlier `Jacquere_PerGuideAnnotations.csv` (60,550 rows), which turned
> out to be scoped to a different, 3-guides/gene version of the library and
> was missing ~19,000 of our guides entirely. See
> `data/drepanos_bernard_difference/drepanos_comparison_notes.md` for the
> full investigation.

# Section 1 — Drepanos CFD100 vs. Bernard alignment classification (UpSet plot)

## Load and join both datasets

`Off-Target CFD100 Hits` is pipe-delimited per sgRNA for guides mapping to
>1 gene annotation; we take the max hit count and flag CFD = 1.0 off-targets.

```{r}
drepanos_cfd100 <- read_csv(
  here::here("data", "drepanos_bernard_difference", "Jacquere_PerGuideAnnotations_Quota4.csv"),
  show_col_types = FALSE
) %>%
  dplyr::select(spacer = `sgRNA Sequence`, gene = `Target Gene Symbol`,
                off_target_cfd100_hits = `Off-Target CFD100 Hits`) %>%
  mutate(
    max_cfd100_hits = map_dbl(str_split(off_target_cfd100_hits, "\\|"), ~ {
      hits <- as.numeric(.x)
      if (all(is.na(hits))) NA_real_ else max(hits, na.rm = TRUE)
    }),
    has_cfd100_offtarget = max_cfd100_hits > 0
  )

# Restricted to the Fig1B universe (reslicing existing classification, not a
# fresh GRCh38 GuideRefine re-run -- see PLAN_B notes). Any Fig1B-universe gene
# with no corresponding Drepanos/GRCh38 data is simply dropped by the inner
# join below, not treated as a gap needing explanation -- this reframes the
# comparison from "does everything match" to "for guides we can compare,
# which build is worse."
jacquere_restricted_sgrna_ids <- read_tsv(
  here::here("data", "library_data", "restricted_library", "Jacquere_PerGuideAnnotations_Quota4.tsv"),
  col_names = c("sgRNA", "spacer", "gene"), show_col_types = FALSE
)$sgRNA

bernard_classification <- read_csv(
  here::here("data", "sgrna_lfc_data", "jacquere_data", "jacquere_sgrna_alignment_classification_hg38.csv"),
  show_col_types = FALSE
) %>%
  dplyr::select(sgRNA, spacer, gene, n_mismatches, num_alignments, alignment) %>%
  filter(sgRNA %in% jacquere_restricted_sgrna_ids)

joined <- inner_join(drepanos_cfd100, bernard_classification, by = "spacer", suffix = c("_drepanos", "_bernard"))

tibble(
  n_drepanos_sgRNA     = nrow(drepanos_cfd100),
  n_bernard_sgRNA      = nrow(bernard_classification),
  n_matched_by_spacer  = nrow(joined),
  pct_drepanos_matched = round(100 * nrow(joined) / nrow(drepanos_cfd100), 2),
  pct_bernard_matched  = round(100 * nrow(joined) / nrow(bernard_classification), 2)
)
```

## Cross-tabulate Drepanos flags vs. our alignment classification

```{r}
joined %>%
  count(has_cfd100_offtarget, alignment, name = "n_sgRNA") %>%
  arrange(desc(has_cfd100_offtarget), desc(n_sgRNA)) %>%
  group_by(has_cfd100_offtarget) %>%
  mutate(pct_within_group = round(100 * n_sgRNA / sum(n_sgRNA), 2)) %>%
  ungroup()
```

```{r}
joined %>%
  filter(alignment %in% c("multi-target guides", "single mismatch", "pam-distal double mismatch")) %>%
  count(alignment, has_cfd100_offtarget, name = "n_sgRNA") %>%
  group_by(alignment) %>%
  mutate(pct_within_category = round(100 * n_sgRNA / sum(n_sgRNA), 2)) %>%
  ungroup() %>%
  arrange(alignment, desc(has_cfd100_offtarget))
```

Guides we flag as multi-target/off-target but Drepanos does not flag at
CFD = 1.0, and vice versa:

```{r}
joined %>%
  filter(alignment %in% c("multi-target guides", "single mismatch", "pam-distal double mismatch"),
         !has_cfd100_offtarget) %>%
  arrange(alignment, sgRNA, gene_bernard)
```

```{r}
joined %>%
  filter(has_cfd100_offtarget, alignment == "perfect") %>%
  arrange(sgRNA, gene_bernard)
```

## UpSet plot: Drepanos CFD = 1.0 flag vs. our alignment categories

Sets are the sgRNAs matched between the two datasets (`joined`), by sgRNA name.
Guides with `NA` `has_cfd100_offtarget` (Drepanos's CFD100 field couldn't be
evaluated, e.g. intergenic/no-site entries) are excluded here — UpSetR's
binary set membership can't distinguish "confirmed not flagged" from
"undeterminable," so leaving them in would silently fold them into the
"not flagged" bars alongside guides Drepanos actually evaluated.

```{r}
joined_for_upset <- joined %>% filter(!is.na(has_cfd100_offtarget))

upset_sets <- list(
  "Drepanos CFD = 1.0"          = joined_for_upset %>% filter(has_cfd100_offtarget) %>% pull(sgRNA),
  "Multi-target"                = joined_for_upset %>% filter(alignment == "multi-target guides") %>% pull(sgRNA),
  "Single mismatch"              = joined_for_upset %>% filter(alignment == "single mismatch") %>% pull(sgRNA),
  "PAM-distal double mismatch"  = joined_for_upset %>% filter(alignment == "pam-distal double mismatch") %>% pull(sgRNA)
)

upset_params <- list(
  order.by        = "degree",
  sets            = c("Drepanos CFD = 1.0", "Multi-target", "Single mismatch", "PAM-distal double mismatch"),
  keep.order      = TRUE,
  main.bar.color  = "#000000",
  sets.bar.color  = "#1E88E5",
  mainbar.y.label = "Number of sgRNA",
  sets.x.label    = "\nTotal sgRNA",
  # text.scale: c(intersection size title, intersection size tick labels,
  #               set size title, set size tick labels, set names, numbers above bars)
  text.scale      = c(2.5, 3, 2.5, 2.2, 2.5, 3),
  point.size      = 4,
  line.size       = 1.5
)

# Same params reused for the saved PNG and the inline preview below, so the two
# can't drift out of sync with each other.
png(here::here("figure", "SuppFig1L_drepanos_vs_bernard_upset.png"),
    width = 14000, height = 7500, res = 1000)
do.call(upset, c(list(fromList(upset_sets)), upset_params))
dev.off()
```

```{r}
do.call(upset, c(list(fromList(upset_sets)), upset_params))
```

## Median CFD score across all sgRNAs shown in the UpSet plot

Per-site CFD (0-1, `crisprScore::getCFDScores()`), not Drepanos's unbounded
per-guide "Aggregate CFD Score." "Perfect" here just means "not single-mismatch
or pam-distal-double-mismatch" — it can still have real off-target sites in our
alignment data, so those guides are scored from any off-target row they
actually have, rather than assumed CFD = 0.

```{r}
upset_group <- joined_for_upset %>%
  mutate(upset_group = case_when(
    alignment == "pam-distal double mismatch" & has_cfd100_offtarget  ~ "PAM-distal double mismatch + Drepanos CFD=1.0",
    alignment == "single mismatch"             & has_cfd100_offtarget  ~ "Single mismatch + Drepanos CFD=1.0",
    alignment == "multi-target guides"         & has_cfd100_offtarget  ~ "Multi-target + Drepanos CFD=1.0",
    alignment == "pam-distal double mismatch"  & !has_cfd100_offtarget ~ "PAM-distal double mismatch only",
    alignment == "single mismatch"             & !has_cfd100_offtarget ~ "Single mismatch only",
    alignment == "perfect"                     & has_cfd100_offtarget  ~ "Drepanos CFD=1.0 only",
    TRUE ~ NA_character_
  )) %>%
  filter(!is.na(upset_group))

# Loaded once here and reused in Section 3 as well.
jacquere_alignment_hg38_raw <- read_csv(
  here::here("data", "sgrna_lfc_data", "hg38_alignment", "Jacquere_PerGuideAnnotations_Quota4_aln.csv"),
  show_col_types = FALSE
)

single_mismatch_sites <- jacquere_alignment_hg38_raw %>% filter(n_mismatches == 1) %>% dplyr::select(sgRNA, spacer, protospacer, pam)
double_mismatch_sites <- jacquere_alignment_hg38_raw %>% filter(n_mismatches == 2) %>% dplyr::select(sgRNA, spacer, protospacer, pam)
any_mismatch_sites    <- jacquere_alignment_hg38_raw %>% filter(n_mismatches >= 1) %>% dplyr::select(sgRNA, spacer, protospacer, pam)

perfect_sgrna <- upset_group %>% filter(alignment == "perfect") %>% distinct(sgRNA, upset_group)

perfect_scores <- perfect_sgrna %>%
  left_join(upset_group %>% dplyr::select(sgRNA, spacer), by = "sgRNA") %>%
  score_by_site(any_mismatch_sites, "upset_group") %>%
  # guides with genuinely zero off-target rows of any kind keep CFD = 0
  right_join(perfect_sgrna, by = c("sgRNA", "upset_group")) %>%
  mutate(cfd_score = coalesce(cfd_score, 0))

multi_scores     <- upset_group %>% filter(alignment == "multi-target guides") %>%
  distinct(sgRNA, upset_group) %>% mutate(cfd_score = 1)
single_scores    <- score_by_site(upset_group %>% filter(alignment == "single mismatch") %>%
                                     dplyr::select(sgRNA, spacer, upset_group), single_mismatch_sites, "upset_group")
pamdistal_scores <- score_by_site(upset_group %>% filter(alignment == "pam-distal double mismatch") %>%
                                     dplyr::select(sgRNA, spacer, upset_group), double_mismatch_sites, "upset_group")

upset_group_cfd_scores <- bind_rows(perfect_scores, multi_scores, single_scores, pamdistal_scores)

upset_group_cfd_summary <- upset_group_cfd_scores %>%
  group_by(upset_group) %>%
  summarise(n_sgRNA = n(), median_cfd_score = median(cfd_score), .groups = "drop") %>%
  arrange(desc(n_sgRNA))

write_csv(upset_group_cfd_summary, here::here(output_dir, "upset_group_median_cfd_score.csv"))

upset_group_cfd_summary
```

# Section 2 — Jacquere single-mismatch guides: Log2FC stratification (GRCh38)

Same approach as SuppFig1K, staying on the GRCh38 classification loaded in
Section 1. Produces **SuppFig1M**. This section's guide universe is
independent of Section 1 / SuppFig1L — it doesn't use Drepanos's CFD100 data
at all, so guides are restricted only to what this section itself requires:
classified by GuideRefine as "perfect" or "single mismatch," with available
LFC data (a guide with no Log2FC value can't be plotted). Joins by `sgRNA`
**and** `spacer` (not the shared `apply_stratification()` helper, which joins
by `sgRNA` alone) because ~0.32% of Jacquere guides have the same `sgRNA`
label pointing at a different spacer across files.

```{r}
single_guides <- c("perfect", "single mismatch")

# [PLAN B, D5 fix] one combined control-term list (see Fig2C-D_SuppFig1B-K_lfc_stratification.Rmd
# for the full rationale) -- load_library_lfc_raw() now requires it explicitly.
generic_terms <- c("CONTROL", "Control", "control", "INTRON", "Intron", "intron",
                    "LacZ", "luciferase", "EGFP", "NO_SITE", "ONE_SITE_INTERGENIC")
terms_for_filtering <- paste(generic_terms, collapse = "|")

jacquere_lfc_raw <- load_library_lfc_raw(
  here::here("data", "library_data", "restricted_library", "Jacquere_PerGuideAnnotations_Quota4.tsv"),
  here::here("data", "sgrna_lfc_data", "output_normalized", "jacquere_normalized_lfc.csv"),
  terms_for_filtering
)

jacquere_single_hg38 <- count_alignment_bin(filter(bernard_classification, alignment %in% single_guides), single_guides)

jacquere_lfc_single_hg38 <- jacquere_lfc_raw %>%
  inner_join(
    jacquere_single_hg38[[1]] %>% dplyr::select(sgRNA, spacer, alignment, num_alignments, alignment_bin),
    by = c("sgRNA", "spacer")
  ) %>%
  mutate(library = "jacquere") %>%
  drop_na()

jacquere_lfc_single_hg38 %>% count(alignment_bin, name = "n_sgRNA")
```

```{r}
visualize_stratify_alignment(jacquere_lfc_single_hg38,
                              "number of single mismatch \noff-target alignment (GRCh38)",
                              here::here("figure", "SuppFig1M_jacquere_single_mismatch_GRCh38.png"))

write_csv(jacquere_lfc_single_hg38, here::here(output_dir, "jacquere_lfc_single_mismatch_GRCh38.csv"))
```

## Jonckheere-Terpstra trend test

Same test as SuppFig1B-K: tests whether median sgRNA LFC decreases
monotonically as the number of single-mismatch off-target alignments
increases across ordered bins.

```{r}
jt_suppfig1m <- run_jt_clinfun(jacquere_lfc_single_hg38, "SuppFig1M", "Jacquere", "Single mismatch off-target (GRCh38)")
jt_suppfig1m
```

# Section 3 — Jacquere single-mismatch guides: CFD score stratification (GRCh38)

Mirrors Section 2 (same guides, same `alignment_bin`) but scores the actual
off-target site instead of plotting Log2FC. "Perfect" guides (bin "0") have no
single-mismatch off-target site, so CFD = 0. Reuses `single_mismatch_sites`
and `score_by_site()` from Section 1 instead of re-reading the alignment file.

```{r}
perfect_guides_hg38 <- jacquere_lfc_single_hg38 %>%
  filter(alignment == "perfect") %>%
  dplyr::select(sgRNA, gene, alignment_bin) %>%
  mutate(max_cfd_score = 0)

jacquere_cfd_by_bin_hg38 <- jacquere_lfc_single_hg38 %>%
  filter(alignment == "single mismatch") %>%
  dplyr::select(sgRNA, spacer, gene, alignment_bin) %>%
  score_by_site(single_mismatch_sites, c("gene", "alignment_bin")) %>%
  dplyr::rename(max_cfd_score = cfd_score) %>%
  bind_rows(perfect_guides_hg38) %>%
  mutate(alignment_bin = fct_relevel(alignment_bin, "0", "1", "2", "3", "4", "5", "> 5"))

jacquere_cfd_by_bin_hg38 %>%
  group_by(alignment_bin) %>%
  summarise(n_sgRNA = n(), median_cfd_score = median(max_cfd_score), .groups = "drop")
```

## Supplementary Table 1A: CFD score by number of off-target alignments

```{r}
supp_table_1a <- jacquere_cfd_by_bin_hg38 %>%
  group_by(alignment_bin) %>%
  summarise(n_sgRNA = n(), median_cfd_score = median(max_cfd_score), .groups = "drop") %>%
  dplyr::rename(num_offtarget_alignments = alignment_bin)

supp_table_1a

write.xlsx(supp_table_1a, here::here("library_survey_supplementary_materials", "supp_table_1a_jacquere_cfd_score_by_alignment_bin.xlsx"))
```

### Median on-target vs. single mismatch off-target => 1 max CFD (manuscript reporting)

```{r}
jacquere_cfd_by_bin_hg38 %>%
  mutate(offtarget_group = if_else(alignment_bin == "0", "0 off-target alignments", ">=1 off-target alignments")) %>%
  group_by(offtarget_group) %>%
  summarise(n_sgRNA = n(), median_max_cfd_score = median(max_cfd_score), .groups = "drop")
```

```{r}
# Same style/annotation as visualize_stratify_alignment() (SuppFig1M), reimplemented
# here since that helper is hardcoded to LFC's y-axis/range.
cfd_annotation_df <- jacquere_cfd_by_bin_hg38 %>% count(alignment_bin)

cfd_boxplot <- jacquere_cfd_by_bin_hg38 %>%
  ggplot(aes(alignment_bin, max_cfd_score, group = alignment_bin)) +
  geom_boxplot(outlier.shape = NA, size = 0.5) +
  theme_minimal() +
  scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, by = 0.2)) +
  labs(title = "",
       x = "number of single mismatch \noff-target alignment (GRCh38)",
       y = "CFD score (max per sgRNA)") +
  theme(
    axis.text.x = element_text(size = 17, vjust = 0.7, colour = "black"),
    axis.text.y = element_text(size = 17, colour = "black"),
    axis.title.x = element_text(size = 17, margin = margin(t = 10)),
    axis.title.y = element_text(size = 17, margin = margin(r = 10)),
    axis.ticks = element_line(size = 1.5),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.line.x = element_line(size = 0.5),
    axis.line.y = element_line(size = 0.5),
    line = element_line((size = 2), colour = "black"))

# upper whisker per box, for label placement
boxplot_whisker <- ggplot_build(cfd_boxplot)$data[[1]]$ymax

cfd_boxplot <- cfd_boxplot +
  geom_text(
    data = cfd_annotation_df,
    aes(x = alignment_bin, y = boxplot_whisker + 0.02, label = paste0("n= ", n)),
    size = 4,
    angle = 45,
    vjust = -0.2,
    hjust = -0.1
  ) +
  coord_cartesian(clip = "off") +
  theme(plot.margin = margin(10, 20, 10, 10)) +
  scale_y_continuous(limits = c(0, 1.2), breaks = seq(0, 1, by = 0.5))

ggsave(here::here("figure", "SuppFig1N_jacquere_single_mismatch_GRCh38_CFD.png"),
       plot = cfd_boxplot, width = 5, height = 5, dpi = 1000)

write_csv(jacquere_cfd_by_bin_hg38, here::here(output_dir, "jacquere_cfd_by_bin_GRCh38.csv"))

cfd_boxplot
```
